---
title: "Adiposity SDS"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adiposity SDS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(purl = FALSE)
```

## Scope
Non-stratified SDS (z-scores) for anthropometric measures using a single reference (mean/sd), with NA policies, optional raw extreme screening, and SDS capping.

## When to use this
- You want simple z-scores against one reference (no sex stratification).
- You need NA handling and optional capping of implausible raw values or extreme SDS magnitudes.
- You plan to standardize BMI/waist/height (or similar) before modeling or QC.

## What you need (rich guide)
- Required: a reference list (`ref`) with `mean` and `sd > 0` per variable, and matching columns in your data (via `col_map$vars` or identity names).
- Units: data units must match the reference (e.g., BMI kg/m^2, waist cm, height cm).
- Policies: choose `na_action` (keep/omit/error), `check_extreme` + `extreme_action` for raw screening, and `sds_cap` + `extreme_action` for large SDS values.
- Optional: `return_summary = TRUE` to get diagnostics, `diagnostics = TRUE` for warnings, `verbose = TRUE` for progress.

## Load packages and data
Use a small slice of the packaged simulated data. Replace `sim_small` with your data.

```{r}
library(HealthMarkers)
library(dplyr)

sim_path <- system.file("extdata", "simulated_hm_data.rds", package = "HealthMarkers")
sim <- readRDS(sim_path)
sim_small <- sim[1:50, ]
```

## Column map
Map each variable to your column names.

```{r}
col_map <- list(
  vars = list(
    BMI = "BMI",
    waist = "waist",
    height = "height"
  )
)
```

## Walkthrough (build ref, compute SDS)

```{r}
ref <- list(
  BMI = c(mean = mean(sim_small$BMI, na.rm = TRUE), sd = sd(sim_small$BMI, na.rm = TRUE)),
  waist = c(mean = mean(sim_small$waist, na.rm = TRUE), sd = sd(sim_small$waist, na.rm = TRUE)),
  height = c(mean = mean(sim_small$height, na.rm = TRUE), sd = sd(sim_small$height, na.rm = TRUE))
)

asds <- adiposity_sds(
  data = sim_small,
  col_map = col_map,
  ref = ref,
  na_action = "keep",
  check_extreme = FALSE,
  extreme_action = "cap",
  sds_cap = 6,
  verbose = FALSE
)

new_cols <- setdiff(names(asds), names(sim_small))
asds |> slice_head(n = 5) |> select(all_of(new_cols))
```

Interpretation: each `<var>_SDS` is `(x - mean)/sd`; values near 0 are average, positive above, negative below the reference. `extreme_action = "cap"` would cap SDS beyond `sds_cap` if present.

## Missing data and extremes
Compare NA policies and raw screening.

```{r}
demo <- sim_small[1:8, c("BMI", "waist", "height")]
demo$BMI[3] <- NA        # missing
demo$waist[5] <- 300     # extreme high waist
demo$height[2] <- 20     # extreme low height

demo_keep <- adiposity_sds(
  data = demo,
  col_map = col_map,
  ref = ref,
  na_action = "keep",
  check_extreme = TRUE,
  extreme_action = "cap",
  sds_cap = 5,
  verbose = FALSE
)

demo_omit <- adiposity_sds(
  data = demo,
  col_map = col_map,
  ref = ref,
  na_action = "omit",
  check_extreme = TRUE,
  extreme_action = "cap",
  sds_cap = 5,
  verbose = FALSE
)

list(
  keep_rows = nrow(demo_keep),
  omit_rows = nrow(demo_omit),
  capped_preview = demo_keep |> select(ends_with("_SDS")) |> slice_head(n = 3)
)
```

## Expectations
- Each mapped variable must exist; `ref` must provide finite `mean` and `sd > 0` for it.
- `na_action` controls row handling: `keep` propagates NA, `omit` drops, `error` aborts.
- `check_extreme` + `extreme_action` operate on raw values before SDS; `sds_cap` + `extreme_action` govern large SDS magnitudes.
- Numeric coercion warns if NAs are introduced; align data units with reference units.
- `return_summary = TRUE` returns data plus summaries of omissions and extremes.

## Common pitfalls
- Unit mismatches (e.g., inches vs cm) distort SDS; ensure units match the reference.
- Using a biased or tiny sample to build `ref` yields unstable SDS; use an appropriate reference population.
- Forgetting to map all variables in `col_map$vars` causes missing-column errors.
- Leaving `extreme_action = "ignore"` or very high `sds_cap` can hide implausible values; review extremes when QC matters.

## Validation notes
- SDS should roughly center at 0 with SD near 1 if your sample matches the reference.
- Spot-check: `(value - ref_mean)/ref_sd` for one row should equal `<var>_SDS`.
- Enable `check_extreme = TRUE` with tailored `extreme_rules` to flag implausible anthropometrics prior to standardizing.

## See also
- `adiposity_sds_strat()` for sex-stratified SDS.
- `obesity_indices()` and `obesity_metrics()` for related adiposity measures.
