---
title: "Adiposity SDS (Sex-Stratified)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adiposity SDS (Sex-Stratified)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(purl = FALSE)
```

## Scope
Sex-stratified SDS (z-scores) for anthropometric measures using separate male/female references, with NA policies, optional extreme screening, and flexible variable mapping.

## When to use this
- You need sex-specific standardization before modeling or flagging outliers.
- Your study has both men and women and you want SDS relative to sex-appropriate reference means/SDs.
- You want built-in options for missing data handling and plausibility screening.

## What you need (rich guide)
- Required: `sex` (coded M/F or 1/2 after mapping) and at least one anthropometric variable present in both `ref$M` and `ref$F`.
- Reference list: `ref$M` and `ref$F` must cover identical variable names, each with `mean` and `sd > 0`.
- Units: use the same units in your data as in the reference (e.g., BMI kg/m^2, waist cm, height cm).
- Optional: `allow_partial = TRUE` to skip missing variables instead of erroring.

## Load packages and data
Use a small slice of the packaged simulated data. Replace `sim_small` with your data.

```{r}
library(HealthMarkers)
library(dplyr)

sim_path <- system.file("extdata", "simulated_hm_data.rds", package = "HealthMarkers")
sim <- readRDS(sim_path)

# Normalize sex labels to M/F and ensure both sexes appear in the demo slice
sim_small <- sim |>
  mutate(sex = case_when(
    sex %in% c("M", "F") ~ sex,
    sex %in% c("male", "Male", "m") ~ "M",
    sex %in% c("female", "Female", "f") ~ "F",
    sex %in% c(1, "1") ~ "M",
    sex %in% c(2, "2") ~ "F",
    TRUE ~ NA_character_
  )) |>
  filter(!is.na(sex)) |>
  slice_head(n = 200)
```

## Column map
Map sex and each variable to your column names.

```{r}
col_map <- list(
  sex = "sex",
  vars = list(
    BMI = "BMI",
    waist = "waist",
    height = "height"
  )
)
```

## Walkthrough (build refs, compute SDS)

```{r, eval=FALSE}
# Build sex-specific references from the demo slice
ref_sex <- sim_small |>
  group_by(sex) |>
  summarise(
    BMI = list(c(mean = mean(BMI, na.rm = TRUE), sd = sd(BMI, na.rm = TRUE))),
    waist = list(c(mean = mean(waist, na.rm = TRUE), sd = sd(waist, na.rm = TRUE))),
    height = list(c(mean = mean(height, na.rm = TRUE), sd = sd(height, na.rm = TRUE)))
  )

if (!all(c("M", "F") %in% ref_sex$sex)) {
  stop("Demo data slice must contain both M and F; normalize or increase sample.")
}

ref <- list(
  M = list(
    BMI = ref_sex$BMI[[ref_sex$sex == "M"]],
    waist = ref_sex$waist[[ref_sex$sex == "M"]],
    height = ref_sex$height[[ref_sex$sex == "M"]]
  ),
  F = list(
    BMI = ref_sex$BMI[[ref_sex$sex == "F"]],
    waist = ref_sex$waist[[ref_sex$sex == "F"]],
    height = ref_sex$height[[ref_sex$sex == "F"]]
  )
)

asds_strat <- adiposity_sds_strat(
  data = sim_small,
  col_map = col_map,
  ref = ref,
  na_action = "keep",
  check_extreme = FALSE,
  extreme_action = "cap",
  allow_partial = FALSE,
  prefix = "",
  verbose = FALSE
)

new_cols <- setdiff(names(asds_strat), names(sim_small))
asds_strat |>
  slice_head(n = 5) |>
  select(all_of(new_cols))
```

Interpretation: each `<var>_SDS` is standardized against sex-specific mean/SD; values near 0 are average for that sex, positive above, negative below.

## Missing data and extremes
Compare row handling and capping.

```{r, eval=FALSE}
demo <- sim_small[1:8, c("sex", "BMI", "waist", "height")]
demo$BMI[3] <- NA         # missing
demo$waist[5] <- 300      # extreme high waist
demo$height[2] <- 20      # extreme low height

demo_keep <- adiposity_sds_strat(
  data = demo,
  col_map = col_map,
  ref = ref,
  na_action = "keep",
  check_extreme = TRUE,
  extreme_action = "cap",
  allow_partial = FALSE,
  prefix = ""
)

demo_omit <- adiposity_sds_strat(
  data = demo,
  col_map = col_map,
  ref = ref,
  na_action = "omit",
  check_extreme = TRUE,
  extreme_action = "cap",
  allow_partial = FALSE,
  prefix = ""
)

list(
  keep_rows = nrow(demo_keep),
  omit_rows = nrow(demo_omit),
  capped_preview = demo_keep |> select(ends_with("_SDS")) |> slice_head(n = 3)
)
```

## Expectations
- `col_map$sex` must be present; sex must map to M/F (or 1/2) or the call errors.
- `ref$M` and `ref$F` must list identical variables with finite `mean` and `sd > 0`.
- All mapped columns must exist unless `allow_partial = TRUE`, in which case missing variables are skipped with an info message.
- `na_action` controls whether rows with missing inputs are kept, dropped, or error; missing values propagate to SDS when kept.
- `check_extreme` + `extreme_action` screen raw inputs before standardizing; SDS outputs themselves are not capped.
- Numeric coercion happens with warnings if NAs are introduced; align your data units with the reference units.

## Common pitfalls
- Sex not normalized to M/F (or 1/2) leads to an error; recode before calling.
- References built from a small or biased subset will give unstable SDS; use appropriate population means/SDs.
- Unit mismatches (e.g., inches vs cm, lb vs kg) will distort SDS; keep units consistent with the reference.
- Forgetting `allow_partial = TRUE` when some variables are absent will error; enable it to compute what’s available.

## Validation notes
- Within each sex, SDS should center near 0 with SD near 1 if references match the data population.
- Spot-check a variable: `(value - sex_mean)/sex_sd` should match the corresponding `<var>_SDS`.
- Use `check_extreme = TRUE` with tailored `extreme_rules` to catch implausible anthropometrics before standardizing.

## See also
- `adiposity_sds()` for pooled-sex SDS.
- `obesity_indices()` and `obesity_metrics()` for related adiposity markers.
