---
title: "Allostatic Load"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{allostatic_load}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(purl = FALSE)
```

## Scope
Composite Allostatic Load score that flags biomarkers above high-risk thresholds and sums the flags. With multiple biomarkers the rule is `>`; with a single biomarker it is `>=`. Optional extreme screening applies only to columns whose names contain "sds".

## When to use this
- You want a simple count of biomarkers exceeding risk cut-points.
- You have a small set of biomarkers with clear thresholds you control.
- You may include pre-computed SDS columns and optionally cap implausible SDS values.

## What you need
- Thresholds: named list of scalar numeric cutoffs; names must match columns (or be mapped via `col_map`).
- Data: matching biomarker columns; units should align with the thresholds you supply.
- Policies: `na_action` (`keep` treats missing as unflagged; `omit` drops rows; `error` aborts). `check_extreme`/`sds_limit`/`extreme_action` only affect columns whose names contain "sds".
- Optional: `return_summary = TRUE` for counts of flags/omissions; `verbose` for progress.

## Load packages and data
Use a small subset of the simulated data. Replace `sim_small` with your data.

```{r}
library(HealthMarkers)
library(dplyr)

sim_path <- system.file("extdata", "simulated_hm_data.rds", package = "HealthMarkers")
sim <- readRDS(sim_path)
sim_small <- sim[1:50, ]
```

## Column map and thresholds
Provide thresholds and, if needed, map them to your column names.

```{r}
thresholds <- list(sbp = 130, dbp = 85, CRP = 3)
col_map <- list(sbp = "sbp", dbp = "dbp", CRP = "CRP")
```

## Walkthrough (compute Allostatic Load)

```{r}
al <- allostatic_load(
  data = sim_small,
  thresholds = thresholds,
  col_map = col_map,
  na_action = "keep",
  check_extreme = FALSE,
  verbose = FALSE
)

al |> slice_head(n = 5)
```

Interpretation: `AllostaticLoad` is the count of biomarkers above thresholds. With >1 biomarker the rule is `>`; with a single biomarker it is `>=`.

## Missing data and screening
Demonstrate row handling and SDS screening (only affects columns containing "sds").

```{r}
demo <- sim_small[1:8, c("sbp", "dbp", "CRP")]
demo$sbp[3] <- NA

demo$CRP_sds <- c(0, 7, -8, 1, NA, 0, 0, 0)  # example SDS column to show screening

al_keep <- allostatic_load(
  data = demo,
  thresholds = thresholds,
  col_map = c(col_map, CRP_sds = "CRP_sds"),
  na_action = "keep",
  check_extreme = TRUE,
  extreme_action = "cap",
  sds_limit = 6,
  verbose = FALSE
)

al_omit <- allostatic_load(
  data = demo,
  thresholds = thresholds,
  col_map = c(col_map, CRP_sds = "CRP_sds"),
  na_action = "omit",
  check_extreme = TRUE,
  extreme_action = "cap",
  sds_limit = 6,
  verbose = FALSE
)

list(
  keep_rows = nrow(al_keep),
  omit_rows = nrow(al_omit),
  head_keep = al_keep |> slice_head(n = 3)
)
```

## Expectations
- Threshold names must match columns (or be mapped); missing columns abort.
- `na_action`: `keep` treats missing as unflagged (flag = 0), `omit` drops rows, `error` aborts on missing.
- `check_extreme`/`sds_limit`/`extreme_action` affect only columns whose names contain "sds" (case-insensitive).
- Multi-biomarker rule is `>`; single-biomarker rule is `>=`.

## Common pitfalls
- Threshold units mismatched to data units will misflag results; align units first.
- Passing only one biomarker changes the rule to `>=`; set thresholds accordingly.
- SDS screening does nothing unless column names contain "sds"; rename if you expect screening.
- `na_action = keep` leaves missing inputs unflagged (zero); use `omit` for stricter handling.

## Validation notes
- Spot-check one biomarker: flag = 1 if value exceeds the threshold (or equals when only one biomarker), else 0.
- Sum of per-biomarker flags should equal `AllostaticLoad` for each row.
- If using `check_extreme`, verify capped SDS values are truncated to +/- `sds_limit` when `extreme_action = "cap"`.

## See also
- `adiposity_sds()` and `adiposity_sds_strat()` for SDS creation.
- `health_markers()` and `health_summary()` for broader panels.
