---
title: "Atherogenic Indices"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{atherogenic_indices}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Scope
Compute atherogenic risk ratios from a lipid panel:
- AIP = log10(TG / HDL_c)
- Castelli Risk Index I (CRI_I) = TC / HDL_c
- Castelli Risk Index II (CRI_II) = LDL_c / HDL_c

All ratios are unitless; use consistent lipid units (mg/dL or mmol/L).

## When to use
- You have triglycerides and HDL cholesterol and want AIP.
- You also have total and/or LDL cholesterol and want Castelli indices.
- You need optional extreme-value screening or strict missing-data handling.

## Requirements checklist
- Packages: HealthMarkers, dplyr (for display).
- Data frame with lipid columns; required keys: TG and HDL_c. Optional: TC and LDL_c.
- Numeric lipids; non-numeric are coerced with warnings; non-finite become NA.
- Decide row policy: na_action = keep (default), omit, or error.
- Optional extreme screening: enable check_extreme and choose extreme_action.

## Load packages and example data
Replace the example slice with your data frame.

```{r}
library(HealthMarkers)
library(dplyr)

sim_path <- system.file("extdata", "simulated_hm_data.rds", package = "HealthMarkers")
sim <- readRDS(sim_path)
sim_small <- sim[1:50, ]
```

## Map lipid columns
TG and HDL_c are required. TC and LDL_c are optional but needed for Castelli indices.

```{r}
col_map <- list(
  TG = "TG",
  HDL_c = "HDL_c",
  TC = "TC",
  LDL_c = "LDL_c"
)
```

## Quick start: compute indices
Defaults keep rows with missing lipids and return NA for their ratios.

```{r}
aip_out <- atherogenic_indices(
  data = sim_small,
  col_map = col_map,
  na_action = "keep",
  check_extreme = FALSE,
  extreme_action = "warn",
  verbose = FALSE
)

new_cols <- setdiff(names(aip_out), names(sim_small))
head(select(aip_out, all_of(new_cols)))
```

## Arguments that matter
- col_map: named list for TG, HDL_c (required), and optional TC, LDL_c. Missing required keys error.
- na_action: keep (default, rows retained; ratios become NA), omit (drop rows with NA in used lipids), error (abort on NA).
- check_extreme: FALSE by default; if TRUE, screen lipids using extreme_rules.
- extreme_action: warn (default, no change), cap, error, ignore, or NA. Defaults use wide mg/dL ranges: TG 0-10000, HDL_c 0-1000, LDL_c 0-10000, TC 0-10000.
- verbose: emit step messages.

## Handling missing and zero denominators
- Non-numeric lipids are coerced to numeric; NA introduced are warned. Non-finite become NA.
- Zero HDL_c denominators yield NA ratios and a warning; Castelli indices use HDL_c as the divisor.
- Row policy: keep vs omit vs error via na_action.

### Compare row policies
```{r}
demo <- sim_small
demo$HDL_c[c(2, 5)] <- NA

aip_keep <- atherogenic_indices(demo, col_map, na_action = "keep")
aip_omit <- atherogenic_indices(demo, col_map, na_action = "omit")

list(
  keep_rows = nrow(aip_keep),
  omit_rows = nrow(aip_omit),
  sample_ratios = head(select(aip_keep, AIP, CRI_I, CRI_II))
)
```

## Extreme-value screening (optional)
Enable check_extreme to cap, warn, error, ignore, or set NA for out-of-range lipids.

```{r}
demo2 <- sim_small
demo2$TG[5] <- 12000  # extreme TG

aip_screen <- atherogenic_indices(
  data = demo2,
  col_map = col_map,
  na_action = "keep",
  check_extreme = TRUE,
  extreme_action = "cap",
  extreme_rules = list(TG = c(0, 10000), HDL_c = c(0, 1000), LDL_c = c(0, 10000), TC = c(0, 10000)),
  verbose = FALSE
)

head(select(aip_screen, AIP, CRI_I, CRI_II))
```

## Outputs
- AIP: log10(TG / HDL_c)
- CRI_I: TC / HDL_c (NA if TC not mapped/present)
- CRI_II: LDL_c / HDL_c (NA if LDL_c not mapped/present)
Rows only drop when na_action = "omit" or when na_action = "error" encounters missing.

## Pitfalls and tips
- Keep lipid units consistent; ratios are unitless but inputs must share units.
- Mapping omissions for TG or HDL_c throw errors; missing TC/LDL_c just yield NA Castelli indices.
- Zero HDL_c produces NA ratios and a warning; address true zeros before computing.
- Tighten extreme_rules if your lab ranges are narrower; use extreme_action = "error" for strict QA.
- Use na_action = "error" for pipelines that must fail fast on missing lipids.

## Validation ideas
- Spot-check AIP: if TG = 150 and HDL_c = 50, expected AIP = log10(3) ~ 0.477.
- CRI_I should decrease if HDL_c rises with other lipids unchanged; similar for CRI_II.
- Confirm that rows with NA in TG/HDL_c are dropped only when na_action = "omit".

## See also
- health_summary() for combined marker panels.
- glycemic_markers() for glucose and insulin-derived indices.
