---
title: "Bone Turnover Markers"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{bone_markers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}

## Scope
Compute OSTA, ALMI, FMI, and BMD_Tscore from DXA and anthropometry, and optionally pass through bone-turnover markers (TBS, HSA, PINP, CTX, BSAP, Osteocalcin). Inputs are coerced to numeric; non-finite become NA. BMD_Tscore uses supplied reference mean and SD; SD must be positive.

## When to use
- You have DXA outputs (ALM, FM, BMD) plus age, weight, height, and reference BMD stats.
- You want OSTA and mass indexes (ALMI/FMI) alongside BMD T-scores.
- You may have additional bone markers to pass through unchanged.

## Requirements checklist
- Packages: HealthMarkers, dplyr (for display).
- Required columns: age, weight, height (meters), ALM, FM (kg), BMD (g/cm^2), BMD_ref_mean, BMD_ref_sd (positive).
- Optional pass-through: TBS, HSA, PINP, CTX, BSAP, Osteocalcin (coerced to numeric if mapped and present).
- Row policy via na_action: keep (default), omit, or error on missing/non-finite required inputs.
- Optional SDS screening: check_extreme on any mapped key containing "sds"; control with sds_limit and extreme_action.

## Load packages and example data
The simulated data lack DXA values; we create simple placeholders for illustration only; replace with your measurements.

```{r}
library(HealthMarkers)
library(dplyr)

sim_path <- system.file("extdata", "simulated_hm_data.rds", package = "HealthMarkers")
sim <- readRDS(sim_path)
sim_small <- sim[1:50, ]

# Illustrative stand-ins; replace with real DXA outputs
sim_small$ALM <- pmax(12, pmin(30, sim_small$weight * 0.32))
sim_small$FM  <- pmax(8,  pmin(40, sim_small$weight * 0.28))
sim_small$BMD <- 0.9 + (sim_small$age - mean(sim_small$age, na.rm = TRUE)) * 0.001
sim_small$BMD_ref_mean <- 1.0
sim_small$BMD_ref_sd   <- 0.12
```

## Map columns
Required keys plus optional markers (commented out below).

```{r}
col_map <- list(
  age = "age",
  weight = "weight",
  height = "height",
  ALM = "ALM",
  FM = "FM",
  BMD = "BMD",
  BMD_ref_mean = "BMD_ref_mean",
  BMD_ref_sd   = "BMD_ref_sd"
  # Optional: TBS = "TBS", HSA = "HSA", PINP = "PINP", CTX = "CTX", BSAP = "BSAP", Osteocalcin = "Osteocalcin"
)
```

## Quick start: compute markers
Defaults keep rows with missing inputs and return NA for their derived scores.

```{r}
bm_out <- bone_markers(
  data = sim_small,
  col_map = col_map,
  na_action = "keep",
  check_extreme = FALSE,
  sds_limit = 6,
  extreme_action = "cap"
)

new_cols <- setdiff(names(bm_out), names(sim_small))
head(select(bm_out, all_of(new_cols)))
```

## Arguments that matter
- col_map: required mappings for age, weight, height, ALM, FM, BMD, BMD_ref_mean, BMD_ref_sd; optional TBS/HSA/PINP/CTX/BSAP/Osteocalcin pass through when mapped.
- na_action: keep (default, retain rows; derived scores become NA when inputs missing), omit (drop rows with missing/non-finite required inputs), error (abort if any missing/non-finite required inputs).
- check_extreme: FALSE by default; TRUE scans mapped keys containing "sds" (case-insensitive) for |value| > sds_limit.
- extreme_action (when check_extreme = TRUE): cap, NA, or error.
- sds_limit: threshold for SDS-like screening (default 6).

## Handling missing inputs
- Non-numeric inputs are coerced; NA introduced are warned. Non-finite become NA.
- BMD_ref_sd and height must be positive or the function aborts.
- Row policy: keep vs omit vs error is governed by na_action.

### Compare row policies
```{r}
demo <- sim_small[1:8, c("age","weight","height","ALM","FM","BMD","BMD_ref_mean","BMD_ref_sd")]
demo$ALM[c(2, 5)] <- NA

a_keep <- bone_markers(demo, col_map, na_action = "keep")
a_omit <- bone_markers(demo, col_map, na_action = "omit")

list(
  keep_rows = nrow(a_keep),
  omit_rows = nrow(a_omit),
  preview = head(select(a_keep, OSTA, ALMI, FMI, BMD_Tscore))
)
```

## Optional SDS screening
If you supply any mapped key containing "sds", you can cap/NA/error on extremes.

```{r}
demo2 <- demo
# Example SDS-like column for illustration
set.seed(1)
demo2$BMD_z_sds <- rnorm(nrow(demo2), mean = 0, sd = 7)
col_map_sds <- c(col_map, list(BMD_z_sds = "BMD_z_sds"))

bm_screen <- bone_markers(
  data = demo2,
  col_map = col_map_sds,
  na_action = "keep",
  check_extreme = TRUE,
  sds_limit = 6,
  extreme_action = "cap"
)

summary(abs(demo2$BMD_z_sds) > 6)
```

## Outputs
- OSTA, ALMI, FMI, BMD_Tscore
- Optional passthrough (if mapped and present): TBS, HSA, PINP, CTX, BSAP, Osteocalcin
Rows drop only with na_action = "omit" or when na_action = "error" aborts.

## Pitfalls and tips
- Ensure BMD_ref_sd > 0 and height > 0; otherwise the function aborts before row filtering.
- Replace placeholder ALM/FM/BMD values with actual DXA measurements; keep units consistent (height meters; masses kg; BMD g/cm^2).
- Optional markers are numeric-coerced; if missing or not mapped, they return NA columns.
- SDS screening applies only to mapped keys containing "sds"; defaults do nothing unless you provide such keys.

## Validation ideas
- Manual check: OSTA for age 70 and weight 60 is (60 - 70) * 0.2 = -2.
- ALMI should equal ALM / height^2; FMI equals FM / height^2.
- BMD_Tscore should drop by ~1 when BMD is one SD below reference.

## See also
- sarc_f_score() and alm_bmi_index() for sarcopenia-related assessments.
- frax_score() for fracture risk estimation.
