---
title: "Charlson Comorbidity Index"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{charlson_index}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
---

```{r, include=FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(purl = FALSE)
```

## Scope
Compute the 19-condition Charlson Comorbidity Index (CCI) without age points. Handles binary indicators, optional bounds screening, and NA policies.

## When to use
- You have binary comorbidity indicators and need the canonical Charlson score (no age points here).
- You want explicit row-retention rules for missing indicators and optional bounds checks on 0/1 flags.
- You may need to cap or error on out-of-bounds inputs.

## Requirements checklist
- Packages: HealthMarkers, dplyr (for display).
- Data columns: binary 0/1 indicators for the 19 conditions.
- Column map: provide all 19 keys -> columns; missing mappings abort.
- Row policy: na_action = keep (default), omit, error; warn/ignore act like keep but warn.
- Optional bounds screening: check_extreme + extreme_action (warn/cap/error/ignore/NA); defaults use 0-1.

## Load packages and example data
Illustrative-only flags below; replace with your real comorbidity indicators.

```{r}
library(HealthMarkers)
library(dplyr)

sim_path <- system.file("extdata", "simulated_hm_data.rds", package = "HealthMarkers")
sim <- readRDS(sim_path)
sim_small <- sim[1:40, ]

# Illustrative binary flags
set.seed(123)
sim_small <- sim_small %>% mutate(
  mi = rbinom(n(), 1, 0.1),
  chf = rbinom(n(), 1, 0.1),
  pvd = rbinom(n(), 1, 0.1),
  stroke = rbinom(n(), 1, 0.1),
  dementia = rbinom(n(), 1, 0.05),
  copd = rbinom(n(), 1, 0.15),
  rheum = rbinom(n(), 1, 0.05),
  ulcer = rbinom(n(), 1, 0.05),
  mild_liver = rbinom(n(), 1, 0.05),
  diabetes = rbinom(n(), 1, 0.2),
  diab_comp = rbinom(n(), 1, 0.05),
  hemiplegia = rbinom(n(), 1, 0.02),
  renal = rbinom(n(), 1, 0.08),
  cancer = rbinom(n(), 1, 0.08),
  leukemia = rbinom(n(), 1, 0.01),
  lymphoma = rbinom(n(), 1, 0.01),
  sev_liver = rbinom(n(), 1, 0.02),
  metastatic_cancer = rbinom(n(), 1, 0.01),
  hiv = rbinom(n(), 1, 0.01)
)
```

## Map columns
Provide all 19 keys. Missing mappings error.

```{r}
col_map <- list(
  mi="mi", chf="chf", pvd="pvd", stroke="stroke", dementia="dementia", copd="copd",
  rheum="rheum", ulcer="ulcer", mild_liver="mild_liver", diabetes="diabetes",
  diab_comp="diab_comp", hemiplegia="hemiplegia", renal="renal", cancer="cancer",
  leukemia="leukemia", lymphoma="lymphoma", sev_liver="sev_liver",
  metastatic_cancer="metastatic_cancer", hiv="hiv"
)
```

## Quick start: compute CCI
Defaults keep rows with missing indicators and return NA scores for them.

```{r}
cci_out <- charlson_index(
  data = sim_small,
  col_map = col_map,
  na_action = "keep",
  check_extreme = FALSE,
  extreme_action = "warn",
  verbose = FALSE
)

head(cci_out)
```

## Arguments that matter
- col_map: all 19 keys required; missing or empty entries abort.
- na_action: keep (default), omit (drop rows with missing indicators), error (abort on missing); warn/ignore behave like keep but warn.
- check_extreme: FALSE by default; TRUE enforces bounds.
- extreme_action: warn (default), cap, error, ignore, NA; default bounds 0-1 per indicator unless overridden via extreme_rules.
- verbose: emit step messages.

## Handling missing and non-binary inputs
- Non-numeric inputs are coerced; NA introduced are warned. Non-finite become NA.
- Missing indicators yield NA scores unless na_action = "omit" or "error".
- Out-of-range (non 0/1) values warn; enable check_extreme to cap/error/NA.

### Compare row policies
```{r}
demo <- sim_small[1:8, names(col_map)]
demo$renal[3] <- NA

a_keep <- charlson_index(demo, col_map, na_action = "keep", check_extreme = FALSE)
a_omit <- charlson_index(demo, col_map, na_action = "omit", check_extreme = FALSE)

list(
  keep_rows = nrow(a_keep),
  omit_rows = nrow(a_omit),
  preview = head(a_keep$charlson_index)
)
```

## Bounds screening (optional)
Cap or error on inputs outside chosen ranges.

```{r}
demo2 <- demo
demo2$mi[5] <- 2  # out of bounds

a_cap <- charlson_index(
  data = demo2,
  col_map = col_map,
  na_action = "keep",
  check_extreme = TRUE,
  extreme_action = "cap",
  extreme_rules = NULL,
  verbose = FALSE
)

head(a_cap$charlson_index)
```

## Outputs
- charlson_index integer total
- Rows drop only with na_action = "omit"; na_action = "error" aborts on missing.
- Optional bounds handling via check_extreme/extreme_action.

## Pitfalls and tips
- Indicators must be truly binary; non-binary values warn and can be capped if screening is on.
- Provide all 19 mappings; missing keys error.
- Age points are not included here; add separately if needed.
- Use extreme_rules to adjust bounds if your inputs differ.

## Validation ideas
- Set all indicators to 0: expected score 0.
- Single metastatic_cancer = 1, all else 0: expected score 6.
- diabetes = 1 and diab_comp = 1: expected diabetes contribution = 2 (max rule).

## See also
- metss and metabolic_risk_features for metabolic burden scoring.
- health_summary() to combine multiple indices.
