---
title: "Chronic Kidney Disease Stage"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ckd_stage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(purl = FALSE)
```

## Scope
Classify eGFR into KDIGO G1-G5, albuminuria (UACR mg/g) into A1-A3, and combine to KDIGO risk. If UACR is absent, albuminuria is NA and KDIGO risk assumes A1 for mapping.

## When to use
- You have eGFR (mL/min/1.73 m^2) and optionally UACR (mg/g) and want KDIGO staging.
- You need explicit handling of missing inputs and optional extreme-value screening.
- You want a lightweight KDIGO risk label without age adjustment.

## Requirements checklist
- Packages: HealthMarkers, dplyr (for display).
- Columns: eGFR required; UACR optional. Units: eGFR mL/min/1.73 m^2; UACR mg/g.
- Column map: list(eGFR = ..., UACR = ...) with UACR optional; mapping missing columns errors (except optional UACR).
- Row policy: na_action = keep (default), omit, error.
- Extreme screening: check_extreme + extreme_action (cap/NA/error); defaults cap eGFR 0-200, UACR 0-5000 when enabled.

## Load packages and example data
Example uses a 50-row slice of simulated data.

```{r}
library(HealthMarkers)
library(dplyr)

sim_path <- system.file("extdata", "simulated_hm_data.rds", package = "HealthMarkers")
sim <- readRDS(sim_path)
sim_small <- sim[1:50, c("eGFR", "UACR")]
```

## Map columns
UACR is optional; omit it from col_map if unavailable.

```{r}
col_map <- list(eGFR = "eGFR", UACR = "UACR")
```

## Quick start: stage CKD
Defaults keep rows with missing inputs and return NA stages for them.

```{r}
ckd_out <- ckd_stage(
  data = sim_small,
  col_map = col_map,
  na_action = "keep",
  check_extreme = FALSE,
  extreme_action = "cap"
)

head(ckd_out)
```

## Arguments that matter
- col_map: eGFR required; UACR optional. If UACR is mapped but missing in data, albuminuria is NA and a warning is issued.
- na_action: keep (default, retain rows; stages become NA), omit (drop rows with missing inputs), error (abort on missing).
- check_extreme: FALSE by default; TRUE enables bounds checks (eGFR 0-200; UACR 0-5000 mg/g by default).
- extreme_action: cap (default), NA, or error when check_extreme = TRUE.

## Handling missing inputs
- Non-numeric inputs are coerced; NA introduced are warned. Non-finite become NA.
- Missing eGFR yields NA stages unless na_action = omit/error. Missing UACR yields NA albuminuria; KDIGO risk treats missing albuminuria as A1 for mapping.

### Compare row policies
```{r}
demo <- sim_small[1:8, ]
demo$eGFR[3] <- NA

a_keep <- ckd_stage(demo, col_map, na_action = "keep", check_extreme = FALSE)
a_omit <- ckd_stage(demo, col_map, na_action = "omit", check_extreme = FALSE)

list(
  keep_rows = nrow(a_keep),
  omit_rows = nrow(a_omit),
  preview = head(a_keep)
)
```

## Extreme-value screening (optional)
Cap or error on out-of-range values.

```{r}
demo2 <- demo
demo2$UACR[5] <- 6000  # extreme albuminuria

a_cap <- ckd_stage(
  data = demo2,
  col_map = col_map,
  na_action = "keep",
  check_extreme = TRUE,
  extreme_action = "cap"
)

head(select(a_cap, Albuminuria_stage, KDIGO_risk))
```

## Outputs
- CKD_stage: G1, G2, G3a, G3b, G4, G5
- Albuminuria_stage: A1, A2, A3 (NA if UACR missing)
- KDIGO_risk: Low, Moderate, High, Very High (assumes A1 when albuminuria is missing for risk mapping)
Rows drop only with na_action = "omit" or when na_action = "error" aborts.

## Pitfalls and tips
- Use consistent units: eGFR mL/min/1.73 m^2; UACR mg/g. Do not mix mg/g with mg/mmol.
- If you do not have UACR, remove it from col_map; albuminuria stays NA and risk assumes A1.
- Use extreme_action = "error" for strict QC; cap for gentle cleaning.

## Validation ideas
- eGFR 92, UACR 10 -> G1/A1, KDIGO risk Low.
- eGFR 52, UACR 400 -> G3a/A3, KDIGO risk High.
- eGFR 22, UACR missing -> G4, Albuminuria NA, KDIGO risk Very High.

## See also
- kidney_kfre() for KFRE risk.
- renal_markers() for additional renal metrics.
