---
title: "Fasting insulin sensitivity indices"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{fasting_is}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(purl = FALSE)
```

## Scope
Compute 10 fasting insulin sensitivity/resistance indices from fasting glucose (mmol/L) and insulin (pmol/L). Internally converts glucose to mg/dL (x18) and insulin to muU/mL (/6). Handles column mapping, numeric coercion, NA policy, optional extreme screening, and optional normalization.

## When to use
- You have fasting glucose (mmol/L) and insulin (pmol/L) and need multiple sensitivity/resistance indices at once.
- You want explicit NA handling (keep/omit/error) plus optional extreme-value screening and normalization.
- Your column names differ from the defaults and need mapping.

## Inputs and requirements
- Required columns (mapped via `col_map`): `G0` (fasting glucose, mmol/L), `I0` (fasting insulin, pmol/L).
- Units are converted internally (mmol/L -> mg/dL; pmol/L -> muU/mL). Convert beforehand if your inputs are already in those units to avoid double-scaling.
- `na_action`: `keep` (default), `omit`, or `error` governs required-input missingness.
- `check_extreme`: optional screening of computed indices against `extreme_limit`; actions: `cap`, `NA`, or `error`.
- `normalize`: `none` (default) or `z`, `range`, `inverse`, `robust` applied after extreme handling.

## Load packages and data
Use a small subset of the simulated data (30-50 rows). Swap `sim_small` with your data frame in practice.

```{r}
library(HealthMarkers)
library(dplyr)

sim_path <- system.file("extdata", "simulated_hm_data.rds", package = "HealthMarkers")
sim <- readRDS(sim_path)
sim_small <- sim[1:50, ]
```

Expected units before internal conversion: glucose mmol/L, insulin pmol/L. If your units differ, convert first or adjust screening parameters.

## Column map (required)
Map the required inputs to your column names; you provide the right-hand-side strings (your column names). The left-hand keys (`G0`, `I0`) are fixed.

### What each input represents
- `G0`: fasting plasma glucose (mmol/L).
- `I0`: fasting insulin (pmol/L).

```{r}
col_map <- list(
  G0 = "G0",
  I0 = "I0"
)
```

If you use different labels (e.g., `glucose_mmol`, `insulin_pmol`), remap accordingly; any column names are fine as long as they hold the right values.

## Core calculation
Compute all indices and display only the newly calculated columns (not the full input data).

```{r}
fis <- fasting_is(
  data = sim_small,
  col_map = col_map,
  na_action = "keep",
  normalize = "none",
  check_extreme = FALSE,
  verbose = FALSE
)

fis_new <- setdiff(names(fis), names(sim_small))
head(select(fis, all_of(fis_new)))
```

The preview shows only the newly computed indices. `_inv` columns are inverted so larger negatives indicate greater insulin resistance; `QUICKI`, `Raynaud`, and `Isi_basal` decrease as resistance worsens. All rows are retained here because `na_action = "keep"`.

### Outputs at a glance
- `Fasting_inv`: inverted fasting insulin (higher negative = higher insulin)
- `Raynaud`: 40 / insulin
- `HOMA_IR_inv`, `HOMA_IR_rev_inv`: inverted HOMA formulations
- `FIRI`: fasting insulin resistance index
- `QUICKI`: insulin sensitivity index (higher = more sensitive)
- `Belfiore_basal`: 2 / (insulin x glucose + 1)
- `Ig_ratio_basal`: -insulin/glucose (inverted)
- `Isi_basal`: 10000 / (glucose x insulin)
- `Bennett`: 1 / (log insulin x log glucose)

## Inputs and units (quick reminder)
- Required columns (map via `col_map`): `G0` glucose (mmol/L), `I0` insulin (pmol/L).
- Internal conversions: glucose x18 -> mg/dL; insulin /6 -> muU/mL.
- Extremes: default `check_extreme = FALSE`; when TRUE, outputs are screened against `extreme_limit`. `extreme_action = "cap"` trims to +/-limit.
- Missingness: `na_action` controls row handling (`keep`/`omit`/`error`).
- Normalization: `normalize = "none"` by default; other options apply columnwise scaling after any extreme handling.

## Missing data and extremes
Show how row handling and extreme scanning behave on a tiny slice.

```{r}
demo <- sim_small[1:6, c("G0", "I0")]
demo$G0[2] <- NA

fis_keep <- fasting_is(demo, col_map, na_action = "keep", check_extreme = FALSE)
fis_omit <- fasting_is(demo, col_map, na_action = "omit", check_extreme = FALSE)

fis_extreme <- fasting_is(
  data = demo,
  col_map = col_map,
  na_action = "keep",
  check_extreme = TRUE,
  extreme_limit = 1e3,
  extreme_action = "cap",
  normalize = "none",
  verbose = FALSE
)

list(
  keep_rows = nrow(fis_keep),
  omit_rows = nrow(fis_omit),
  extreme_head = head(select(fis_extreme, all_of(fis_new)))
)
```

Here `na_action = "keep"` preserves all six rows (with NA-derived outputs), while `na_action = "omit"` drops the row with missing `G0`. Extreme scanning with `extreme_action = "cap"` would trim outputs exceeding `extreme_limit`; none were capped in this slice.

## Expectations
- Both required inputs must be mapped and present; missing mappings or columns abort.
- `na_action`: `keep` leaves NA-derived outputs; `omit` drops rows; `error` stops on any missing input.
- `check_extreme` screens outputs for magnitudes beyond `extreme_limit`; adjust `extreme_limit` and choose `extreme_action` (`cap`/`NA`/`error`).
- `normalize` is optional post-hoc scaling; set to `"none"` to preserve native scales.

## Tips
- Non-numeric inputs are coerced to numeric; coercion that introduces NAs triggers a warning.
- For pipelines, prefer `na_action = "omit"` or `"error"` to avoid silent gaps; use `"keep"` for exploratory summaries.
- Enable `check_extreme = TRUE` when working with noisy or simulated data; use `extreme_action = "error"` for strict QA.
- Set `normalize` to align scales across indices if feeding into distance-based models; leave as `"none"` for interpretability.
- See `?fasting_is` for full argument reference and diagnostics.
