---
title: "Frailty index"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{frailty_index}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(purl = FALSE)
```

## Scope
Compute a deficit-based frailty index (FI) via `di::di()` with HealthMarkers validation, NA/extreme handling, optional rescaling, and tidy output. By default, numeric columns are auto-selected as deficits when `cols = NULL` (excluding `age` if present). Higher FI indicates more deficits (roughly 0–1).

## When to use
- You have deficit variables (binary/logical or [0,1]-scaled) and need a standard FI quickly.
- You want guardrails: NA policy, out-of-range screening/capping, and optional rescaling before FI computation.
- You plan to consume the FI as a tibble (return = "data") or keep the original `di::di` list (return = "list").

## Inputs and requirements
- Dependency: the suggested package `di` must be installed; examples and calls are skipped/stop otherwise.
- Deficits: supply via `cols` (character vector) or allow auto-selection of numeric columns when `cols = NULL` (age is excluded if provided).
- Optional `age`: kept for plotting/bins but excluded from deficit selection.
- `na_action`: `ignore`/`warn`/`error` (legacy) plus aliases `keep` (like ignore) and `omit` (drops rows with any missing selected deficit).
- `check_extreme`: `NULL` (default; scans only when `rescale = FALSE`), `TRUE` (always scan <0 or >1), or `FALSE` (never scan). `extreme_action`: `warn`/`ignore`/`error`/`cap`/`NA`.
- `rescale`: default `TRUE`; when `FALSE`, range scanning is active by default (`check_extreme = NULL`).
- `return`: `list` (di object) or `data` (tibble with `di` plus selected deficits [+ age]).

## Load packages and data
These examples run only if the `di` package is installed.

```{r}
library(HealthMarkers)
library(dplyr)

has_di <- requireNamespace("di", quietly = TRUE)

demo_df <- data.frame(
  age = c(70, 75, 80, 78, 74, 69, 72, 77),
  d1  = c(0, 1, 1, 0, 0, 1, 0, 1),
  d2  = c(0.2, 0.8, 1.0, 0.4, 0.1, 0.9, 0.0, 0.7),
  d3  = c(TRUE, FALSE, TRUE, TRUE, FALSE, TRUE, FALSE, TRUE)
)
```

Expected inputs: deficits ideally scaled to [0,1] (or logical/binary). `age` is optional; when present it is excluded from auto-selected deficits.

## Columns to use
You can supply `cols` or let the function auto-select numeric deficits when `cols = NULL`. Here we name them explicitly and keep `age` for optional plotting.

```{r}
cols <- c("d1", "d2", "d3")
```

## Core calculation
Return a tidy tibble (`return = "data"`) so we can preview FI alongside deficits.

```{r}
if (has_di) {
  fi_tbl <- frailty_index(
    data = demo_df,
    cols = cols,
    age = "age",
    rescale = TRUE,
    na_action = "keep",
    check_extreme = NULL,   # legacy: scan only if rescale = FALSE
    extreme_action = "warn",
    return = "data",
    verbose = FALSE
  )

  head(select(fi_tbl, di, all_of(cols)))
} else {
  message("Install the 'di' package to run the examples.")
}
```

FI ranges from 0 (no deficits) toward 1 (all deficits present). Here `rescale = TRUE` allows non-binary numeric inputs; `na_action = "keep"` retains rows even if deficits contain `NA` (none in this slice).

## Inputs and behavior (quick reminder)
- Deficits: binary/logical or [0,1]-scaled; supply via `data` and `cols` (or let auto-select numeric deficits when `cols = NULL`).
- `age` (optional): excluded from auto-selected deficits; useful only for plotting/bins.
- `na_action`: `keep`/`ignore` retain rows with missing deficits; `omit` drops rows; `error` stops on any missing selected deficit; `warn` emits warnings.
- `check_extreme`: `TRUE` scans for <0 or >1; `NULL` scans only when `rescale = FALSE`; `FALSE` skips scanning.
- `extreme_action`: `cap` truncates to [0,1]; `NA` blanks; `warn`/`ignore` leave as-is; `error` stops.
- `return`: `data` yields a tibble (di + selected deficits [+ age]); `list` returns the `di::di` object.

## Missing data and extremes
Demonstrate row dropping, capping, and tidy return.

```{r}
if (has_di) {
  demo_miss <- demo_df
  demo_miss$d2[3] <- NA       # introduce missing
  demo_miss$d1[2] <- 1.2      # out-of-range to be capped

  fi_keep <- frailty_index(
    data = demo_miss,
    cols = cols,
    age = "age",
    rescale = FALSE,          # triggers range scan when check_extreme is NULL
    na_action = "keep",
    check_extreme = NULL,
    extreme_action = "cap",
    return = "data",
    verbose = FALSE
  )

  fi_omit <- frailty_index(
    data = demo_miss,
    cols = cols,
    age = "age",
    rescale = FALSE,
    na_action = "omit",
    check_extreme = NULL,
    extreme_action = "cap",
    return = "data",
    verbose = FALSE
  )

  list(
    keep_rows = nrow(fi_keep),
    omit_rows = nrow(fi_omit),
    capped_example = head(select(fi_keep, di, all_of(cols)))
  )
} else {
  message("Install the 'di' package to run the examples.")
}
```

With `na_action = "keep"`, all rows remain (missing deficits propagate to FI). With `na_action = "omit"`, rows with any missing selected deficit are dropped. `extreme_action = "cap"` trims out-of-range deficits (e.g., 1.2) into [0,1] before FI calculation.

## Expectations
- `di` package must be installed; otherwise calls stop (examples skip when missing).
- Provide deficits via `cols` or rely on auto-selection of numeric columns (age excluded when present).
- Choose `na_action` to control row retention vs. dropping on missing deficits.
- Use `check_extreme`/`extreme_action` to scan and cap or blank values outside [0,1] when desired.
- `return = "data"` gives a tibble (`di` + selected deficits [+ age]); `return = "list"` returns the raw `di::di` output.

## Tips
- Keep deficits close to [0,1]; `rescale = TRUE` can normalize mild deviations, but extreme outliers may still need capping.
- Use `extreme_action = "cap"` with `check_extreme = TRUE` (or `rescale = FALSE` + `check_extreme = NULL`) when your deficits might stray outside [0,1].
- Choose `na_action = "omit"` or `"error"` for stricter pipelines; `keep`/`ignore` for exploratory summaries.
- `return = "data"` pairs FI with the selected deficits for quick binding into downstream tibbles.
- Set `verbose = TRUE` to see the selected deficits, NA scan counts, and completion range.
