---
title: "Getting Started with HealthMarkers"
author: "HealthMarkers Team"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with HealthMarkers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE)
knitr::opts_knit$set(purl = FALSE)
```

## Overview
Load the bundled simulated dataset, pick which markers to compute, and view only the outputs you care about. All examples use shipped data; swap in your data when ready.

## When to use
- You want a quick tour of core helpers and the `all_health_markers()` orchestrator.
- You need to see column inference vs explicit mapping for your data.
- You prefer ready-to-run code snippets (no optional dependencies required for these examples).

## What the package offers
- Many clinical/metabolic markers: glycemic, lipid, renal, inflammatory, liver, anthropometric/obesity, atherogenic indices, frailty/Charlson, sweat/urine panels, micronutrients, pulmonary indices, and more.
- Single-purpose helpers (e.g., `cvd_marker_aip()`, `sweat_markers()`, `fasting_is()`) plus `all_health_markers()` to run multiple groups at once.
- Column mapping diagnostics with `hm_col_report()` and override via `col_map` when your labels differ.
- Light missing-data utilities (`impute_missing()`) to prep data before computing markers.

## Quick starts
- Targeted helper: `cvd_marker_aip(sim_small)` or `renal_markers(sim_small)`.
- Multiple groups: `all_health_markers(sim_small, which = c("glycemic", "lipid"))`.
- Everything: `all_health_markers(sim_small, which = "all")` (wide output; filter afterward).
- Insulin sensitivity/resistance: `fasting_is()`, `ogtt_is()`, `adipo_is()`, or include insulin groups in `all_health_markers()` with a `col_map` for glucose/insulin.

## Common helpers (at a glance)
- `all_health_markers(data, which = ...)` — run multiple groups or all markers at once.
- `cvd_marker_aip(data, col_map = list(TG = "TG", HDL_c = "HDL_c"))` — atherogenic index of plasma.
- `renal_markers(data)` — renal and eGFR-related outputs.
- `glycemic_markers(data)` — glycemic metrics; pair with `fasting_is()` / `ogtt_is()` / `adipo_is()` for insulin indices.
- `sweat_markers(data, check_extreme = TRUE)` and `urine_markers(data)` — alternate biofluid panels.

## Load the package and simulated data
Inputs: the package plus the bundled RDS file at `inst/extdata/simulated_hm_data.rds`.

Outputs: two data frames—`sim` (full) and `sim_small` (first 30 rows for fast examples).

```{r}
if (!requireNamespace("HealthMarkers", quietly = TRUE)) {
  if (requireNamespace("pkgload", quietly = TRUE)) {
    pkgload::load_all("..")
  } else {
    stop("Install HealthMarkers (or pkgload for development) before knitting.", call. = FALSE)
  }
}
library(HealthMarkers)

sim_path <- system.file("extdata", "simulated_hm_data.rds", package = "HealthMarkers")
sim <- readRDS(sim_path)

# Work with a small subset using slice_head
sim_small <- dplyr::slice_head(sim, n = 30)

dim(sim_small)
```

## Minimal example: broad panels
Input: `sim_small` with inferred columns.

What it does: `all_health_markers()` auto-detects columns and computes the requested groups.

Output: a tibble of computed markers. Below we print only the new columns.

```{r}
hm_out <- all_health_markers(
  data = sim_small,
  which = c("glycemic", "lipid", "renal", "inflammatory"),
  verbose = FALSE
)

hm_new_cols <- setdiff(names(hm_out), names(sim_small))
head(dplyr::select(hm_out, dplyr::all_of(hm_new_cols)))

# New columns added (names only):
head(hm_new_cols)
```

## Data readiness: columns at a glance
Most functions infer columns automatically. If your names differ, provide a `col_map`. Typical defaults:

| Panel | Typical columns | Notes |
| --- | --- | --- |
| Glycemic | `FPG`, `insulin_fasting`, `HbA1c` | Uses fasting and HbA1c where available |
| Lipid | `TG`, `HDL_c`, `LDL_c`, `TC` | Supports TG/HDL aliasing for AIP |
| Renal | `creatinine`, `uacr`, `age`, `sex` | Infers eGFR and kidney markers |
| Blood pressure (used by CVD risk helpers) | `SBP`, `DBP` | Inputs to CVD risk functions; not a standalone `all_health_markers()` group |
| Anthropometrics | `height`, `weight`, `BMI`, `waist_circumference` | Supports z-score and obesity indices |

### The easiest way: `hm_col_report()`

Call `hm_col_report()` **before running any computation** to see a full
column-mapping report for your data:

```{r, eval = FALSE}
# See which columns are matched and how — no computation performed
hm_col_report(my_data)

# Show all 258 keys, including unmatched ones
hm_col_report(my_data, show_unmatched = TRUE)

# Capture the matched col_map to pass directly to any function
cm <- hm_col_report(my_data, verbose = FALSE)
all_health_markers(my_data, col_map = cm)
```

The report shows:
- **key** — the internal name all functions use
- **data_column** — the matched column in your data
- **how matched** — `exact`, `case-insensitive`, `col contains synonym`, or `synonym contains col`
- A ready-to-paste `col_map` template for any keys not found

Column inference is usually enough; if your data use different labels, provide a `col_map` override.

## Column inference vs explicit mapping
Input: the same data but with explicit column aliases.

What it does: provides TG/HDL names to compute AIP when your columns differ.

Output: a tibble with the AIP result.

```{r}
lipid_cols <- list(TG = "TG", HDL_c = "HDL_c")

aip <- cvd_marker_aip(sim_small, col_map = lipid_cols, na_action = "keep")
aip_new <- setdiff(names(aip), names(sim_small))
head(dplyr::select(aip, dplyr::all_of(aip_new)))
```

## Handling missingness
Most functions accept `na_action`. This shows how outputs differ when required inputs are missing.

```{r}
missing_demo <- sim_small[1:5, c("TG", "HDL_c")]
missing_demo$TG[2] <- NA

# Keep rows (produces NA where inputs are missing)
cvd_marker_aip(missing_demo, col_map = lipid_cols, na_action = "keep")

# Drop rows with required NA
cvd_marker_aip(missing_demo, col_map = lipid_cols, na_action = "omit")
```

> **`na_action` aliases:** `"ignore"` is a backward-compatible alias for `"keep"` (same
> behavior; retained so older code continues to work). `"warn"` is also an alias for
> `"keep"` that additionally emits a missingness warning. If you are reading another
> function's help page and see `"ignore"` listed first in the choices, it behaves
> identically to `"keep"`.

If you want to impute before computing markers:

```{r, eval = FALSE}
imputed <- impute_missing(sim_small, method = "mice", m = 1, maxit = 3)
head(imputed)
```

### Imputation options (what happens under the hood)
- `impute_missing()` is a preprocessing helper; marker functions do not impute internally. Run it first, then call calculators.
- Methods:
  - `method = "mice"`: multiple imputation by chained equations via **mice**; control `m` (imputations) and `maxit` (iterations).
  - `method = "missForest"`: nonparametric random-forest imputation via **missForest**; good for mixed data types.
  - `method = "mean" | "median" | "constant"`: simple deterministic fills for numeric columns.
- Output: a data frame with imputed values replacing missing entries; then pass it to `all_health_markers()` or any helper (e.g., `cvd_marker_aip()`, `renal_markers()`).

## Extreme value checks
Scan for extreme sweat values and warn without stopping. Output: sweat markers with any warnings emitted.

```{r}
sweat <- sweat_markers(sim_small, check_extreme = TRUE, extreme_action = "warn")
sweat_new <- setdiff(names(sweat), names(sim_small))
head(dplyr::select(sweat, dplyr::all_of(sweat_new)))
```

## Verbose mode: skipped optional markers

When a function supports optional inputs (columns that extend outputs when present),
set `verbose = TRUE` to see which optional markers were not computed because their
columns were absent from the data. This helps you distinguish "column not found"
from "marker intentionally absent".

```{r}
# Provide only required columns; optional columns are absent
glyc_min <- sim_small[, c("HDL_c", "TG", "BMI")]

# verbose = TRUE lists which optional markers were skipped
glyc_out <- glycemic_markers(glyc_min, verbose = TRUE)
```

The informational message lists all optional keys (glucose, HbA1c, C_peptide, G0,
I0, leptin, adiponectin) that had no matching column, so the caller knows which
derived metrics (e.g., prediabetes flag, HOMA_CP, LAR) will be `NA` in the output.
Once you add those columns to your data, the message disappears and the metrics
are computed automatically.

## Common functions
### Cardio-metabolic 
Input: `sim_small`; groups: glycemic, lipid, renal, inflammatory.

Output: a tibble with those markers (first rows shown).

```{r}
cardio_panel <- all_health_markers(
  data = sim_small,
  which = c("glycemic", "lipid", "renal", "inflammatory"),
  verbose = FALSE
)

cardio_new <- setdiff(names(cardio_panel), names(sim_small))
head(dplyr::select(cardio_panel, dplyr::all_of(cardio_new)))
```

### Urine markers 
Input: `sim_small`; output: urine markers with rows containing required inputs.
```{r}
urine <- urine_markers(sim_small, na_action = "omit")
urine_new <- setdiff(names(urine), names(sim_small))
head(dplyr::select(urine, dplyr::all_of(urine_new)))
```

### All markers (wide; optional)
To compute every available group on your data (can produce many columns), set `which = "all"`. Here we keep it example-only to avoid long output:

```{r, eval = FALSE}
all_out <- all_health_markers(data = sim_small, which = "all", verbose = FALSE)
head(setdiff(names(all_out), names(sim_small)))
```

## Export-ready output
Select columns and write to a CSV (disabled during vignette build):

```{r, eval = FALSE}
export_cols <- c("aip", "egfr_ckd_epi", "homa_ir")
out_path <- tempfile("healthmarkers_export_", fileext = ".csv")
hm_out_subset <- dplyr::select(cardio_panel, dplyr::any_of(export_cols))
utils::write.csv(hm_out_subset, out_path, row.names = FALSE)
out_path
```

## Next steps
- Explore focused vignettes for insulin sensitivity, cardio-renal panels, and frailty.
- See individual help pages (e.g., `?fasting_is`, `?cvd_risk`, `?frailty_index`) for detailed arguments and references.

> **All 46 domain vignettes** are available on the package website — only 12 core vignettes are bundled with the CRAN package.  
> Browse the full collection at <https://sufyansuleman.github.io/HealthMarkers/articles/>

