---
title: "Glycemic and metabolic markers"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{glycemic_markers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(purl = FALSE)
```

## Scope
Compute SPISE, METS_IR, prediabetes/diabetes flags (HbA1c), HOMA_CP, LAR, ASI, and TyG_index from fasting labs and anthropometry. Required inputs: HDL_c (mmol/L), TG (mmol/L), BMI (kg/m^2); optional glucose/HbA1c/C-peptide/insulin/leptin/adiponectin if present. Uses the packaged simulated data for runnable examples.

## When to use
- You have fasting lipids and BMI and want quick insulin resistance/sensitivity proxies.
- You can supply glucose/HbA1c/C-peptide/insulin/adipokines when available to enrich outputs.
- You need explicit NA handling and optional extreme-value screening.

## Inputs and requirements
- Required (mapped via `col_map`): `HDL_c` (mmol/L), `TG` (mmol/L), `BMI` (kg/m^2).
- Optional: `glucose`, `HbA1c`, `C_peptide`, `G0`, `I0`, `leptin`, `adiponectin`.
- Unit notes: TyG converts TG to mg/dL (x88.57) and glucose to mg/dL (x18); HOMA_CP uses `(G0 * (C_peptide/6)) / 22.5`.
- `na_action`: `keep`/`ignore`/`warn` retain rows with NA outputs; `omit` drops rows with missing used inputs; `error` aborts.
- `check_extreme`: when TRUE, scans inputs with default broad ranges; `extreme_action` = `warn`/`cap`/`NA`/`error`/`ignore`.

## Load packages and data
Use a small subset of the simulated data (30-50 rows). Swap `sim_small` with your data frame in practice.

```{r}
library(HealthMarkers)

sim_path <- system.file("extdata", "simulated_hm_data.rds", package = "HealthMarkers")
sim <- readRDS(sim_path)
sim_small <- dplyr::slice_head(sim, n = 30)
```

## Column map (required)
Map required and optional inputs to your column names; left-hand keys are fixed.

### What each input represents
- `HDL_c`: HDL cholesterol (mmol/L)  required
- `TG`: triglycerides (mmol/L)  required
- `BMI`: body mass index (kg/m^2)  required
- `glucose`: fasting glucose (mmol/L)
- `HbA1c`: hemoglobin A1c (mmol/mol)
- `C_peptide`: C-peptide (pmol/L)
- `G0`: fasting glucose (mmol/L) for HOMA_CP
- `I0`: fasting insulin (pmol/L) for ASI
- `leptin`: leptin (ng/mL)
- `adiponectin`: adiponectin (ng/mL)

```{r}
col_map <- list(
  HDL_c = "HDL_c",
  TG = "TG",
  BMI = "BMI",
  glucose = "glucose",
  HbA1c = "HbA1c",
  C_peptide = "C_peptide",
  G0 = "G0",
  I0 = "I0",
  leptin = "leptin",
  adiponectin = "adiponectin"
)
```

## Core calculation
Compute all markers and show only newly created columns.

```{r}
gm <- glycemic_markers(
  data     = sim_small,
  col_map  = col_map,
  na_action = "keep",
  verbose  = FALSE
)

gm_new <- setdiff(names(gm), names(sim_small))
head(dplyr::select(gm, dplyr::all_of(gm_new)))
```

Interpretation: SPISE and METS_IR summarize insulin sensitivity/resistance; TyG_index rises with higher TG/glucose; HOMA_CP approximates insulin resistance via C-peptide and glucose; LAR and ASI relate adipokines to insulin; HbA1c flags populate `prediabetes`/`diabetes` when available. Rows are retained here because `na_action = "keep"`.

## Inputs and units (quick reminder)
- Required: `HDL_c` (mmol/L), `TG` (mmol/L), `BMI` (kg/m^2).
  Auto-derived: BMI from `weight` (kg) + `height` (m/cm) if BMI absent.
- Optional if present/mapped: `glucose`, `HbA1c`, `C_peptide`, `G0`, `I0`, `leptin`, `adiponectin`.
  Auto-derived: `glucose` from `G0` (or vice versa) if one is absent.
- TyG converts TG to mg/dL (88.57) and glucose to mg/dL (18) internally.
- HOMA_CP uses `(G0 * (C_peptide/6)) / 22.5`; verify your units for C-peptide.
- Missingness: `na_action` controls row handling (`keep`/`omit`/`error`/`warn`/`ignore`).

## Pre-computation example

Providing `weight` and `height` instead of `BMI` works automatically:

```{r}
df_no_bmi <- data.frame(
  HDL_c   = c(1.0, 1.2),
  TG      = c(1.3, 1.5),
  weight  = c(70, 85),    # kg
  height  = c(175, 180),  # cm
  glucose = c(5.5, 6.0)
)
gm_precomp <- glycemic_markers(df_no_bmi, verbose = FALSE)
gm_precomp$SPISE  # computed via auto-derived BMI
```

Similarly, providing `G0` without `glucose` enables METS_IR and TyG_index via alias:

```{r}
df_g0_only <- data.frame(
  HDL_c = 1.2, TG = 1.3, BMI = 24, G0 = 5.5
)
gm_alias <- glycemic_markers(df_g0_only, verbose = FALSE)
c(METS_IR = gm_alias$METS_IR, TyG_index = gm_alias$TyG_index)
```

## Missing data handling
Show row handling on a small slice.

```{r}
demo <- sim_small[1:6, c("HDL_c", "TG", "BMI")]
demo$glucose   <- c(5.5, 6.2, 7.0, 5.8, 6.0, 5.4)
demo$HbA1c     <- c(42, NA, 41, 45, 44, 40)
demo$C_peptide <- c(300, 450, 500, 400, 380, 360)
demo$G0        <- demo$glucose
demo$I0        <- c(60, 90, 120, 80, 75, 70)
demo$leptin    <- c(10, 12, 15, 11, 9, 13)
demo$adiponectin <- c(8, 7, 6, 9, 10, 7)

cm2 <- list(
  HDL_c = "HDL_c", TG = "TG", BMI = "BMI",
  glucose = "glucose", HbA1c = "HbA1c",
  C_peptide = "C_peptide", G0 = "G0", I0 = "I0",
  leptin = "leptin", adiponectin = "adiponectin"
)

gm_keep <- glycemic_markers(data = demo, col_map = cm2, na_action = "keep",  verbose = FALSE)
gm_omit <- glycemic_markers(data = demo, col_map = cm2, na_action = "omit",  verbose = FALSE)

list(keep_rows = nrow(gm_keep), omit_rows = nrow(gm_omit))
```

`na_action = "keep"` preserves all rows (missing inputs flow to `NA` outputs); `na_action = "omit"` drops rows with missing/invalid used inputs.

## Expectations
- Required inputs must be present/mapped or derivable (BMI from weight/height); missing required columns abort.
- `na_action` governs whether missing/invalid inputs drop rows, warn, or error.
- Optional columns drive optional markers; absent inputs yield `NA` in their markers.
- Range checks are informational only — values outside physiological plausibility ranges are noted in `verbose` output but not altered.

## Verbose diagnostics
Set `verbose = TRUE` (the default) to surface five structured messages: column mapping, optional inputs (what is present/missing and which indices they affect), pre-computation actions, the full list of markers being computed with their inputs, and a per-column results summary.

```{r, message=TRUE}
df_v <- data.frame(HDL_c = 1.0, TG = 1.3, BMI = 24, glucose = 5.6)
glycem_out <- glycemic_markers(
  df_v,
  col_map = list(HDL_c = "HDL_c", TG = "TG", BMI = "BMI", glucose = "glucose"),
  verbose = TRUE
)
```

## Tips
- Start with fasting lipids and BMI (or weight/height); add glucose/HbA1c/C-peptide/insulin/adipokines for richer markers.
- Provide `weight` and `height` instead of BMI when BMI is not directly available — it is computed automatically.
- Convert units beforehand if your data are not in the expected mmol/L/pmol/L/ng/mL units.
- For stricter pipelines, use `na_action = "omit"` or `"error"`; keep/warn for exploratory work.
- Use `verbose = TRUE` (default) for detailed output messages and diagnostics.
- See `?glycemic_markers` for full argument details and marker definitions.
