title: "Hormone marker ratios"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{hormone_markers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(purl = FALSE)
```

## Scope
Compute nine hormone ratios with built-in QA: `FAI`, `LH_FSH`, `E2_P`, `T3_T4`, `ARR`, `Ins_Glu`, `GH_IGF1`, `PRL_T`, and `CAR_slope`. Inputs are coerced to numeric, non-finite become `NA`, and optional extreme screening can warn, cap, blank, or error on out-of-range labs.

## When to use
- You have a complete hormone panel (mapped via `col_map`) and need quick ratio features.
- You want explicit NA handling and optional extreme-value screening to avoid crashes on bad labs.
- You need outputs that preserve row counts unless you explicitly `omit` missing rows.

## Inputs and requirements
- All keys are required: total testosterone, SHBG, LH, FSH, estradiol, progesterone, free T3, free T4, aldosterone, renin, insulin, glucagon, GH, IGF1, prolactin, cortisol_0, cortisol_30.
- `col_map`: map each required key to your column names; missing mappings or columns abort.
- Units: keep units consistent within each ratio (e.g., cortisol nmol/L; aldosterone/renin in matching units).
- `na_action`: `keep`/`ignore` retain rows with NA outputs; `omit` drops rows with missing required inputs; `error` aborts.
- `check_extreme`: optional; `extreme_action` = `warn`/`cap`/`NA`/`error`/`ignore`.

## Load packages and demo data
Use a small synthetic tibble (covers all required inputs) so the chunk runs anywhere. Replace `horm_demo` with your lab data.

```{r}
library(HealthMarkers)
library(tibble)

horm_demo <- tibble(
  tt = c(5.0, 7.0, 4.5, 6.2),
  shbg = c(30, 40, 28, 35),
  lh = c(6, 7, 5, 6),
  fsh = c(8, 9, 7, 8),
  e2 = c(120, 140, 110, 135),
  p4 = c(2, 3, 1.8, 2.5),
  ft3 = c(4.5, 4.8, 4.3, 4.6),
  ft4 = c(16, 15, 17, 16),
  aldo = c(150, 120, 140, 130),
  renin = c(12, 10, 11, 9),
  ins = c(15, 20, 18, 17),
  gluc = c(80, 70, 75, 78),
  gh = c(1.2, 0.9, 1.0, 1.1),
  igf1 = c(180, 160, 170, 165),
  prl = c(10, 12, 11, 9),
  cort0 = c(300, 280, 290, 310),
  cort30 = c(480, 450, 470, 500)
)
```

## Column map (required)
Map every required key to your column names; all are mandatory for computation.

- `total_testosterone`, `SHBG`
- `LH`, `FSH`
- `estradiol`, `progesterone`
- `free_T3`, `free_T4`
- `aldosterone`, `renin`
- `insulin`, `glucagon`
- `GH`, `IGF1`
- `prolactin`
- `cortisol_0`, `cortisol_30` (nmol/L; slope uses the 30-minute delta / 30)

```{r}
col_map <- list(
  total_testosterone = "tt",
  SHBG = "shbg",
  LH = "lh",
  FSH = "fsh",
  estradiol = "e2",
  progesterone = "p4",
  free_T3 = "ft3",
  free_T4 = "ft4",
  aldosterone = "aldo",
  renin = "renin",
  insulin = "ins",
  glucagon = "gluc",
  GH = "gh",
  IGF1 = "igf1",
  prolactin = "prl",
  cortisol_0 = "cort0",
  cortisol_30 = "cort30"
)
```

## Core calculation
Compute all ratios and preview the outputs. Non-finite values are set to `NA`; rows are retained when `na_action = "keep"`/`"ignore"`.

```{r}
hm_out <- hormone_markers(
  data = horm_demo,
  col_map = col_map,
  na_action = "keep",
  check_extreme = FALSE,
  verbose = FALSE
)

hm_out
```

## Missing data and row handling
Show the difference between keeping rows (propagating `NA`) and omitting rows when required inputs are missing.

```{r}
horm_demo_na <- horm_demo
horm_demo_na$ft3[2] <- NA

keep_res <- hormone_markers(
  data = horm_demo_na,
  col_map = col_map,
  na_action = "keep",
  check_extreme = FALSE
)

omit_res <- hormone_markers(
  data = horm_demo_na,
  col_map = col_map,
  na_action = "omit",
  check_extreme = FALSE
)

list(keep_rows = nrow(keep_res), omit_rows = nrow(omit_res))
```

## Extreme screening
Enable range scanning and cap out-of-range inputs into allowed bounds.

```{r}
horm_demo_ext <- horm_demo
horm_demo_ext$aldo[1] <- 2000  # intentionally extreme

cap_res <- hormone_markers(
  data = horm_demo_ext,
  col_map = col_map,
  na_action = "keep",
  check_extreme = TRUE,
  extreme_action = "cap",
  verbose = TRUE
)

cap_res[1, c("ARR", "FAI")]
```

`extreme_action = "cap"` trims inputs into the allowed ranges; `warn` or `error` emit warnings or abort instead.

## Expectations
- All required inputs must be mapped and present; missing mappings/columns abort.
- `na_action`: `keep`/`ignore` retains rows with `NA` outputs; `omit` drops rows with missing required inputs; `error` stops on missingness.
- Extreme screening is optional; when enabled, choose how to handle flagged values with `extreme_action`.
- Outputs are numeric ratios; divisions by zero/non-finite are converted to `NA`.

## Tips
- Keep units consistent across rows (e.g., cortisol nmol/L, aldosterone/renin in matching units) to make ratios meaningful.
- Use `na_action = "omit"` when you prefer complete-case ratios; keep/ignore if you want row counts preserved.
- Tighten `extreme_rules` if your lab reference ranges are narrower than the broad defaults.
- Turn on `verbose = TRUE` to see coercion, missingness scan, and extreme handling summaries.
