---
title: "Imputing missing values"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{impute_missing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(purl = FALSE)
```

## Scope
Three imputation helpers for preparing data before marker calculations:

- `impute_missing()` — fast deterministic imputation (mean, median, zero, or a constant). Operates column by column over all numeric columns with at least one `NA`.
- `impute_mice()` — multiple imputation via `mice::mice()`. Requires the **mice** package.
- `impute_missforest()` — random-forest imputation via `missForest::missForest()`. Requires the **missForest** package. Falls back to mean imputation if the forest fails.

All three: (1) leave non-numeric columns untouched, (2) respect an optional `cols` argument to limit the columns imputed, (3) warn on high-missingness columns, and (4) emit structured progress messages when `verbose = TRUE`.

## Load packages and demo data

```{r}
library(HealthMarkers)
library(tibble)

df <- tibble::tibble(
  glucose  = c(5.2, NA, 6.0, 5.8, NA),
  insulin  = c(60,  80, NA,  90, 110),
  age      = c(45,  52, 38,  60,  NA),
  sex      = c("M", "F", "M", "F", "M")   # non-numeric; never touched
)
```

## `impute_missing()` — deterministic methods

```{r}
out_mean <- impute_missing(df, method = "mean")
out_mean
```

Replace NAs column-by-column with the column mean. `sex` is untouched.

### Other methods

```{r}
# Median
out_med <- suppressWarnings(impute_missing(df, method = "median"))
# Zero
out_zero <- suppressWarnings(impute_missing(df, method = "zero", cols = "glucose"))
# Constant
out_const <- suppressWarnings(impute_missing(df, method = "constant", constant = -1))

list(
  median_glucose  = out_med$glucose,
  zero_glucose    = out_zero$glucose,
  const_glucose   = out_const$glucose
)
```

### Target specific columns

```{r}
# Only impute insulin; leave glucose NA
out_sel <- suppressWarnings(impute_missing(df, method = "mean", cols = "insulin"))
list(glucose = out_sel$glucose, insulin = out_sel$insulin)
```

## `impute_mice()` — multiple imputation

Requires at least two numeric columns with missing values. Returns one completed dataset from `m` imputations.

```{r}
if (requireNamespace("mice", quietly = TRUE)) {
  out_mice <- suppressWarnings(impute_mice(df, m = 5))
  print(out_mice)
} else {
  message("mice package not installed; skipping example.")
}
```

## `impute_missforest()` — random forest imputation

```{r}
if (requireNamespace("missForest", quietly = TRUE)) {
  out_rf <- suppressWarnings(impute_missforest(df, ntree = 50))
  print(out_rf)
} else {
  message("missForest package not installed; skipping example.")
}
```

If missForest errors (e.g., too few unique values), a mean-imputation fallback is applied automatically.

## Handling high-missingness columns

All three functions warn when a numeric column has missing values above `na_warn_prop` (default 0.2 = 20 %). The warning carries class `healthmarkers_impute_warn_high_missing`.

```{r, warning=TRUE}
df_hi <- tibble::tibble(x = c(1, NA, NA, NA), y = c(2, NA, 4, 5))
suppressWarnings(
  tryCatch(
    impute_missing(df_hi, method = "mean", verbose = FALSE),
    warning = function(w) { message("High-missingness warning: ", conditionMessage(w)); NULL }
  )
)
```

## Verbose diagnostics

Set `verbose = TRUE` (and `healthmarkers.verbose = "inform"`) to surface structured messages for each call: preparing inputs and a results summary.

```{r, message=TRUE}
old_opt <- options(healthmarkers.verbose = "inform")

df_v <- tibble::tibble(a = c(1, NA, 3), b = c(4, 5, NA), c = c(NA, 2, 2))
suppressWarnings(impute_missing(df_v, method = "mean", verbose = TRUE))

options(old_opt)
```

The results message reports how many values were imputed per column:
`impute_missing(): results: imputed N values across M columns [col1=n1, col2=n2, ...]`

## Tips
- Run imputation **before** calling marker functions when data-level NA handling is insufficient (e.g., when you need a complete matrix for PCA or multivariate ML upstream).
- Prefer `impute_missing(method = "median")` for skewed biomarker distributions.
- `impute_mice()` and `impute_missforest()` provide MAR-robust estimates at the cost of speed; use them for final analysis datasets, not exploratory runs.
- Non-numeric columns (factors, characters) are always passed through unchanged.
