---
title: "Inflammatory markers"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{inflammatory_markers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(purl = FALSE)
```

## Scope
Compute inflammatory ratios/panels. Panels:
- Classic: NLR, PLR, LMR, dNLR, SII, SIRI, AISI, CRP_category.
- Eosinophil: NLR, PLR, LMR, NER, SII, SIRI, PIV, CLR, CAR, PCR, mGPS, ESR passthrough.
- Both: union of classic + eos metrics.
Inputs are coerced to numeric; non-finite become `NA`. Zero denominators warn. Optional extreme handling can warn/cap/blank/error.

## Load packages and demo data
Synthetic counts to run both panels.

```{r}
library(HealthMarkers)
library(tibble)

df <- tibble(
  neutrophils = c(4, 2, 3.5),
  lymphocytes = c(2, 1.0, 1.5),
  monocytes = c(0.5, 0.3, 0.4),
  platelets = c(200, 150, 180),
  WBC = c(7, 4.5, 6),
  CRP = c(2.5, 0.8, 12),
  albumin = c(40, 42, 33),
  eosinophils = c(0.2, 0.1, 0.15),
  ESR = c(12, 15, 18)
)
```

## Column map (required keys per panel)
- Classic: `neutrophils`, `lymphocytes` (monocytes/platelets/WBC/CRP improve coverage).
- Eos: classic keys plus `eosinophils`, `CRP`, `albumin` (for CLR/CAR/mGPS); `ESR` optional passthrough.

```{r}
cm <- list(
  neutrophils = "neutrophils",
  lymphocytes = "lymphocytes",
  monocytes = "monocytes",
  platelets = "platelets",
  WBC = "WBC",
  CRP = "CRP",
  albumin = "albumin",
  eosinophils = "eosinophils",
  ESR = "ESR"
)
```

## Core calculation (classic panel)

```{r}
classic_out <- inflammatory_markers(
  data = df,
  col_map = cm,
  panel = "classic",
  na_action = "keep",
  check_extreme = FALSE,
  verbose = FALSE
)

classic_out
```

## Eosinophil panel with extreme capping

```{r}
df_ext <- df
df_ext$CRP[3] <- 400  # intentional extreme

eos_out <- inflammatory_markers(
  data = df_ext,
  col_map = cm,
  panel = "eos",
  na_action = "keep",
  check_extreme = TRUE,
  extreme_action = "cap",
  verbose = TRUE
)

eos_out
```

`extreme_action = "cap"` trims out-of-range inputs; `warn` only warns; `error` aborts; `NA` blanks flagged values.

## Missing data handling
`na_action = "keep"` preserves rows with `NA` outputs; `omit` drops rows with missing required inputs; `error` aborts if any required input is missing/non-finite.

```{r}
df_na <- df
df_na$lymphocytes[2] <- NA

keep_out <- inflammatory_markers(df_na, cm, panel = "classic", na_action = "keep")
omit_out <- inflammatory_markers(df_na, cm, panel = "classic", na_action = "omit")

list(keep_rows = nrow(keep_out), omit_rows = nrow(omit_out))
```

## Expectations
- Provide required keys per panel; missing mappings abort.
- Inputs are coerced to numeric; non-finite become `NA`; zero denominators warn.
- Extreme scan uses broad defaults for blood counts/CRP/albumin; adjust via `extreme_action` and built-in ranges.
- Panel selection: `auto` uses eos if `eosinophils` mapping is present; `both` returns the union.

## Tips
- If your lymphocyte/eosinophil counts can be zero, consider `na_action = "keep"` to avoid dropping rows; zeros still trigger divide-by-zero warnings.
- Tighten ranges when using capping to match lab references.
- If your data use different units (e.g., CRP mg/dL), convert before calling to keep ratios interpretable.
