title: "Inflammatory age (iAge proxy)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{inflammatory_age}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(purl = FALSE)
```

## Scope
Compute a simplified inflammatory age proxy (`iAge`) as a weighted sum of CRP, IL6, and TNFa. Supports NA policies, high-missingness warnings, and optional extreme scanning (warn/cap/NA/error). Units assumed: CRP mg/L; IL6, TNFa pg/mL (no internal conversions).

## Load packages and demo data
Synthetic data with one missing value to show NA handling.

```{r}
library(HealthMarkers)
library(tibble)

df <- tibble(
  CRP  = c(1.2, 3.5, NA),  # mg/L
  IL6  = c(2.0, 4.1, 1.5), # pg/mL
  TNFa = c(1.0, 1.8, 0.9)  # pg/mL
)
```

## Column map (required)
Map each marker to your column names; all three are mandatory.

```{r}
col_map <- list(CRP = "CRP", IL6 = "IL6", TNFa = "TNFa")
```

## Core calculation
Default behavior (`na_action = "omit"`/`ignore`/`warn`): NAs are ignored in the weighted sum (treated as 0 contribution), preserving row count.

```{r}
ia_default <- iAge(
  data = df,
  col_map = col_map,
  na_action = "omit",
  check_extreme = FALSE,
  verbose = FALSE
)

ia_default
```

## Keep NA vs drop
`na_action = "keep"` returns NA if any required marker is missing in a row. `na_action = "error"` stops on missing required inputs.

```{r}
ia_keep <- iAge(
  data = df,
  col_map = col_map,
  na_action = "keep",
  check_extreme = FALSE
)

ia_keep
```

## Extreme screening
Scan for out-of-range inputs and cap into allowed bounds.

```{r}
df_ext <- df
df_ext$CRP[2] <- 500  # extreme on purpose

ia_cap <- iAge(
  data = df_ext,
  col_map = col_map,
  check_extreme = TRUE,
  extreme_action = "cap",
  verbose = TRUE
)

ia_cap
```

`extreme_action = "warn"` would only warn; `error` aborts; `NA` blanks flagged values before computing.

## Weights
- Default weights: CRP 0.33, IL6 0.33, TNFa 0.34 (sum to 1). Named weights are reordered to CRP/IL6/TNFa.
- For cohorts where one marker is more informative, adjust weights (ensure they sum to 1):

```{r}
iAge(df, col_map = col_map, weights = c(CRP = 0.5, IL6 = 0.25, TNFa = 0.25))
```

## Expectations
- All three markers must be present and numeric; coercion to numeric will error if NAs are introduced.
- `na_action` behaviors:
  - `omit`/`ignore`/`warn`: missing markers contribute 0 to the sum; rows preserved.
  - `keep`: any missing required marker yields `NA` for that row.
  - `error`: abort if any required marker is missing/NA.
- Extreme scan uses broad defaults (CRP 0–300, IL6 0–1000, TNFa 0–2000); adjust via `extreme_rules` for your lab ranges.
- Output: one-column tibble `iAge`; rows match input unless you subset beforehand.

## Tips
- Keep units consistent (CRP mg/L; IL6/TNFa pg/mL). Convert before calling if needed.
- Prefer `na_action = "keep"` when you want to see which rows are incomplete; use `omit` when you want continuity of the score despite single missing markers.
- Tighten `extreme_rules` for cleaner QA; use `verbose = TRUE` during setup to see mapping and warnings.
- Impute upstream (e.g., `impute_missing()`) if you prefer not to down-weight missing markers to zero.
