---
title: "Kynurenine/Tryptophan ratio"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{kyn_trp_ratio}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(purl = FALSE)
```

## Scope
Compute the kynurenine/tryptophan ratio (KTR), a marker of IDO activity and immune activation. Supports NA policies, optional extreme screening (warn/cap/NA/error), and verbose progress. Inputs must already be Kyn (nmol/L) and Trp (µmol/L); no unit conversion.

## Load packages and demo data

```{r}
library(HealthMarkers)
library(tibble)

df <- tibble(
  Kyn_nM = c(1800, 2500, 3200),
  Trp_uM = c(50, 42, 12)
)
```

## Column map (required)

```{r}
col_map <- list(kynurenine = "Kyn_nM", tryptophan = "Trp_uM")
```

## Core calculation
Default `na_action = "keep"`: ratios are `NA` where inputs are missing; no extreme scan.

```{r}
ktr <- kyn_trp_ratio(
  data = df,
  col_map = col_map,
  na_action = "keep",
  check_extreme = FALSE,
  verbose = FALSE
)

ktr
```

## Extreme screening
Scan and cap into default ranges.

```{r}
df_ext <- df
df_ext$Kyn_nM[2] <- 50000  # extreme

ktr_cap <- kyn_trp_ratio(
  data = df_ext,
  col_map = col_map,
  na_action = "warn",
  check_extreme = TRUE,
  extreme_action = "cap",
  verbose = TRUE
)

ktr_cap
```

`extreme_action = "warn"` only warns; `error` aborts; `NA` blanks flagged values.

## Missing data handling
`keep`/`ignore`/`warn` retain rows with NA ratios; `omit` drops rows with missing inputs; `error` stops on missingness.

```{r}
df_na <- df
df_na$Trp_uM[3] <- NA

keep_out <- kyn_trp_ratio(df_na, col_map, na_action = "keep")
omit_out <- kyn_trp_ratio(df_na, col_map, na_action = "omit")

list(keep_rows = nrow(keep_out), omit_rows = nrow(omit_out))
```

## Expectations
- Both markers must be mapped and present; non-numeric inputs are coerced with warnings if NAs are introduced.
- Trp must be positive; nonpositive values are set to `NA` with a warning.
- Extreme defaults: Kyn 100–20000 nmol/L, Trp 10–150 µmol/L, ratio 0–200; adjust with `extreme_rules`.
- Output: one-column tibble `kyn_trp_ratio`; row count follows `na_action` if you omit rows.

## Tips
- Re-check units when ratios are very high (>100); Kyn should be nmol/L and Trp µmol/L.
- Use `extreme_action = "cap"` sparingly and adjust ranges to your cohort.
- For modeling, consider scaling/transforming KTR separately after computation.
