---
title: "marker_summary"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{marker_summary}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(comment = "#", collapse = TRUE)
knitr::opts_knit$set(purl = FALSE)
```

## Overview

`marker_summary()` provides quick descriptive statistics (mean, SD, IQR) for numeric columns returned by marker functions. It validates input type, optionally emits progress messages, and returns a tidy tibble.

## Required Inputs

- `x`: data.frame/tibble, typically output from marker functions. Must contain at least one numeric column to produce summaries.

## Key Arguments

- `verbose = FALSE`: when `TRUE`, prints start/completion messages and counts all-NA variables.

## Output Description

Returns a tibble with columns `variable`, `mean`, `sd`, and `iqr` for each numeric column in `x`. If no numeric columns are present, returns an empty tibble (0 rows).

## Minimal Reproducible Example

```{r}
library(HealthMarkers)
library(tibble)

df <- tibble(
  marker1 = c(1.1, 2.3, NA, 4.0),
  marker2 = c(10, 12, 11, 9),
  group   = c("a", "a", "b", "b")
)

marker_summary(df)
```

## Extended Example

Enable verbose messages:

```{r}
marker_summary(df, verbose = TRUE)
```

## Notes / Workflow Integration

- Only numeric columns are summarized; non-numeric columns are ignored.
- Missing values are removed for each statistic via `na.rm = TRUE`.
- Use after marker pipelines to get a quick quality check or to feed into reporting tables.