---
title: "Multi-Biobank Compatibility: Automatic Column Recognition"
output: rmarkdown::html_document
vignette: >
  %\VignetteIndexEntry{Multi-Biobank Compatibility}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(purl = FALSE)
```

## Overview

HealthMarkers is designed to work with data from a wide range of international
cohort studies and biobanks **without any manual column renaming**. The internal
synonym dictionary covers naming conventions from 15+ major studies. This article
explains how the matching works, which biobanks are supported, and how to diagnose
and fix cases where a column is not recognised.

---

## How column inference works

Every HealthMarkers function looks up your column names through a five-layer
matching pipeline before any computation:

| Layer | Method | Example |
|---|---|---|
| 1 | Exact match in synonym dictionary | `LBXGLU` → `fasting_glucose` |
| 2 | Case-insensitive exact match | `lbxglu` → `fasting_glucose` |
| 3 | Data column name contains a synonym (≥ 4 chars) | `plasma_glukoosi` → `fasting_glucose` |
| 4 | Synonym contains the data column name (≥ 4 chars) | `glukos` → `fasting_glucose` |
| 5 | Fuzzy (Jaro–Winkler) match, only when `fuzzy = TRUE` | `glucos` → `fasting_glucose` |

The first four layers are active by default. Layer 5 is opt-in and is most useful
for catching typos in column names.

---

## Supported biobanks and naming systems

### UK Biobank (UKB)

UK Biobank fields follow the pattern `analyte_0_0` (assessment visit 0, instance 0)
and `analyte_1_0` (reassessment). HealthMarkers recognises these for all major
analytes:

| Internal key | UKB field | UKB field code |
|---|---|---|
| `fasting_glucose` | `glucose_0_0` | 30740 |
| `total_cholesterol` | `cholesterol_0_0` | 30690 |
| `HDL_c` | `hdl_cholesterol_0_0` | 30760 |
| `LDL_c` | `ldl_direct_0_0` | 30780 |
| `TG` | `triglycerides_0_0` | 30870 |
| `ALT` | `alanine_aminotransferase_0_0` | 30620 |
| `AST` | `aspartate_aminotransferase_0_0` | 30650 |
| `creatinine` | `creatinine_0_0` | 30700 |
| `albumin` | `albumin_0_0` | 30600 |
| `HbA1c` | `glycated_haemoglobin_hba1c_0_0` | 30750 |
| `vitaminD` | `vitamin_d_0_0` | 30890 |
| `Hgb` | `haemoglobin_concentration_0_0` | 30020 |
| `WBC` | `white_blood_cell_leucocyte_count_0_0` | 30000 |
| `platelets` | `platelet_count_0_0` | 30080 |
| `SBP` | `systolic_blood_pressure_0_0` | 4080 |
| `DBP` | `diastolic_blood_pressure_0_0` | 4079 |
| `height` | `standing_height_0_0` | 50 |
| `weight` | `body_weight_0_0` | 21002 |
| `BMI` | `body_mass_index_bmi_0_0` | 21001 |
| `waist` | `waist_circumference_0_0` | 48 |
| `testosterone` | `testosterone_0_0` | 30850 |
| `TSH` | `thyroid_stimulating_hormone_tsh_0_0` | 30830 |
| `urea_serum` | `urea_0_0` | 30670 |
| `uric_acid` | `urate_0_0` | 30880 |
| `sodium` | `sodium_0_0` | 30530 |
| `potassium` | `potassium_0_0` | 30520 |
| `calcium` | `calcium_0_0` | 30680 |
| `phosphate` | `phosphate_0_0` | 30810 |

### NHANES (National Health and Nutrition Examination Survey, USA)

NHANES uses uppercase prefix codes. HealthMarkers recognises both the `LBX`
(examination) and `LBD` (derived) prefixes:

| Internal key | NHANES variable | Questionnaire |
|---|---|---|
| `fasting_glucose` | `LBXGLU` | Biochemistry Profile |
| `total_cholesterol` | `LBXSCH` | Biochemistry Profile |
| `HDL_c` | `LBDHDD` | HDL-Cholesterol |
| `LDL_c` | `LBDLDL` | Cholesterol-LDL |
| `TG` | `LBXSTR` | Biochemistry Profile |
| `ALT` | `LBXSATSI` | Biochemistry Profile |
| `AST` | `LBXSASSI` | Biochemistry Profile |
| `creatinine` | `LBXSCR` | Biochemistry Profile |
| `albumin` | `LBXSAL` | Biochemistry Profile |
| `HbA1c` | `LBXGH` | Glycohemoglobin |
| `vitaminD` | `LBXVD2` | Vitamin D |
| `Hgb` | `LBXHGB` | CBC |
| `WBC` | `LBXWBCSI` | CBC |
| `platelets` | `LBXPLTSI` | CBC |
| `ferritin` | `LBXFER` | Ferritin |
| `CRP` | `LBXHSCRP` | hs-CRP |
| `SBP` | `BPXSY1` | Blood Pressure |
| `DBP` | `BPXDI1` | Blood Pressure |
| `u_albumin` | `URXUMA` | Albumin-Creatinine Urine |
| `u_creatinine` | `URXUCR` | Albumin-Creatinine Urine |
| `Homocysteine` | `LBXHCY` | Homocysteine |

---

## Diagnosing column matching with `hm_col_report()`

Before running any computation on a new dataset, call `hm_col_report()` to
see exactly which columns are matched and how:

```{r, eval=FALSE}
library(HealthMarkers)

# Load your biobank extract (example: HUNT data with Norwegian column names)
hunt_data <- read.csv("hunt_extract.csv")

# Run the column report
hm_col_report(hunt_data)
```

The output looks like:

```
── HealthMarkers column report ────────────────────────────────────────────
 Data: 56301 rows × 187 columns   |   Keys in dictionary: 258

 key                  data_column              how matched
 -------------------- ------------------------ ------------------
 age                  alder                    col contains synonym ✔
 sex                  kjonn                    exact ✔
 height               hoyde                    exact ✔
 fasting_glucose      fastende_blodsukker      col contains synonym ✔
 TG                   triglyserider            col contains synonym ✔
 creatinine           kreatinin                exact ✔
 SBP                  systolisk_blodtrykk      col contains synonym ✔
 eGFR                 ─                        NOT FOUND ✘

 ✔ 142 keys matched   ✘ 116 keys not found

── col_map template for missing keys ──────────────────────────────────────
 col_map <- list(
   eGFR = "from_your_data",  # fill in your column name
 )
```

For the `NOT FOUND` keys, either:
1. Add the column to your data (or compute it from raw inputs),
2. Provide an explicit `col_map` entry, or
3. Ignore it — functions will produce `NA` for any marker that requires that key.

```{r, eval=FALSE}
# Capture the auto-detected map and add manual overrides
cm <- hm_col_report(hunt_data, verbose = FALSE)
cm$eGFR <- "ckd_epi_gfr_ml_min"   # fill in the actual column name

results <- all_health_markers(hunt_data, col_map = cm,
                              which = c("glycemic", "lipid", "renal",
                                        "inflammatory", "liver"))
```

---

## Auto-derived columns

Even when raw computed inputs are not in your data, HealthMarkers can derive
them automatically from more basic measurements before marker computation begins.
The following secondary variables are auto-derived:

| Derived key | Derived from | Formula/method |
|---|---|---|
| `eGFR` | `creatinine`, `age`, `sex` | CKD-EPI 2021 |
| `UACR` | `u_albumin`, `u_creatinine` | ratio (mg/g) |
| `LDL_c` | `TC`, `HDL_c`, `TG` | Friedewald equation |
| `WHR` | `waist`, `hip` | ratio |
| `BMI` | `height`, `weight` | kg/m² |
| `MAP` | `SBP`, `DBP` | (SBP + 2×DBP) / 3 |
| `PP` | `SBP`, `DBP` | SBP − DBP |
| `non_HDL_c` | `TC`, `HDL_c` | TC − HDL_c |
| `TC_HDL_ratio` | `TC`, `HDL_c` | ratio |
| `TG_HDL_ratio` | `TG`, `HDL_c` | log₁₀(TG/HDL_c) |
| `creatinine_ratio` | `creatinine`, `u_creatinine` | ratio |

This means, for example, that `kidney_failure_risk()` will work even if `eGFR`
is absent from your data, as long as `creatinine`, `age`, and `sex` are present.

---

## Example: running on UK Biobank data

```{r, eval=FALSE}
library(HealthMarkers)

# Load UKB extract (columns follow _0_0 naming)
ukb <- readRDS("ukb_pheno.rds")

# Check column matching — most UKB columns are auto-detected
hm_col_report(ukb)

# Run a broad panel — no col_map needed for UKB standard names
results <- all_health_markers(
  data    = ukb,
  which   = c("glycemic", "lipid", "liver", "renal",
              "inflammatory", "obesity_metrics", "vitamin"),
  verbose = TRUE
)
```

## Example: running on NHANES data

```{r, eval=FALSE}
# NHANES lab data (LBXGLU, LBXSCH, LBXWBCSI, etc.)
nhanes_lab <- read.csv("nhanes_lab.csv")
nhanes_bp  <- read.csv("nhanes_bp.csv")
nhanes_dem <- read.csv("nhanes_demo.csv")

nhanes <- nhanes_lab |>
  dplyr::left_join(nhanes_bp,  by = "SEQN") |>
  dplyr::left_join(nhanes_dem, by = "SEQN")

results <- all_health_markers(nhanes, which = c("glycemic", "lipid",
                                                 "renal", "inflammatory"),
                               verbose = FALSE)
```

## Example: running on OMOP / All of Us data (LOINC codes)

```{r, eval=FALSE}
# Columns named LOINC_2345_7, LOINC_2160_0, LOINC_718_7, etc.
omop_labs <- dplyr::collect(tbl(con, "measurement_wide"))

# All LOINC_XXXX_X columns are recognised automatically
hm_col_report(omop_labs)

results <- all_health_markers(omop_labs,
                              which = c("glycemic", "lipid", "renal",
                                        "inflammatory", "vitamin"))
```

---

## Adding a custom biobank

If your biobank uses names not already in the dictionary, you have two options:

**Option A — explicit `col_map` (immediate, no code changes):**

```{r, eval=FALSE}
my_col_map <- list(
  fasting_glucose  = "p_glu_0",
  total_cholesterol = "tot_kol",
  creatinine       = "s_krea",
  age              = "alder_bij_opname",
  sex              = "geslacht_f"
)

results <- all_health_markers(data, col_map = my_col_map,
                              which = c("glycemic", "lipid", "renal"))
```

**Option B — open a GitHub issue / pull request** to add permanent recognition
of the naming system to the dictionary:

> <https://github.com/sufyansuleman/HealthMarkers/issues>

Please include: biobank name, the analyte/variable, and the column name(s) used.

---

## Summary

| Biobank / system | Language | Key feature |
|---|---|---|
| UK Biobank | English | `_0_0` field suffix notation |
| NHANES | English | `LBX`/`LBD`/`BPX`/`URX` prefixes |
| Danish registers | Danish + NPU codes | `NPU01994` etc.; `kreatinin`, `kolesterol` |
| HUNT / Tromsø | Norwegian | `triglyserider`, `karbamid`, double-k `blodtrykk` |
| FinnGen / THL | Finnish | `-iini` endings; `glukoosi`, `kreatiniini` |
| Estonian Biobank | Estonian | `kolesterool` (double-o), `naatrium` (≠ natrium) |
| LifeLines / Rotterdam | Dutch | `ureum` (= urea), `urinezuur` (uric acid) |
| Generation Scotland | English | `SBP_mean`, `genetic_sex` |
| All of Us / OMOP | LOINC | `LOINC_XXXX_X` format |
| NAKO / KORA | German | `Cholesterin`, `Triglyzeride`, `Harnsäure` |

