---
title: "Neurofilament light chain (NfL) marker"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{nfl_marker}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(purl = FALSE)
```

## Scope
Incorporate a neurofilament light chain (NfL) concentration measurement into the analysis pipeline. NfL is a biomarker of neuroaxonal injury; elevated plasma or CSF levels indicate neurodegeneration and typically increase with age and neurological disease. This function validates and passes through the input values (assumed plasma pg/mL) with optional NA handling and extreme-value scanning, returning a tidy tibble ready for downstream analysis.

## Load packages and demo data

```{r}
library(HealthMarkers)
library(tibble)

df <- tibble::tibble(
  NfL = c(8.5, 14.2, 22.1, NA, 35.0)
)
```

## Column map
The only required key is `nfl`.

```{r}
cm <- list(nfl = "NfL")
```

## Core calculation

```{r}
out <- nfl_marker(
  data = df,
  col_map = cm,
  na_action = "keep",
  check_extreme = FALSE,
  verbose = FALSE
)

out
```

## Missing data handling
`na_action = "keep"` preserves rows with `NA` outputs; `omit` drops rows with missing required inputs; `error` aborts if any required input is missing; `warn` keeps rows but emits a warning.

```{r}
df_na <- tibble::tibble(NfL = c(8.5, NA, 22.1))

keep_out <- nfl_marker(df_na, cm, na_action = "keep")
omit_out <- nfl_marker(df_na, cm, na_action = "omit")

list(keep_rows = nrow(keep_out), omit_rows = nrow(omit_out))
```

## Extreme-value handling
`check_extreme = TRUE` scans values against a broad default range of [0, 1e6] pg/mL.  
`extreme_action` controls the response: `"warn"` emits a warning without altering values, `"cap"` clips to the range boundary, `"NA"` blanks flagged values, `"error"` aborts.

```{r, warning=TRUE}
df_ext <- tibble::tibble(NfL = c(8.5, -5, 22.1))  # negative value is out of range

nfl_marker(
  data = df_ext,
  col_map = cm,
  check_extreme = TRUE,
  extreme_action = "warn"
)
```

Custom range override:

```{r}
nfl_marker(
  data = tibble::tibble(NfL = c(8.5, 5000, 22.1)),
  col_map = cm,
  check_extreme = TRUE,
  extreme_action = "cap",
  extreme_rules = list(nfl = c(0, 1000))
)
```

## Verbose diagnostics
Set `verbose = TRUE` (and `healthmarkers.verbose = "inform"`) to surface three structured messages on each call: preparing inputs, the column map, and a results summary.

```{r, message=TRUE}
old_opt <- options(healthmarkers.verbose = "inform")
nfl_marker(tibble::tibble(NfL = c(8.5, 14.2, 22.1)), cm, verbose = TRUE)
options(old_opt)
```

## Tips
- Units are preserved as-is; ensure all values are in the same unit (e.g., pg/mL) before calling.
- Age-adjusted reference ranges are not applied by this function; use downstream tools or stratified z-scores for age normalisation.
- For batch processing with imputed datasets, combine with `impute_missing()` upstream to handle systematic missingness before running `nfl_marker()`.
