---
title: "Obesity Indices"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Obesity Indices}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  message = FALSE,
  warning = FALSE
)
knitr::opts_knit$set(purl = FALSE)
```

## Scope

Compute anthropometric adiposity/shape indices (BMI + category, WHR/WHtR, AVI, BAI, ABSI, BRI, Conicity Index, waist/BMI and weight/height ratios, optional WHRadjBMI and RFM) with unit conversion, NA/extreme handling, and safe-division guards.

## When to use

- You have weight/height/waist/hip (and optionally sex) and want a comprehensive set of body composition proxies in one call.
- You need built-in unit conversion (kg/lb, cm/m) and NA policy controls.
- You want optional BMI-adjusted WHR and Relative Fat Mass when sex coding is available.

## Inputs

- `data`: data frame/tibble with numeric columns for weight, height, waist, hip; `sex` needed only if `include_RFM = TRUE` (coded 0=male, 1=female).
- Column arguments are unquoted: `weight`, `height`, `waist`, `hip`, optional `sex`.
- Units: weight kg (or lb if `weight_unit = "lb"`), height m (or cm if `height_unit = "cm"`), waist/hip cm.
- `weight_unit` / `height_unit`: convert to kg/m internally.
- `adjust_WHR`: add BMI-adjusted WHR residuals.
- `include_RFM`: compute Relative Fat Mass (requires `sex`).
- `na_action`: `keep` (default) propagates NA; `omit` drops rows missing required inputs; `error` aborts when required inputs contain NA.
- `na_warn_prop`: threshold for high-missingness debug messages (default 0.2).
- `verbose`: optional progress and summary logging.

## Quick start

```{r}
library(HealthMarkers)
library(tibble)

df <- tibble::tibble(
  wt    = c(70, 80),   # kg
  ht    = c(175, 165), # cm
  waist = c(80, 90),   # cm
  hip   = c(100, 95),  # cm
  sex   = c(0, 1)
)

obesity_indices(
  data = df,
  weight = wt,
  height = ht,
  waist = waist,
  hip = hip,
  sex = sex,
  weight_unit = "kg",
  height_unit = "cm",
  adjust_WHR = TRUE,
  include_RFM = TRUE,
  na_action = "keep",
  verbose = FALSE
)
```

## Verbose diagnostics

```{r, message=TRUE}
old_opt <- options(healthmarkers.verbose = "inform")
obesity_indices(
  data = df,
  weight = wt,
  height = ht,
  waist = waist,
  hip = hip,
  sex = sex,
  weight_unit = "kg",
  height_unit = "cm",
  adjust_WHR = FALSE,
  include_RFM = TRUE,
  na_action = "omit",
  verbose = TRUE
)
options(old_opt)
```

## Missing-data policy

```{r}
try(
  obesity_indices(
    data = df,
    weight = wt,
    height = ht,
    waist = waist,
    hip = hip,
    na_action = "error"
  )
)
```

## Outputs and expectations

- Adds: `weight_kg`, `height_m`, `BMI`, `BMI_cat`, `WHR`, `waist_to_height_ratio`, `AVI`, `BAI`, `ABSI`, `BRI`, `CI`, `waist_to_BMI_ratio`, `weight_to_height_ratio`, optional `WHRadjBMI`, optional `RFM`.
- Zero denominators yield `NA` with warnings; WHRadjBMI requires variance in WHR and BMI; RFM requires sex coded 0/1.

## Tips

- Set weight/height units correctly; waist/hip are assumed cm.
- Enable `include_RFM = TRUE` only with valid sex coding; leave FALSE otherwise.
- WHRadjBMI needs variation in WHR and BMI; otherwise it returns `NA` with a warning.
- Use `na_action = "omit"` for row-complete outputs; keep during QA to see missingness.
