---
title: "OGTT Insulin Sensitivity"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{OGTT Insulin Sensitivity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  message = FALSE,
  warning = FALSE
)
knitr::opts_knit$set(purl = FALSE)
```

## Scope
Compute OGTT-based insulin sensitivity indices (Matsuda, Cederholm, Gutt, Stumvoll variants, BIGTT, HIRI_inv, Belfiore, etc.) from glucose/insulin at 0/30/120 minutes plus weight, BMI, age, and sex. Includes validation, missingness policies, and optional normalization of the returned indices.

## When to use this
- You have OGTT glucose (mmol/L) and insulin (pmol/L) at 0, 30, 120 minutes, plus weight, BMI, age, and sex, and need multiple published insulin sensitivity indices.
- You want consistent NA handling and optional normalization for modeling pipelines.

## What you need (inputs & options)

| Argument       | Purpose / Options                                                        | Notes                                                     |
|----------------|--------------------------------------------------------------------------|-----------------------------------------------------------|
| data           | Data frame/tibble with OGTT labs and demographics                        | Columns mapped via `col_map`                              |
| col_map        | Named list mapping required keys                                         | G0/G30/G120 (glucose mmol/L); I0/I30/I120 (insulin pmol/L); weight (kg); bmi (kg/m^2); age (years); sex (1=male, 2=female; 0/1 coerced) |
| normalize      | Output normalization method                                              | "none", "z", "inverse", "range", "robust" (default "none") |
| na_action      | Missing-data policy for required inputs                                  | "keep" (default), "omit", "error"                      |
| na_warn_prop   | Proportion to trigger high-missingness diagnostics                       | Default 0.2 (emitted in debug/verbose)                    |
| verbose        | Emit progress and completion summaries                                   | Default FALSE                                             |

**Units:** Inputs must already be in mmol/L (glucose) and pmol/L (insulin); weight kg; BMI kg/m^2; age years; sex coded 1/2 (0/1 coerced). The function internally converts glucose to mg/dL (x18) and insulin to muU/mL (/6) where specific formulas require it.

## Handling and expectations
- Validation: required columns must exist; non-numeric inputs are coerced with warnings; non-finite values become NA.
- Missingness: `keep` propagates NA; `omit` drops rows with any required NA; `error` aborts if required inputs contain NA (before computation).
- High-missing diagnostics: `na_warn_prop` controls debug/verbose messages on missingness; enable `verbose = TRUE` to see them.
- Unit handling: inputs are expected in mmol/L (glucose) and pmol/L (insulin); internally converted to mg/dL (*18) and muU/mL (/6) where formulas need them.
- Normalization (optional): applied via `normalize_vec()`; use "none" for raw interpretable values.
- Sex coding: expects 1 = male, 2 = female; 0 is treated as female after coercion; other strings become NA.

## Outputs
- Returns a tibble with: Isi_120, Cederholm_index, Gutt_index, Avignon_Si0, Avignon_Si120, Avignon_Sim, Modified_stumvoll, Stumvoll_Demographics, Matsuda_AUC, Matsuda_ISI, BigttSi, Ifc_inv, HIRI_inv, Belfiore_isi_gly.
- NA indicates missing/invalid inputs, zero/negative arguments to logs, or values blanked by normalization after NA inputs.

## Defaults and validation details
- Output screening happens before normalization; outputs are then normalized if `normalize != "none"`.
- High-missingness diagnostics are only printed when verbose/debug logging is on.
- Zero/negative inputs to log-based terms become NA by design (safe log guard).

## Worked example 1: Basic usage (raw outputs)
```{r}
library(HealthMarkers)
library(tibble)

df <- tibble::tibble(
  G0 = 5.5, I0 = 60,
  G30 = 7.8, I30 = 90,
  G120 = 6.2, I120 = 50,
  weight = 70, bmi = 24, age = 30, sex = 1
)

ogtt_is(
  data = df,
  col_map = list(
    G0 = "G0", I0 = "I0",
    G30 = "G30", I30 = "I30",
    G120 = "G120", I120 = "I120",
    weight = "weight", bmi = "bmi",
    age = "age", sex = "sex"
  ),
  normalize = "none",
  na_action = "keep"
)
```
*Interpretation:* Returns all indices with NA only where inputs are missing or non-finite.

## Worked example 2: Normalization
```{r}
ogtt_is(
  data = df,
  col_map = list(
    G0 = "G0", I0 = "I0",
    G30 = "G30", I30 = "I30",
    G120 = "G120", I120 = "I120",
    weight = "weight", bmi = "bmi",
    age = "age", sex = "sex"
  ),
  normalize = "z",
  na_action = "keep",
  verbose = TRUE
)
```
*Interpretation:* Outputs are z-scaled; verbose logs summarize column-map resolution and result counts.

## Troubleshooting & common pitfalls
- Missing YAML: ensure this header stays at the top; Markdown before it will break knitting.
- Missing columns: every required `col_map` key must exist in `data` and be mapped.
- Units: inputs must be mmol/L (glucose) and pmol/L (insulin); only internal conversions are applied for formulas.
- Sex coding: expects 1/2; 0/1 are coerced; other strings become NA and propagate.
- All NA outputs: usually missing required inputs, zero/negative values passed to logs, or NA produced by normalization after NA inputs.

## Verbose diagnostics

```{r, message=TRUE}
old_opt <- options(healthmarkers.verbose = "inform")
ogtt_is(
  data = df,
  col_map = list(
    G0 = "G0", I0 = "I0",
    G30 = "G30", I30 = "I30",
    G120 = "G120", I120 = "I120",
    weight = "weight", bmi = "bmi",
    age = "age", sex = "sex"
  ),
  normalize = "none",
  verbose = TRUE
)
options(old_opt)
```

## Tips for best results
- Keep `normalize = "none"` for clinical interpretation; use "z" or "robust" only when scaling is needed for modeling.
- Use `na_action = "omit"` for modeling-ready datasets; `keep` for exploratory review.
- Inspect distributions of glucose and insulin values to catch unit errors before running.

## Validation notes
- Glucose values are multiplied by 18 and insulin divided by 6 inside formulas that require mg/dL or muU/mL.
- Logs are safe: nonpositive arguments become NA.
- Normalization occurs after computing raw indices; disable to mirror published formulas.

## See also
- Function documentation: `?ogtt_is`, `?normalize_vec`
- Related vignettes: fasting_is, health_markers, glycemic_markers
