---
title: "Oxidative Markers"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Oxidative Markers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  message = FALSE,
  warning = FALSE
)
```

```{r, include=FALSE}
knitr::opts_knit$set(purl = FALSE)
```
## Scope
Compute the glutathione redox ratio (`GSH_GSSG_Ratio`) from reduced (GSH) and oxidized (GSSG) glutathione, with robust input validation, missing-data handling, and clear output. No unit conversion is performed.


## When to use this
- You need a reliable marker of oxidative stress for clinical or research data.
- Your dataset includes both reduced (GSH) and oxidized (GSSG) glutathione measurements.
- You want to handle missing or non-numeric values gracefully, with options to omit, propagate, or error on missing data.



## What you need (detailed inputs)

| Argument    | Purpose/Options                                                                 | Details/Notes                                  |
|-------------|--------------------------------------------------------------------------------|------------------------------------------------|
| data        | Data frame/tibble with GSH and GSSG columns                                    | Columns can be any name, mapped via col_map     |
| col_map     | Named list: GSH, GSSG → your column names                                      | Defaults to identity mapping                    |
| units       | No conversion performed; must match (e.g., µmol/L)                             | Mismatched units distort results                |
| na_action   | "keep" (propagate NA), "omit" (drop rows), "error" (abort if missing)         | Controls missing data handling                  |
| verbose     | TRUE/FALSE: print progress and summary                                         | Useful for diagnostics                          |

### Example input table
| GSH | GSSG |
|-----|------|
| 5   | 1    |
| 3   | 0.5  |
| NA  | 2    |
| 2   | 0    |


## Handling and expectations
- **Input validation:** Non-numeric columns are coerced; warnings are issued if NAs are introduced. Non-finite values become NA.
- **Missing data:**
  - `na_action = "keep"`: Output is NA for rows with missing GSH or GSSG.
  - `na_action = "omit"`: Rows with missing GSH or GSSG are dropped.
  - `na_action = "error"`: Function aborts if any required value is missing.
- **Division by zero:** If GSSG is zero or non-finite, the ratio is set to NA with a warning.
- **Verbose output:** Shows counts of omitted/kept rows and NA ratios for troubleshooting.
- **Troubleshooting:** If you see many NAs, check for zero or non-numeric GSSG values, or mismatched column names in `col_map`.


## Output
- Returns a tibble with one column: `GSH_GSSG_Ratio`.
- NA values indicate missing or invalid input, or division by zero.

## Worked example 1: Basic usage
*Interpretation:* The third row yields NA because GSH is missing. Ratios are computed as GSH/GSSG for each row.

```{r}
library(HealthMarkers)
library(tibble)

df <- tibble(GSH = c(5, 3, NA), GSSG = c(1, 0.5, 2))
oxidative_markers(
  data = df,
  col_map = list(GSH = "GSH", GSSG = "GSSG"),
  na_action = "keep"
)
```
*Interpretation:* The third row yields NA because GSH is missing. Ratios are computed as GSH/GSSG for each row.

## Worked example 2: Omit missing data and verbose output
```{r}
oxidative_markers(
  data = df,
  col_map = list(GSH = "GSH", GSSG = "GSSG"),
  na_action = "omit",
  verbose = TRUE
)
```
*Interpretation:* Only complete rows are returned. Verbose output shows counts of omitted rows.



## Troubleshooting & common pitfalls
- **Unit mismatch:** If GSH and GSSG are in different units, the ratio is meaningless. Always check units.
- **Zero or near-zero GSSG:** Division by zero yields NA. Inspect your data for implausible values.
- **Non-numeric input:** Non-numeric columns are coerced; check for unexpected NAs after coercion.
- **High missingness:** If many NAs are present, check your input data and consider using `na_action = "omit"`.
- **Column mapping errors:** If you get a missing column error, check your `col_map` and input names.

## Tips for best results
- Use `na_action = "omit"` for modeling or summary statistics to avoid NA propagation.
- Set `verbose = TRUE` to get a summary of omitted/kept rows and NA counts.
- Always verify assay units before computing ratios.
- Spot-check a few rows manually to confirm expected output.
- If you see many NAs, check for zero or non-numeric GSSG values.
- For large datasets, run a summary (e.g., `summary(df$GSH)`, `summary(df$GSSG)`) before analysis.
- Use `dplyr::filter()` to pre-screen for implausible values if needed.

## Validation notes
- Spot-check: For a row, `GSH_GSSG_Ratio = GSH / GSSG`.
- NA output means missing or invalid input, or division by zero.
- Compare summary statistics (mean, SD) to expected ranges for your population.
- If output is all NA, check for column mapping errors or all-zero GSSG.

## See also
- Other marker functions in HealthMarkers for related stress and metabolic indices: `inflammatory_markers`, `metabolic_markers`, `nutrient_markers`.
- Package vignettes for metabolic, inflammatory, and nutrient markers.
- HealthMarkers documentation for advanced usage and troubleshooting.
- **Non-numeric input:** Non-numeric columns are coerced; check for unexpected NAs after coercion.

