---
title: "Pulmonary Markers"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Pulmonary Markers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  message = FALSE,
  warning = FALSE
)
knitr::opts_knit$set(purl = FALSE)
```

## Scope
Compute spirometry-based pulmonary markers using GLI/GLIgl/NHANES3 reference equations from `rspiro`: predicted normals, z-scores, percent-predicted, LLN for FEV1, FVC, and the FEV1/FVC ratio. Handles missing-data policies, basic height unit detection, and verbose diagnostics.

## When to use this
- You need standardized pulmonary markers (predicted, z, percent-predicted, LLN) for FEV1, FVC, and FEV1/FVC using established reference equations.
- Your dataset has age/sex/height/ethnicity plus measured FEV1 and FVC (liters).
- You want controlled handling of missingness and clear warnings on implausible inputs (e.g., unmapped sex/ethnicity, zero/negative heights).

## What you need (inputs & options)

| Argument      | Purpose / Options                                                                    | Notes                                                  |
|---------------|--------------------------------------------------------------------------------------|--------------------------------------------------------|
| data          | Data frame/tibble with spirometry fields                                             | Required columns listed below                          |
| equation      | "GLI", "GLIgl" (ignores ethnicity), or "NHANES3"                                   | Default GLI                                            |
| na_action     | keep (propagate NA), omit (drop rows), error (abort if required missing)             | Default keep                                           |
| na_warn_prop  | Proportion to trigger high-missingness warnings (default 0.2)                        | Applies to required inputs                             |
| verbose       | TRUE/FALSE for progress and completion summaries                                     | Default FALSE                                          |

**Required columns** (names must match exactly): age (years), sex ("male"/"female" or 1/2 or 0/1), height (cm or m; auto-detected), ethnicity (text), fev1 (L), fvc (L). No col_map is needed; use these names.

**Units:** FEV1/FVC in liters. Height auto-detected as cm if any value > 3; otherwise treated as meters. No other unit conversions are performed.

## Handling and expectations
- Validation & coercion: required columns are checked; age/height/fev1/fvc coerced to numeric with warnings if NAs are introduced.
- Missingness: keep = propagate NA; omit = drop rows with any required NA; error = abort if any required NA.
- Height handling: any height > 3 is treated as cm and divided by 100; non-positive heights warn and yield NA outputs.
- Sex/Ethnicity mapping: case-insensitive mapping; unmapped values warn and become NA (sex) or Other/Mixed (ethnicity code 5).
- Division by zero/non-finite: ratios set to NA when the denominator is zero or non-finite.
- Dependency: requires `rspiro`; aborts if missing or if the chosen equation is unsupported.

## Outputs
Tibble (rows follow na_action) with:
- fev1_pred, fev1_z, fev1_pctpred, fev1_LLN
- fvc_pred, fvc_z, fvc_pctpred, fvc_LLN
- fev1_fvc_ratio, fev1_fvc_pred, fev1_fvc_z, fev1_fvc_pctpred, fev1_fvc_LLN (LLN currently NA placeholder)
Values are numeric; NA where inputs are missing/invalid or division-by-zero occurred.

## Worked example 1: GLI (default), keep NAs
```{r}
library(HealthMarkers)
library(tibble)

df <- tibble(
  age = c(45, 62, 30),
  sex = c("male", "female", "female"),
  height = c(178, 162, 1.70),  # mixed cm/m accepted (auto-detect)
  ethnicity = c("Caucasian", "African-American", "Other"),
  fev1 = c(2.8, 1.4, 3.0),
  fvc  = c(3.5, 2.1, 3.8)
)

pulmo_markers(
  data = df,
  equation = "GLI",
  na_action = "keep",
  verbose = TRUE
)
```
*Interpretation:* Mixed height units are handled; missingness (none here) would propagate. Outputs include predicted, z, percent-predicted, and ratios.

## Worked example 2: GLIgl, drop incomplete rows
```{r}
df2 <- tibble(
  age = c(45, 62, 30),
  sex = c(1, 2, 2),
  height = c(178, NA, 170),
  ethnicity = c("Caucasian", "African-American", "Other"),
  fev1 = c(2.8, 1.4, 3.0),
  fvc  = c(3.5, 2.1, 3.8)
)

pulmo_markers(
  data = df2,
  equation = "GLIgl",
  na_action = "omit",
  verbose = TRUE
)
```
*Interpretation:* Row 2 is dropped due to missing height. Ethnicity is ignored for GLIgl. Outputs are returned only for complete cases.

## Troubleshooting & common pitfalls
- Missing package: install `rspiro` if absent.
- Unsupported equation: ensure equation is GLI, GLIgl, or NHANES3.
- Unit issues: heights must be cm or m; volumes in liters. Convert inches or other units upstream.
- Unmapped sex/ethnicity: check spelling/casing; unmapped values warn and become NA/Other.
- High missingness: heed `na_warn_prop` warnings; consider `na_action = "omit"` for modeling.
- All NA outputs: likely missing required fields, zero/negative heights, or failed mapping.

## Tips for best results
- Validate units before running; store heights in cm or m and volumes in liters.
- Use `verbose = TRUE` during development/QC to see omissions and NA counts.
- For modeling, prefer `na_action = "omit"` to avoid NA propagation.
- Spot-check a few rows: `fev1_pctpred` should be `100 * fev1 / fev1_pred` when finite; `fev1_fvc_ratio` should match `fev1 / fvc`.
- Pre-screen data with `summary()` to catch extremes before calling the function.

## Validation notes
- Expect ratios ~0.6–0.9 in adults; percent-predicted near 100% in healthy subjects.
- If many z-scores are extreme, re-check age/height/sex/ethnicity coding and units.
- If output is all NA, verify required columns and `rspiro` availability.

## See also
- Related vignettes: inflammatory_markers, metabolic_markers, nutrient_markers.
- Function docs: `pulmo_markers()` and `rspiro` reference equation documentation.
