---
title: "Saliva markers"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{saliva_markers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(purl = FALSE)
```

## Scope

Compute four salivary-based markers from cortisol awakening response (CAR) measurements and salivary amylase / glucose:

- **`log_cortisol_wake`**: log(cort1) — log-transformed waking cortisol.
- **`CAR_AUC`**: trapezoidal area under the cortisol curve over the awakening period (default: 0–60 min using cort1/cort2/cort3).
- **`log_amylase`**: log(amylase) — log-transformed salivary α-amylase.
- **`saliva_glucose`**: raw salivary glucose, passed through unchanged.

## When to use

- You collected serial salivary cortisol (wake, +30 min, +60 min), α-amylase, and/or salivary glucose and want standard derived markers.
- You need NA propagation control and optional verbose diagnostics in a tidy pipeline.

## Inputs

- **Required keys** (all five must be mappable or inferrable): `cort1`, `cort2`, `cort3` (nmol/L), `amylase` (U/mL), `glucose` (mg/dL).
- Default column names expected: `saliva_cort1`, `saliva_cort2`, `saliva_cort3`, `saliva_amylase`, `saliva_glucose` — matched automatically when no `col_map` is provided.
- `na_action`: `keep` (default) propagates NA; `omit` drops rows missing any required input; `error` aborts on any NA.
- `times`: numeric vector of length 3 giving sampling minutes for the trapezoidal AUC. Default `c(0, 30, 60)`.

## Quick start

Data with the default column names — no `col_map` needed:

```{r}
library(HealthMarkers)
library(tibble)

sal <- tibble::tibble(
  saliva_cort1   = c(12.5, 10.0, 15.3),  # nmol/L
  saliva_cort2   = c(18.0, 20.3, 22.1),
  saliva_cort3   = c(14.2, 16.0, 18.0),
  saliva_amylase = c(85, 110, 72),        # U/mL
  saliva_glucose = c(4.2, 5.0, 3.9)      # mg/dL
)

saliva_markers(sal, verbose = FALSE)
```

## Non-standard column names

Only supply `col_map` entries for the keys that differ from default names. Synonym patterns cover common alternatives (`cortisol_wake` → `cort1`, `cortisol_30` → `cort2`, `cortisol_60` → `cort3`).

```{r}
df_ns <- data.frame(
  cort_t0   = 11.2,
  cort_t30  = 17.5,
  cort_t60  = 13.8,
  saa       = 95,    # salivary alpha-amylase
  sal_glu   = 4.6
)

saliva_markers(
  df_ns,
  col_map = list(
    cort1   = "cort_t0",
    cort2   = "cort_t30",
    cort3   = "cort_t60",
    amylase = "saa",
    glucose = "sal_glu"
  ),
  verbose = FALSE
)
```

## Custom sampling times

When cortisol was sampled at different time points, pass the actual minutes via `times`. The AUC is computed with the trapezoidal rule using those intervals.

```{r}
sal_45 <- tibble::tibble(
  saliva_cort1 = 12.0,
  saliva_cort2 = 17.0,
  saliva_cort3 = 15.0,
  saliva_amylase = 90,
  saliva_glucose = 4.5
)

# Cortisol sampled at 0, 20, and 45 minutes instead of 0/30/60
saliva_markers(sal_45, times = c(0, 20, 45), verbose = FALSE)
```

## Missing data handling

```{r}
sal_na <- sal
sal_na$saliva_cort2[2] <- NA

saliva_markers(sal_na, verbose = FALSE)             # na_action = "keep": NA propagates into CAR_AUC
saliva_markers(sal_na, na_action = "omit", verbose = FALSE)  # row 2 dropped
```

## Verbose diagnostics

```{r, message=TRUE}
old_opt <- options(healthmarkers.verbose = "inform")
saliva_markers(
  sal,
  verbose = TRUE
)
options(old_opt)
```

## Outputs

Returns a tibble with columns `log_cortisol_wake`, `CAR_AUC`, `log_amylase`, and `saliva_glucose`. An ID column is prepended when detected in `data`.

- Safe-log is applied: zero or negative cortisol/amylase values yield `NA` (not `-Inf`).
- `CAR_AUC` requires all three cortisol readings to be finite; any NA in cort1/2/3 gives `NA` for that row.

## Tips

- Ensure cortisol values are in nmol/L and amylase in U/mL; no unit conversion is applied.
- When your protocol uses fewer than three cortisol time points, you cannot obtain `CAR_AUC` — supply `NA` for the missing time point and use `na_action = "keep"`.
- Pass `times` explicitly when your sampling schedule deviates from 0/30/60.
- See `?saliva_markers` for full argument details and references.
