---
title: "SARC-F Score"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SARC-F Score}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  message = FALSE,
  warning = FALSE
)
knitr::opts_knit$set(purl = FALSE)
```

## Scope
Compute the 5-item SARC-F sarcopenia screening score (Strength, Assistance in walking, Rise from chair, Climb stairs, Falls). Items are expected on a 0–2 scale; total ranges 0–10. Flags `sarc_f_high_risk` when total ≥ 4. Includes mapping validation, NA policies, optional extreme screening/capping, and non-numeric coercion safeguards.

## When to use this
- You have SARC-F questionnaire responses and need the total score plus a high-risk flag.
- You want controlled handling of missing or miscoded responses (0–2 expected) with optional capping.
- You need outputs suitable for downstream modeling or screening summaries.

## What you need (inputs & options)

| Argument       | Purpose / Options                                               | Notes                                               |
|----------------|-----------------------------------------------------------------|-----------------------------------------------------|
| data           | Data frame/tibble with SARC-F responses                         | Columns mapped via `col_map`                        |
| col_map        | Named list mapping required items                               | strength, walking, chair, stairs, falls             |
| na_action      | Missing-data policy                                             | "keep" (default), "omit", "error", "ignore", "warn" |
| check_extreme  | Enable range checks (0–2 expected)                              | Default FALSE                                       |
| extreme_action | Handling for out-of-range values                                | "warn" (default), "cap", "error", "ignore", "NA"   |
| extreme_rules  | Optional per-item bounds `c(min, max)`                          | Defaults to 0–2 per item                           |
| verbose        | Emit progress messages                                          | Default FALSE                                       |

**Required columns (col_map):** strength, walking, chair, stairs, falls (expected 0/1/2). Mapped columns must exist.

**Units:** None. Inputs should already be coded 0–2; no transformation beyond coercion/capping/NA handling.

## Handling and expectations
- Validation & coercion: mapped columns must exist; non-numeric inputs are coerced to numeric with warnings if NAs are introduced; non-finite values become NA.
- Missingness: `keep`/`ignore`/`warn` propagate NA to the total; `omit` drops rows with any required NA; `error` aborts if missing.
- Out-of-range check: even without `check_extreme`, a warning is issued if values fall outside 0–2. With `check_extreme = TRUE`, `extreme_action` decides warn/cap/error/ignore/NA.
- Extreme rules: defaults cap each item to 0–2; can be overridden per item via `extreme_rules`.
- Scoring: `sarc_f_score` sums the five items; `sarc_f_high_risk` is TRUE when score ≥ 4, FALSE otherwise, NA if score is NA.
- Empty result: if `na_action = "omit"` drops all rows, returns a zero-row tibble with expected columns.

## Outputs
- Returns a tibble with `sarc_f_score` (0–10, or NA) and `sarc_f_high_risk` (TRUE/FALSE/NA).
- NA arises from missing inputs, NA introduced during coercion, NA set by extreme handling, or dropped rows under `omit`.

## Worked example 1: Basic usage (keep NAs)
```{r}
library(HealthMarkers)
library(tibble)

df <- tibble::tibble(
  Strength = c(0, 1, 2, NA),
  Walking  = c(0, 0, 2, 1),
  Chair    = c(0, 1, 2, 1),
  Stairs   = c(0, 1, 2, 1),
  Falls    = c(0, 0, 1, 0)
)

sarc_f_score(
  data = df,
  col_map = list(
    strength = "Strength",
    walking  = "Walking",
    chair    = "Chair",
    stairs   = "Stairs",
    falls    = "Falls"
  ),
  na_action = "keep",
  verbose = TRUE
)
```
*Interpretation:* Rows with complete items return scores; the row with an NA item yields `sarc_f_score = NA` and `sarc_f_high_risk = NA`.

## Worked example 2: Detect and cap miscoded values
```{r}
df2 <- tibble::tibble(
  str = c(0, 3, -1),
  walk = c(1, 2, 1),
  chair = c(0, 2, 1),
  stairs = c(0, 2, 2),
  falls = c(0, 0, 1)
)

sarc_f_score(
  data = df2,
  col_map = list(strength = "str", walking = "walk", chair = "chair", stairs = "stairs", falls = "falls"),
  na_action = "keep",
  check_extreme = TRUE,
  extreme_action = "cap",
  verbose = TRUE
)
```
*Interpretation:* Values outside 0–2 are capped into range before summing; warnings note the corrections. Scores and risk flags are returned for all rows.

## Troubleshooting & common pitfalls
- YAML/header: keep the YAML at the top; any text before it breaks knitting.
- Missing columns: ensure every required `col_map` key is mapped to an existing column.
- Non-numeric inputs: coerced with warnings; resulting NAs propagate unless `omit` is used.
- Out-of-range values: a warning is emitted even without `check_extreme`; set `check_extreme = TRUE` + `extreme_action = "cap"` or `"NA"` to enforce bounds.
- All NA scores: usually due to missing items, coercion to NA, or NA introduced by extreme handling with `extreme_action = "NA"`.

## Verbose diagnostics

Enable verbose output to inspect column mapping, row counts, and result summaries during QC:

```{r, message=TRUE}
old_opt <- options(healthmarkers.verbose = "inform")
sarc_f_score(
  data.frame(Strength = 1, Walking = 1, Chair = 1, Stairs = 0, Falls = 0),
  col_map = list(strength="Strength", walking="Walking", chair="Chair", stairs="Stairs", falls="Falls"),
  verbose = TRUE
)
options(old_opt)
```

## Tips for best results
- Use `check_extreme = TRUE` with `extreme_action = "cap"` to hard-limit inputs to 0–2 when data may be noisy.
- Choose `na_action = "omit"` when you need complete-case scores for modeling; use `keep`/`warn` for exploratory summaries where partial data is expected.
- Treat `sarc_f_high_risk` as a screening flag (score ≥ 4); follow up with clinical assessment as appropriate.

## Validation notes
- Score is a simple sum of the five items; risk flag is `score >= 4` when the score is finite.
- Default bounds are 0–2 per item; override via `extreme_rules` if your instrument coding differs.
- No unit conversion or weighting is applied; inputs should already reflect the standard SARC-F coding.

## See also
- Function docs: `?sarc_f_score`
- Related: frailty_index, health_summary
