---
title: "Adiposity SDS"
output: rmarkdown::html_document
---

```{r, include=FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(purl = FALSE)
```

## Scope
Non-stratified SDS (z-scores) for anthropometric measures using a single reference (mean/sd), with NA policies and optional SDS capping.

## When to use this
- You want simple z-scores against one reference (no sex stratification).
- You need NA handling and optional capping of implausible raw values or extreme SDS magnitudes.
- You plan to standardize BMI/waist/height (or similar) before modeling or QC.

## What you need (rich guide)
- Required: a reference list (`ref`) with `mean` and `sd > 0` per variable, and matching columns in your data (via `col_map$vars` or identity names).
- Units: data units must match the reference (e.g., BMI kg/m^2, waist cm, height cm).
- Policies: choose `na_action` (keep/omit/error) and `sds_cap` + `extreme_action` to cap large SDS magnitudes.
- Optional: `return_summary = TRUE` to get diagnostics, `diagnostics = TRUE` for warnings, `verbose = TRUE` for progress.

## Load packages and data
Use a small slice of the packaged simulated data. Replace `sim_small` with your data.

```{r}
library(HealthMarkers)
library(dplyr)

sim_path <- system.file("extdata", "simulated_hm_data.rds", package = "HealthMarkers")
sim <- readRDS(sim_path)
sim_small <- dplyr::slice_head(sim, n = 30)
```

## Column map
Map each variable to your column names.

```{r}
col_map <- list(
  vars = list(
    BMI = "BMI",
    waist = "waist",
    height = "height"
  )
)
```

## Walkthrough (build ref, compute SDS)

`adiposity_sds()` returns a tibble of **only the `<var>_SDS` columns** — not the original data. Bind back with `dplyr::bind_cols()` if needed.

```{r}
ref <- list(
  BMI = c(mean = mean(sim_small$BMI, na.rm = TRUE), sd = sd(sim_small$BMI, na.rm = TRUE)),
  waist = c(mean = mean(sim_small$waist, na.rm = TRUE), sd = sd(sim_small$waist, na.rm = TRUE)),
  height = c(mean = mean(sim_small$height, na.rm = TRUE), sd = sd(sim_small$height, na.rm = TRUE))
)

asds <- adiposity_sds(
  data = sim_small,
  col_map = col_map,
  ref = ref,
  na_action = "keep",
  extreme_action = "cap",
  sds_cap = 6,
  verbose = FALSE
)

new_cols <- setdiff(names(asds), names(sim_small))
asds %>% slice_head(n = 5) %>% select(all_of(new_cols))
```

Interpretation: each `<var>_SDS` is `(x - mean)/sd`; values near 0 are average, positive above, negative below the reference. `extreme_action = "cap"` trims SDS values beyond `sds_cap`.

## Missing data and SDS capping
Compare NA policies and SDS capping.

```{r}
demo <- sim_small[1:8, c("BMI", "waist", "height")]
demo$BMI[3] <- NA        # missing

demo_keep <- adiposity_sds(
  data = demo,
  col_map = col_map,
  ref = ref,
  na_action = "keep",
  extreme_action = "cap",
  sds_cap = 5,
  verbose = FALSE
)

demo_omit <- adiposity_sds(
  data = demo,
  col_map = col_map,
  ref = ref,
  na_action = "omit",
  extreme_action = "cap",
  sds_cap = 5,
  verbose = FALSE
)

list(
  keep_rows = nrow(demo_keep),
  omit_rows = nrow(demo_omit),
  capped_preview = demo_keep %>% select(ends_with("_SDS")) %>% slice_head(n = 3)
)
```

## Diagnostics summary with `return_summary`

Set `return_summary = TRUE` to get a list with the SDS tibble plus counts of omitted rows, raw extreme adjustments, and SDS extremes:

```{r}
res <- adiposity_sds(
  data           = demo,
  col_map        = col_map,
  ref            = ref,
  na_action      = "omit",
  extreme_action = "cap",
  sds_cap        = 5,
  return_summary = TRUE,
  verbose        = FALSE
)

res$summary
head(res$data)     # SDS tibble
res$warnings       # any diagnostic messages collected
```

## Verbose diagnostics

Set `verbose = TRUE` to emit three structured messages per call:

1. **Preparing inputs** — start-of-function signal.
2. **Column map** — confirms which data column each mapped variable resolved to.
   Example: `adiposity_sds(): column map: BMI -> 'BMI', waist -> 'waist', ...`
3. **Results summary** — shows how many rows computed successfully (non-NA) per output column.
   Example: `adiposity_sds(): results: BMI_SDS 28/30, waist_SDS 30/30, ...`

`verbose = TRUE` emits at the `"inform"` level; you also need `options(healthmarkers.verbose = "inform")` active:

```{r, message=TRUE}
old_opt <- options(healthmarkers.verbose = "inform")
invisible(adiposity_sds(
  data  = sim_small[1:5, ],
  col_map = col_map,
  ref   = ref,
  na_action = "keep"
))
options(old_opt)
```

Reset with `options(healthmarkers.verbose = NULL)` or `"none"`.

## Expectations
- Each mapped variable must exist; `ref` must provide finite `mean` and `sd > 0` for it.
- `na_action` controls row handling: `keep` propagates NA, `omit` drops, `error` aborts.
- `sds_cap` + `extreme_action` govern large SDS magnitudes.
- Numeric coercion warns if NAs are introduced; align data units with reference units.
- `return_summary = TRUE` returns data plus summaries of omissions and extremes.

## Common pitfalls
- Unit mismatches (e.g., inches vs cm) distort SDS; ensure units match the reference.
- Using a biased or tiny sample to build `ref` yields unstable SDS; use an appropriate reference population.
- Forgetting to map all variables in `col_map$vars` causes missing-column errors.

## Validation notes
- SDS should roughly center at 0 with SD near 1 if your sample matches the reference.
- Spot-check: `(value - ref_mean)/ref_sd` for one row should equal `<var>_SDS`.

## Column recognition

Run `hm_col_report(your_data)` to check which anthropometric columns are
auto-detected before building your `col_map`. See the
[Multi-Biobank Compatibility](multi_biobank.html) article for recognised synonyms.

```{r, eval=FALSE}
hm_col_report(your_data)
```

## See also
- `adiposity_sds_strat()` for sex-stratified SDS.
- `obesity_indices()` and `obesity_metrics()` for related adiposity measures.
