---
title: "Adipose insulin sensitivity indices"
output: rmarkdown::html_document
---

```{r, include=FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(purl = FALSE)
```

## Scope
Fasting adipose insulin sensitivity/resistance indices in one call: Revised_QUICKI, McAuley_index, Adipo_inv, Belfiore_inv_FFA, TyG_inv, TG_HDL_C_inv, and sex-specific VAI/LAP (inverted so more negative = worse). Inputs are fasting; common glucose/insulin/TG/HDL unit conversions are handled internally.

## When to use this
- You need multiple adipose IR markers for metabolic phenotyping or clustering.
- You have fasting glucose, insulin, lipids, and adiposity (waist/BMI); sex is optional but improves VAI/LAP labeling.
- You want built-in plausibility screening and NA policy controls.

## What you need
- Required columns: `G0` (mmol/L), `I0` (pmol/L), `TG` (mmol/L), `HDL_c` (mmol/L), `FFA` (mmol/L), `waist` (cm), `bmi` (kg/m^2).
- Fasting values only; post-prandial inputs will misstate IR.
- Units before conversion: glucose mmol/L, insulin pmol/L, TG mmol/L, HDL mmol/L, FFA mmol/L, waist cm, BMI kg/m^2. If different, convert first.
- Optional but helpful: `sex` (1/2 or labels) to understand which VAI/LAP variant applies.

## Load packages and data
Use a small subset of the simulated data. Swap `sim_small` for your own data.

```{r}
library(HealthMarkers)
library(dplyr)

sim_path <- system.file("extdata", "simulated_hm_data.rds", package = "HealthMarkers")
sim <- readRDS(sim_path)
sim_small <- dplyr::slice_head(sim, n = 30)
```

## Column map
Map required inputs; left-hand keys are fixed.

```{r}
col_map <- list(
  G0 = "G0",
  I0 = "I0",
  TG = "TG",
  HDL_c = "HDL_c",
  FFA = "FFA",
  waist = "waist",
  bmi = "BMI"
)
```

## Walkthrough (compute and read results)

`adipo_is()` returns a tibble of **only the computed index columns** — not the original data. Bind back to your source rows with `dplyr::bind_cols()` if needed.

```{r}
ais <- adipo_is(
  data = sim_small,
  col_map = col_map,
  na_action = "keep",
  normalize = "none"
)

# Group 1 — non-sex-specific indices
ais %>%
  slice_head(n = 5) %>%
  select(Revised_QUICKI, McAuley_index, Adipo_inv, Belfiore_inv_FFA, TyG_inv)

# Group 2 — sex-specific and ratio indices
ais %>%
  slice_head(n = 5) %>%
  select(TG_HDL_C_inv, VAI_Men_inv, VAI_Women_inv, LAP_Men_inv, LAP_Women_inv)
```

Values ending in `_inv` are inverted so **more negative = worse adipose IR**. Both sex variants (VAI_Men/Women, LAP_Men/Women) are always returned; pick the appropriate column per subject based on biological sex:

```{r, eval=FALSE}
# Bind sex back from source, then derive a single VAI and LAP per subject.
# Adjust the if_else condition to match your sex coding (e.g. 1/2, "M"/"F").
ais_with_sex <- dplyr::bind_cols(dplyr::select(sim_small, sex), ais)

ais_with_sex %>%
  dplyr::mutate(
    VAI_inv = dplyr::if_else(sex == 1, VAI_Men_inv, VAI_Women_inv),
    LAP_inv = dplyr::if_else(sex == 1, LAP_Men_inv, LAP_Women_inv)
  ) %>%
  dplyr::select(sex, Revised_QUICKI, TyG_inv, VAI_inv, LAP_inv) %>%
  slice_head(n = 5)
```

## Missing data handling
Compare NA policies on a small slice.

```{r}
demo <- sim_small[1:6, c("G0", "I0", "TG", "HDL_c", "FFA", "waist", "BMI")]
demo$G0[2] <- NA        # missing glucose

ais_keep <- adipo_is(
  data = demo,
  col_map = col_map,
  na_action = "keep",
  normalize = "none",
  verbose = FALSE
)

ais_omit <- adipo_is(
  data = demo,
  col_map = col_map,
  na_action = "omit",
  normalize = "none",
  verbose = FALSE
)

list(
  keep_rows = nrow(ais_keep),
  omit_rows = nrow(ais_omit),
  keep_preview = ais_keep %>% select(ends_with("_inv")) %>% slice_head(n = 3)
)
```

## Normalise outputs

Use `normalize` to put all indices on a common scale before modelling or dimension reduction:

```{r}
ais_z <- adipo_is(
  data      = sim_small,
  col_map   = col_map,
  na_action = "keep",
  normalize = "z"      # z-score: mean 0, SD 1
)

ais_z %>%
  slice_head(n = 5) %>%
  select(Revised_QUICKI, McAuley_index, TyG_inv, Adipo_inv)
```

Available options: `"none"` (default — raw scale), `"z"` (mean 0 / SD 1), `"range"` (0–1 min–max), `"robust"` (median / IQR), `"inverse"` (1/x).

## Verbose diagnostics

Set `verbose = TRUE` to emit three structured messages per call:

1. **Preparing inputs** — start-of-function signal.
2. **Column map** — confirms which data column each required key resolved to.
   Example: `adipo_is(): column map: G0 -> 'G0', I0 -> 'I0', TG -> 'TG', ...`
3. **Results summary** — shows how many rows computed successfully (non-NA) per output column.
   Example: `adipo_is(): results: Revised_QUICKI 28/30, VAI_Men_inv 28/30, McAuley_index 27/30, ...`

`verbose = TRUE` emits at the `"inform"` level; you also need `options(healthmarkers.verbose = "inform")`
(or `"debug"`) active in the session for the messages to print:

```{r, message=TRUE}
old_opt <- options(healthmarkers.verbose = "inform")

invisible(adipo_is(
  data          = sim_small[1:5, ],
  col_map       = col_map,
  na_action     = "keep",
  normalize     = "none",
  verbose       = TRUE
))

options(old_opt)  # restore
```

Enable globally for an entire session: `options(healthmarkers.verbose = "inform")`.
Reset with `options(healthmarkers.verbose = NULL)` or `options(healthmarkers.verbose = "none")`.

## Expectations
- All seven required mappings must be provided; missing mappings or columns abort.
- `na_action` controls row handling: `keep` propagates NA outputs; `omit` drops rows with missing required inputs; `error` aborts.
- Built-in conversions: glucose *18 (mg/dL), insulin /6 (muU/mL), TG *88.57 (mg/dL), HDL *38.67 (mg/dL); FFA/waist/BMI unchanged.

## Common pitfalls
- Non-fasting samples inflate TG/glucose and distort TyG/VAI/LAP.
- Mis-specified units (e.g., mg/dL glucose without converting) will skew every index; convert upstream if your lab units differ.
- Missing `sex` does not block computation but only one of VAI/LAP applies per subject; interpret accordingly.
- For modelling, use `normalize = "z"` (z-score), `"range"` (0–1), or `"robust"` (median/IQR); default `"none"` keeps raw index scales. See the *Normalise outputs* section above.
## Validation notes
- **Revised_QUICKI**: typical adults ~0.3–0.6; lower is worse IR. Computed from log10(I0 [mU/L]) + log10(G0 [mg/dL]) + log10(FFA [mmol/L]).
- **McAuley_index**: typical adults ~3–10; higher is better insulin sensitivity. Uses `ln(TG [mmol/L])` and `ln(I0 [mU/L])` — TG is kept in mmol/L before the internal mg/dL conversion.
- **TyG_inv**: typically −8 to −12 for adults; more negative = worse IR. Computed from TG and G0 both in mg/dL.
- **LAP_Men_inv / LAP_Women_inv**: computed as `−(WC−threshold) × TG [mmol/L]` (Kahn 2005). Typical values −10 to −80; more negative = worse.
- **VAI_Men_inv / VAI_Women_inv**: uses TG and HDL in mmol/L with the Amato 2010 reference constants (1.03, 1.31, 0.81, 1.52). Typical range −1 to −5; more negative = worse.
- Spot-check `TG_HDL_C_inv`: should equal `−(TG_mg/dL / HDL_mg/dL)`.

## See also
- `glycemic_markers()` for glucose-centric indices.
- `insulin_secretion()` and `insulin_sensitivity()` vignettes for complementary beta-cell metrics.
