---
title: "Allostatic Load"
output: rmarkdown::html_document
---

```{r, include=FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(purl = FALSE)
```

## Scope
Compute a composite Allostatic Load (AL) score: each biomarker exceeding its threshold contributes 1 to the count. With multiple biomarkers the rule is `>`; with a single biomarker it is `>=`. Returns a tibble with a single integer column `AllostaticLoad`. Optional extreme screening applies only to columns whose names contain "sds".

## When to use this
- You want a simple count of biomarkers exceeding risk cut-points.
- You have a small set of biomarkers with clear thresholds you control.
- You may include pre-computed SDS columns and optionally cap implausible SDS values.

## What you need
- Thresholds: named list of scalar numeric cutoffs; names must match columns (or be mapped via `col_map`).
- Data: matching biomarker columns; units should align with the thresholds you supply.
- Policies: `na_action` (`keep` treats missing as unflagged; `omit` drops rows; `error` aborts).
- Optional: `return_summary = TRUE` for counts of flags/omissions; `verbose` for progress.

## Load packages and data
Use a small subset of the simulated data. Replace `sim_small` with your data.

```{r}
library(HealthMarkers)
library(dplyr)

sim_path <- system.file("extdata", "simulated_hm_data.rds", package = "HealthMarkers")
sim <- readRDS(sim_path)
sim_small <- dplyr::slice_head(sim, n = 30)
```

## Column map and thresholds
Provide thresholds and, if needed, map them to your column names.

```{r}
thresholds <- list(sbp = 130, dbp = 85, CRP = 3)
col_map <- list(sbp = "sbp", dbp = "dbp", CRP = "CRP")
```

## Walkthrough (compute Allostatic Load)

```{r}
al <- allostatic_load(
  data = sim_small,
  thresholds = thresholds,
  col_map = col_map,
  na_action = "keep",
  verbose = FALSE
)

al %>% slice_head(n = 5)
```

Interpretation: `AllostaticLoad` is the count of biomarkers above thresholds. With >1 biomarker the rule is `>`; with a single biomarker it is `>=`.

## Missing data handling
Demonstrate row handling.

```{r}
demo <- sim_small[1:8, c("sbp", "dbp", "CRP")]
demo$sbp[3] <- NA

al_keep <- allostatic_load(
  data = demo,
  thresholds = thresholds,
  col_map = col_map,
  na_action = "keep",
  verbose = FALSE
)

al_omit <- allostatic_load(
  data = demo,
  thresholds = thresholds,
  col_map = col_map,
  na_action = "omit",
  verbose = FALSE
)

list(
  keep_rows = nrow(al_keep),
  omit_rows = nrow(al_omit),
  head_keep = al_keep %>% slice_head(n = 3)
)
```

## Diagnostics summary with `return_summary`

`return_summary = TRUE` returns `list(data, summary, warnings)` instead of a bare tibble. The `summary` element contains `rows`, `biomarkers`, `total_flags`, and `mean_flags`.

```{r}
df_rs <- data.frame(
  SBP = c(118, 142, 130, 155, 120),
  DBP = c(76,   92,  85,  98,  78),
  CRP = c(1.2,  4.8, 2.1, 5.0, 0.9)
)
thr_rs <- list(SBP = 130, DBP = 85, CRP = 3)

res <- allostatic_load(
  data      = df_rs,
  thresholds = thr_rs,
  na_action  = "keep",
  return_summary = TRUE,
  verbose    = FALSE
)
res$summary
res$data
```

## Verbose diagnostics

Set `verbose = TRUE` to emit three structured messages per call:

1. **Preparing inputs** — start-of-function signal.
2. **Column map** — confirms which data column each threshold key resolved to.
   Example: `allostatic_load(): column map: SBP -> 'SBP', DBP -> 'DBP', ...`
3. **Results summary** — shows how many rows computed successfully (non-NA) per output column.
   Example: `allostatic_load(): results: AllostaticLoad 30/30, ...`

`verbose = TRUE` emits at the `"inform"` level; you also need `options(healthmarkers.verbose = "inform")` active:

```{r, message=TRUE}
old_opt <- options(healthmarkers.verbose = "inform")

df_v <- data.frame(SBP = c(118, 142), DBP = c(76, 92))
allostatic_load(df_v, thresholds = list(SBP = 130, DBP = 85), verbose = TRUE)

options(old_opt)
```

Reset with `options(healthmarkers.verbose = NULL)` or `"none"`.

## Expectations
- Threshold names must match columns (or be mapped); missing columns abort.
- `na_action`: `keep` treats missing as unflagged (flag = 0), `omit` drops rows, `error` aborts on missing.
- Multi-biomarker rule is `>`; single-biomarker rule is `>=`.

## Common pitfalls
- Threshold units mismatched to data units will misflag results; align units first.
- Passing only one biomarker changes the rule to `>=`; set thresholds accordingly.
- `na_action = keep` leaves missing inputs unflagged (zero); use `omit` for stricter handling.

## Validation notes
- Spot-check one biomarker: flag = 1 if value exceeds the threshold (or equals when only one biomarker), else 0.
- Sum of per-biomarker flags should equal `AllostaticLoad` for each row.

## Column recognition

Run `hm_col_report(your_data)` to check which biomarker columns are auto-detected
before building your `col_map`. See the
[Multi-Biobank Compatibility](multi_biobank.html) article for recognised synonyms.

```{r, eval=FALSE}
hm_col_report(your_data)
```

## See also
- `adiposity_sds()` and `adiposity_sds_strat()` for SDS creation.
- `health_markers()` and `health_summary()` for broader panels.
