---
title: "Atherogenic Indices"
output: rmarkdown::html_document
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
knitr::opts_knit$set(purl = FALSE)
```

## Scope
Compute atherogenic risk ratios from a lipid panel:
- AIP = log10(TG / HDL_c)
- Castelli Risk Index I (CRI_I) = TC / HDL_c
- Castelli Risk Index II (CRI_II) = LDL_c / HDL_c

The Castelli indices (CRI_I, CRI_II) are true scale-invariant ratios — identical
in mg/dL or mmol/L. **AIP is different:** because TG and HDL-c convert to mg/dL
with different factors (×88.57 vs ×38.67), `log10(TG / HDL_c)` is **not**
scale-invariant — mg/dL inputs shift AIP by about +0.36 and invalidate the
published risk strata. AIP (Dobiasova 2004) is defined for **mmol/L**, so supply
TG and HDL_c in mmol/L.

## When to use
- You have triglycerides and HDL cholesterol and want AIP.
- You also have total and/or LDL cholesterol and want Castelli indices.
- You need strict missing-data handling.

## Requirements checklist
- Packages: HealthMarkers, dplyr (for display).
- Data frame with lipid columns; required keys: TG and HDL_c. Optional: TC and LDL_c.
- Numeric lipids; non-numeric are coerced with warnings; non-finite become NA.
- Decide row policy: na_action = keep (default), omit, or error.

## Load packages and example data
Replace the example slice with your data frame.

```{r}
library(HealthMarkers)
library(dplyr)

sim_path <- system.file("extdata", "simulated_hm_data.rds", package = "HealthMarkers")
sim <- readRDS(sim_path)
sim_small <- dplyr::slice_head(sim, n = 30)
```

## Map lipid columns
TG and HDL_c are required. TC and LDL_c are optional but needed for Castelli indices.

```{r}
col_map <- list(
  TG = "TG",
  HDL_c = "HDL_c",
  TC = "TC",
  LDL_c = "LDL_c"
)
```

## Quick start: compute indices
Defaults keep rows with missing lipids and return NA for their ratios.

```{r}
aip_out <- atherogenic_indices(
  data = sim_small,
  col_map = col_map,
  na_action = "keep",
  verbose = FALSE
)

new_cols <- setdiff(names(aip_out), names(sim_small))
head(select(aip_out, all_of(new_cols)))
```

## Arguments that matter
- col_map: named list for TG, HDL_c (required), and optional TC, LDL_c. Missing required keys error.
- na_action: keep (default, rows retained; ratios become NA), omit (drop rows with NA in used lipids), error (abort on NA).
- verbose: emit step messages.

## Handling missing and zero denominators
- Non-numeric lipids are coerced to numeric; NA introduced are warned. Non-finite become NA.
- Zero HDL_c denominators yield NA ratios and a warning; Castelli indices use HDL_c as the divisor.
- Row policy: keep vs omit vs error via na_action.

### Compare row policies
```{r}
demo <- sim_small
demo$HDL_c[c(2, 5)] <- NA

aip_keep <- atherogenic_indices(demo, col_map, na_action = "keep")
aip_omit <- atherogenic_indices(demo, col_map, na_action = "omit")

list(
  keep_rows = nrow(aip_keep),
  omit_rows = nrow(aip_omit),
  sample_ratios = head(select(aip_keep, AIP, CRI_I, CRI_II))
)
```

## Extreme-value screening (optional)
Zero or negative HDL_c denominators yield NA ratios and a warning. Filter or remove such values before calling.

```{r}
demo2 <- sim_small
demo2$TG[5] <- 12000  # extreme TG
# Note: extreme values will produce extreme ratio outputs; pre-filter if needed
head(select(atherogenic_indices(demo2, col_map = col_map), AIP, CRI_I, CRI_II))
```

## Outputs
- AIP: log10(TG / HDL_c)
- CRI_I: TC / HDL_c (NA if TC not mapped/present)
- CRI_II: LDL_c / HDL_c (NA if LDL_c not mapped/present)
Rows only drop when na_action = "omit" or when na_action = "error" encounters missing.

## Pitfalls and tips
- CRI_I and CRI_II are scale-invariant ratios; AIP is not, so supply TG and HDL_c in mmol/L (mg/dL shifts AIP by ~+0.36 and breaks the published cut-offs).
- Mapping omissions for TG or HDL_c throw errors; missing TC/LDL_c just yield NA Castelli indices.
- Zero HDL_c produces NA ratios and a warning; address true zeros before computing.
- Use na_action = "error" for pipelines that must fail fast on missing lipids.

## Validation ideas
- Spot-check AIP (mmol/L): if TG = 1.5 and HDL_c = 1.0, expected AIP = log10(1.5) ~ 0.176.
- CRI_I should decrease if HDL_c rises with other lipids unchanged; similar for CRI_II.
- Confirm that rows with NA in TG/HDL_c are dropped only when na_action = "omit".

## Verbose diagnostics

Set `verbose = TRUE` to emit three structured messages per call:

1. **Preparing inputs** — start-of-function signal.
2. **Column map** — confirms which data column each lipid key resolved to.
   Example: `atherogenic_indices(): column map: TG -> 'TG', HDL_c -> 'HDL_c', TC -> 'TC', LDL_c -> 'LDL_c'`
3. **Results summary** — shows how many rows computed successfully (non-NA) per output column.
   Example: `atherogenic_indices(): results: AIP 28/30, CRI_I 30/30, CRI_II 30/30, ...`

`verbose = TRUE` emits at the `"inform"` level; you also need `options(healthmarkers.verbose = "inform")` active:

```{r, message=TRUE}
old_opt <- options(healthmarkers.verbose = "inform")

df_v <- tibble::tibble(
  TG    = c(1.7, 2.3),   # mmol/L
  HDL_c = c(1.3, 1.0),   # mmol/L
  TC    = c(5.2, 5.7),   # mmol/L
  LDL_c = c(3.1, 3.9)    # mmol/L
)
atherogenic_indices(
  df_v,
  col_map = list(TG = "TG", HDL_c = "HDL_c", TC = "TC", LDL_c = "LDL_c"),
  verbose = TRUE
)

options(old_opt)
```

Reset with `options(healthmarkers.verbose = NULL)` or `"none"`.

## Column recognition

Run `hm_col_report(your_data)` to check which lipid/analyte columns are
auto-detected before building your `col_map`. See the
[Multi-Biobank Compatibility](multi_biobank.html) article for recognised synonyms.

```{r, eval=FALSE}
hm_col_report(your_data)
```

## See also
- health_summary() for combined marker panels.
- glycemic_markers() for glucose and insulin-derived indices.
