---
title: "Charlson Comorbidity Index"
output: rmarkdown::html_document
---

```{r, include=FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(purl = FALSE)
```

## Scope
Compute the 19-condition Charlson Comorbidity Index (CCI) without age points. Handles binary indicators, optional bounds screening, and NA policies.

## When to use
- You have binary comorbidity indicators and need the canonical Charlson score (no age points here).
- You want explicit row-retention rules for missing indicators.

## Requirements checklist
- Packages: HealthMarkers, dplyr (for display).
- Data columns: binary 0/1 indicators for the 19 conditions.
- Column map: provide all 19 keys -> columns; missing mappings abort.
- Row policy: na_action = keep (default), omit, error; warn/ignore act like keep but warn.

## Load packages and example data
Illustrative-only flags below; replace with your real comorbidity indicators.

```{r}
library(HealthMarkers)
library(dplyr)

sim_path <- system.file("extdata", "simulated_hm_data.rds", package = "HealthMarkers")
sim <- readRDS(sim_path)
sim_small <- dplyr::slice_head(sim, n = 30)

# Illustrative binary flags
set.seed(123)
sim_small <- sim_small %>% mutate(
  mi = rbinom(n(), 1, 0.1),
  chf = rbinom(n(), 1, 0.1),
  pvd = rbinom(n(), 1, 0.1),
  stroke = rbinom(n(), 1, 0.1),
  dementia = rbinom(n(), 1, 0.05),
  copd = rbinom(n(), 1, 0.15),
  rheum = rbinom(n(), 1, 0.05),
  ulcer = rbinom(n(), 1, 0.05),
  mild_liver = rbinom(n(), 1, 0.05),
  diabetes = rbinom(n(), 1, 0.2),
  diab_comp = rbinom(n(), 1, 0.05),
  hemiplegia = rbinom(n(), 1, 0.02),
  renal = rbinom(n(), 1, 0.08),
  cancer = rbinom(n(), 1, 0.08),
  leukemia = rbinom(n(), 1, 0.01),
  lymphoma = rbinom(n(), 1, 0.01),
  sev_liver = rbinom(n(), 1, 0.02),
  metastatic_cancer = rbinom(n(), 1, 0.01),
  hiv = rbinom(n(), 1, 0.01)
)
```

## Map columns
Provide all 19 keys. Missing mappings error.

```{r}
col_map <- list(
  mi="mi", chf="chf", pvd="pvd", stroke="stroke", dementia="dementia", copd="copd",
  rheum="rheum", ulcer="ulcer", mild_liver="mild_liver", diabetes="diabetes",
  diab_comp="diab_comp", hemiplegia="hemiplegia", renal="renal", cancer="cancer",
  leukemia="leukemia", lymphoma="lymphoma", sev_liver="sev_liver",
  metastatic_cancer="metastatic_cancer", hiv="hiv"
)
```

## Quick start: compute CCI
Defaults keep rows with missing indicators and return NA scores for them.

```{r}
cci_out <- charlson_index(
  data = sim_small,
  col_map = col_map,
  na_action = "keep",
  verbose = FALSE
)

head(cci_out)
```

## Arguments that matter
- col_map: all 19 keys required; missing or empty entries abort.
- na_action: keep (default), omit (drop rows with missing indicators), error (abort on missing); warn/ignore behave like keep but warn.
- verbose: emit step messages.

## Handling missing and non-binary inputs
- Non-numeric inputs are coerced; NA introduced are warned. Non-finite become NA.
- Missing indicators yield NA scores unless na_action = "omit" or "error".
- Out-of-range (non 0/1) values warn; recode non-binary values to 0/1 before passing.

### Compare row policies
```{r}
demo <- sim_small[1:8, names(col_map)]
demo$renal[3] <- NA

a_keep <- charlson_index(demo, col_map, na_action = "keep")
a_omit <- charlson_index(demo, col_map, na_action = "omit")

list(
  keep_rows = nrow(a_keep),
  omit_rows = nrow(a_omit),
  preview = head(a_keep$charlson_index)
)
```

## Out-of-bounds indicators
Indicators outside 0-1 will warn. Pre-filter or recode non-binary values before calling.

```{r}
demo2 <- demo
demo2$mi[5] <- 2  # out of bounds — will warn
# Recode to valid binary before passing if needed
demo2$mi[5] <- as.integer(demo2$mi[5] > 0)
head(charlson_index(demo2, col_map = col_map, na_action = "keep")$charlson_index)
```

## Outputs
- charlson_index integer total
- Rows drop only with na_action = "omit"; na_action = "error" aborts on missing.

## Pitfalls and tips
- Indicators must be truly binary; non-binary values warn; recode before passing.
- Provide all 19 mappings; missing keys error.
- Age points are not included here; add separately if needed.

## Validation ideas
- Set all indicators to 0: expected score 0.
- Single metastatic_cancer = 1, all else 0: expected score 6.
- diabetes = 1 and diab_comp = 1: expected diabetes contribution = 2 (max rule).

## Verbose diagnostics

Set `verbose = TRUE` to emit three structured messages per call:

1. **Preparing inputs** — start-of-function signal.
2. **Column map** — confirms which data column each comorbidity key resolved to.
   Example: `charlson_index(): column map: mi -> 'mi', chf -> 'chf', ...`
3. **Results summary** — shows how many rows computed successfully (non-NA) per output column.
   Example: `charlson_index(): results: charlson_index 30/30, charlson_index_age 30/30, ...`

`verbose = TRUE` emits at the `"inform"` level; you also need `options(healthmarkers.verbose = "inform")` active:

```{r, message=TRUE}
old_opt <- options(healthmarkers.verbose = "inform")

patient <- tibble::tibble(
  mi=0, chf=0, pvd=0, stroke=0, dementia=0, copd=0, rheum=0, ulcer=0,
  mild_liver=0, diabetes=1, diab_comp=1, hemiplegia=0, renal=1,
  cancer=0, leukemia=0, lymphoma=0, sev_liver=0, metastatic_cancer=0, hiv=0
)
cm_v <- as.list(stats::setNames(names(patient), names(patient)))
charlson_index(patient, col_map = cm_v, verbose = TRUE)

options(old_opt)
```

Reset with `options(healthmarkers.verbose = NULL)` or `"none"`.

## See also
- metss and metabolic_risk_features for metabolic burden scoring.
- health_summary() to combine multiple indices.
