---
title: "Corrected Calcium"
output: rmarkdown::html_vignette
---

```{r, include=FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(purl = FALSE)
```

## Scope
Apply the Payne formula to albumin-adjust serum calcium. Works with conventional (mg/dL Ca, g/dL Alb) or SI (mmol/L Ca, g/L Alb); can auto-detect units; supports optional extreme screening.

## When to use
- You have total serum calcium and albumin and need albumin-corrected calcium.
- Your labs may be in mg/dL+g/dL or mmol/L+g/L, and you want unit-aware handling.
- You need row-retention rules for missing inputs and optional bounds checks.

## Requirements checklist
- Packages: HealthMarkers, dplyr (for display).
- Columns: calcium and albumin; units must be consistent with the chosen `units` argument.
- Column map: list(calcium = ..., albumin = ...). Defaults assume Ca/Alb.
- Units: conventional (mg/dL, g/dL), si (mmol/L, g/L), or auto (infers; warns when assuming SI).
- Row policy: na_action = keep (default), omit, error; warn/ignore behave like keep but warn.
- Optional screening: check_extreme + extreme_action (warn/cap/error/ignore/NA); defaults use Ca 4-15 mg/dL and Alb 2-5 g/dL in working units when enabled.

## Load packages and example data
Simulated data lack Ca/Alb; we generate illustrative valuesreplace with real labs.

```{r}
library(HealthMarkers)
library(dplyr)

sim_path <- system.file("extdata", "simulated_hm_data.rds", package = "HealthMarkers")
sim <- readRDS(sim_path)
sim_small <- dplyr::slice_head(sim, n = 30)

set.seed(123)
# Illustrative labs (mg/dL Ca, g/dL Alb)
sim_small$Ca <- pmax(7.5, pmin(11.5, rnorm(nrow(sim_small), 9.2, 0.6)))
sim_small$Alb <- pmax(2.5, pmin(5.0, rnorm(nrow(sim_small), 4.0, 0.4)))
```

## Map columns
```{r}
col_map <- list(calcium = "Ca", albumin = "Alb")
```

## Quick start: compute corrected calcium
Defaults keep rows with missing inputs and do not screen extremes.

```{r}
cc_out <- corrected_calcium(
  data = sim_small,
  col_map = col_map,
  units = "conventional",  # use "si" for mmol/L; "auto" to infer
  na_action = "keep",
  check_extreme = FALSE,
  extreme_action = "warn"
)

head(cc_out)
```

## Arguments that matter
- col_map: calcium and albumin mappings are required; empty or missing mappings abort.
- units: conventional | si | auto. Auto infers based on medians and warns if assuming SI.
- na_action: keep (rows retained, NA outputs for missing), omit (drop rows with missing inputs), error (abort on missing).
- check_extreme: FALSE by default; TRUE applies bounds in working units (mg/dL, g/dL) before correction.
- extreme_action: warn (default, no change), cap, error, ignore, or NA; override bounds via extreme_rules.

## Handling missing and non-numeric inputs
- Non-numeric Ca/Alb are coerced; NA introduced are warned. Non-finite become NA.
- Missing inputs yield NA outputs unless na_action = omit/error.

### Compare row policies
```{r}
demo <- sim_small[1:8, c("Ca", "Alb")]
demo$Ca[3] <- NA

a_keep <- corrected_calcium(demo, col_map, units = "conventional", na_action = "keep")
a_omit <- corrected_calcium(demo, col_map, units = "conventional", na_action = "omit")

list(
  keep_rows = nrow(a_keep),
  omit_rows = nrow(a_omit),
  preview = head(a_keep)
)
```

## Extreme-value screening (optional)
Cap or error on implausible inputs before correction.

```{r}
demo2 <- demo
demo2$Ca[5] <- 2.0  # implausibly low if treated as mg/dL

a_cap <- corrected_calcium(
  data = demo2,
  col_map = col_map,
  units = "conventional",
  na_action = "keep",
  check_extreme = TRUE,
  extreme_action = "cap",
  extreme_rules = list(ca_mgdl = c(4, 15), alb_gdl = c(2, 5))
)

head(a_cap)
```

## Verbose diagnostics

Set `verbose = TRUE` to emit three structured messages per call:

1. **Preparing inputs** — start-of-function signal.
2. **Column map** — confirms which data column each required key (`calcium`, `albumin`) resolved to.
   Example: `corrected_calcium(): column map: calcium -> 'Ca', albumin -> 'Alb'`
3. **Results summary** — shows how many rows computed successfully (non-NA) per output column.
   Example: `corrected_calcium(): results: corrected_calcium 28/30, ...`

`verbose = TRUE` emits at the `"inform"` level; you also need `options(healthmarkers.verbose = "inform")` active.
Reset with `options(healthmarkers.verbose = NULL)` or `options(healthmarkers.verbose = "none")`.

```{r, message=TRUE}
old_opt <- options(healthmarkers.verbose = "inform")

cc_verbose <- corrected_calcium(
  data    = sim_small,
  col_map = col_map,
  units   = "conventional",
  verbose = TRUE
)

options(old_opt)   # restore original setting
```

## Outputs
- corrected_calcium (mg/dL when units = conventional; mmol/L when units = si or auto-inferred SI)
- Rows drop only with na_action = "omit"; na_action = "error" aborts on missing.

## Pitfalls and tips
- Prefer explicit units; auto mode may warn and convert if SI is inferred.
- Ensure Ca and Alb units match the chosen mode; mismatched units will mis-scale results.
- Tighten extreme_rules to your lab ranges; use extreme_action = "error" for strict QC.
- Downstream consumers expecting mg/dL should keep units = "conventional" or convert explicitly.

## Validation
```{r}
# Payne formula spot check (conventional): Ca = 9.0, Alb = 3.0 => 9.0 + 0.8*(4-3) = 9.8
v1 <- corrected_calcium(data.frame(Ca = 9.0, Alb = 3.0), col_map, units = "conventional")
stopifnot(abs(v1$corrected_calcium - 9.8) < 1e-10)

# SI round-trip: Ca = 2.25 mmol/L, Alb = 35 g/L
# working: Ca_mg = 2.25*4 = 9.0, Alb_gdl = 35/10 = 3.5
# corrected_mg = 9.0 + 0.8*(4 - 3.5) = 9.4; corrected_mmol = 9.4/4 = 2.35
v2 <- corrected_calcium(data.frame(Ca = 2.25, Alb = 35), col_map, units = "si")
stopifnot(abs(v2$corrected_calcium - 2.35) < 1e-10)

cat("Payne conventional:", v1$corrected_calcium, "mg/dL\n")
cat("Payne SI round-trip:", v2$corrected_calcium, "mmol/L\n")
```

## See also
- renal_markers() for broader kidney panels.
- health_summary() to combine lab markers.
