---
title: "FRAX fracture risk"
output: rmarkdown::html_document
---

```{r, include=FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(purl = FALSE)
```

## Scope
Compute a simplified, non-validated approximation of FRAX 10-year fracture risk (major osteoporotic and hip). Normalizes sex labels, applies HM-CS missing/extreme policies, and returns placeholder risks; it is **not** the proprietary FRAX algorithm.

## When to use
- You need a placeholder fracture-risk proxy for demos or pipelines while acknowledging it is not clinical FRAX.
- You want explicit NA and extreme-value handling (age/BMD bounds) with column mapping flexibility.
- You plan to swap in real FRAX later; this keeps interfaces stable in the meantime.

## Inputs and requirements
- Required columns (map via `col_map`): `age` (years), `sex` (codes starting with m/f or 1/0).
- Optional binary risk factors: prior fracture, parent fracture, glucocorticoids, rheumatoid, secondary osteoporosis, current smoker, high alcohol.
- Optional: femoral neck BMD T-score (`bmd`).
- `na_action`: `keep`/`ignore`/`warn` retain rows with NA outputs; `omit` drops rows missing required inputs; `error` aborts.

## Load packages and data
Illustrative demo data only; replace with your cohort.

```{r}
library(HealthMarkers)

demo <- tibble::tibble(
  Age = c(65, 72, 80, 68, 74, 60),
  Sex = c("F", "M", "F", "F", "M", "F"),
  prior_fx   = c(1, 0, 1, 0, 0, 0),
  parent_fx  = c(0, 1, 0, 0, 0, 0),
  steroids   = c(0, 1, 0, 0, 1, 0),
  rheumatoid = c(0, 0, 0, 1, 0, 0),
  secondary  = c(0, 0, 0, 0, 0, 0),
  smoker     = c(0, 1, 0, 0, 1, 0),
  alcohol    = c(1, 0, 0, 0, 0, 1),
  bmd        = c(-2.0, -1.0, -2.5, -1.5, -0.8, NA)
)
```

## Column map (required)
Provide your column names; left-hand keys are fixed.

```{r}
col_map <- list(
  age = "Age",
  sex = "Sex",
  prior_fracture  = "prior_fx",
  parent_fracture = "parent_fx",
  steroids        = "steroids",
  rheumatoid      = "rheumatoid",
  secondary_op    = "secondary",
  smoker          = "smoker",
  alcohol         = "alcohol",
  bmd             = "bmd"
)
```

### What each input represents
- `age`: age in years (required).
- `sex`: sex code; first letter `m`/`f` or `1`/`0` (required; normalized to male/female).
- Optional binary risk factors: prior fracture, parent fracture, glucocorticoids, rheumatoid, secondary osteoporosis, current smoker, high alcohol.
- `bmd`: femoral neck T-score (optional); placeholder adjustment when present.

## Core calculation
Return the placeholder risks and the normalized age/sex/BMD used.

```{r}
frax_out <- frax_score(
  data = demo,
  col_map = col_map,
  na_action = "keep",
  verbose = FALSE
)

head(dplyr::select(frax_out, frax_major_percent, frax_hip_percent, frax_sex_norm, frax_age_used, frax_bmd_tscore))
```

Interpretation: `frax_major_percent` and `frax_hip_percent` are placeholder probabilities (not clinical). `frax_sex_norm` shows normalized sex; `frax_age_used` and `frax_bmd_tscore` reflect any capping/blanking applied by extreme checks.

## Inputs and reminders
- Required: `age` (years) and `sex` codes mapped in `col_map`.
- Optional: binary risk factors and BMD T-score. Non-numeric inputs are coerced; non-finite values become NA.
- `na_action`: `keep`/`ignore`/`warn` retain rows with NA outputs; `omit` drops rows missing required inputs; `error` aborts.

## Missing data handling
Show row handling on a small slice.

```{r}
demo_miss <- demo
demo_miss$Age[3] <- NA      # missing age
demo_miss$Sex[4] <- "X"    # unknown sex -> NA

keep_res <- frax_score(
  data = demo_miss,
  col_map = col_map,
  na_action = "keep",
  verbose = FALSE
)

omit_res <- frax_score(
  data = demo_miss,
  col_map = col_map,
  na_action = "omit",
  verbose = FALSE
)

list(
  keep_rows = nrow(keep_res),
  omit_rows = nrow(omit_res),
  preview = head(dplyr::select(keep_res, frax_age_used, frax_bmd_tscore, frax_major_percent, frax_hip_percent))
)
```

Here `na_action = "keep"` retains rows even with missing/unknown sex or age (outputs become NA). `na_action = "omit"` drops rows missing required inputs.

## Expectations
- Placeholder approximation only; not a validated FRAX calculator and not for clinical use.
- You must map `age` and `sex`; missing mappings or columns abort.
- Choose `na_action` to enforce or relax completeness; `error` is strict, `omit` drops, `keep` retains.
- `verbose = TRUE` surfaces progress and QA messages.

## Verbose diagnostics
Set `verbose = TRUE` (and `healthmarkers.verbose = "inform"`) to surface three structured messages on each call: preparing inputs, the column map, and a results summary.

```{r, message=TRUE}
old_opt <- options(healthmarkers.verbose = "inform")
df_v <- data.frame(Age = c(65, 72), Sex = c("F", "M"),
                   bmd = c(-2.0, -1.0))
frax_score(
  df_v,
  col_map = list(age = "Age", sex = "Sex", bmd = "bmd"),
  verbose = TRUE
)
options(old_opt)
```

## Column recognition

Run `hm_col_report(your_data)` to check which analyte columns are auto-detected
before building your `col_map`. See the
[Multi-Biobank Compatibility](multi_biobank.html) article for recognised synonyms
across major biobanks.

```{r, eval=FALSE}
hm_col_report(your_data)
```

## Tips
- Normalize sex coding to `M`/`F` or 1/0 before calling to avoid NA normalization.
- Use `na_action = "error"` for strict pipelines; `omit` to filter incomplete rows; `keep`/`warn` for exploratory use.
- Outputs are illustrative only; replace with a validated FRAX implementation for clinical workflows.
