---
title: "Kidney failure risk (KFRE)"
output: rmarkdown::html_document
---

```{r, include=FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(purl = FALSE)
```

## Scope
4-variable Kidney Failure Risk Equation (KFRE) for 2- and 5-year ESKD risk. Includes NA policies, high-missingness warnings, and verbose summaries. Units: age (years), sex coded 1=male/2=female, eGFR (mL/min/1.73 m^2), UACR (mg/g); no unit conversion is performed.

## Clinical context

The KFRE (Tangri *et al.*, 2011, *JAMA*; validated in 3-continent cohort) estimates
2- and 5-year probability of kidney failure (ESKD) from four routine variables.
It is now embedded in the KDIGO 2022 CKD guidelines for risk stratification and
nephrology referral decisions.

| CKD Stage | eGFR (mL/min/1.73 m²) | 5-yr KFRE (typical range) |
|-----------|----------------------|---------------------------|
| G3a | 45–59 | < 1% |
| G3b | 30–44 | 1–5% |
| G4 | 15–29 | 5–40% |
| G5 | < 15 | > 40% (often > 70%) |

Outputs `KFRE_2yr` and `KFRE_5yr` are probabilities (0–1); multiply by 100 for
percentages. The equation was developed in adults with CKD G3–G5; interpret
cautiously outside this range.

## Load packages and demo data

```{r}
library(HealthMarkers)
library(tibble)

# Ten participants spanning CKD G3a → G5; mixed sex; UACR A1 → A3
df <- tibble::tibble(
  age  = c(45, 52, 60, 68, 72, 55, 63, 75, 80, 48),
  sex  = c(1,  2,  1,  2,  1,  2,  1,  2,  1,  2 ),
  eGFR = c(58, 50, 42, 35, 28, 22, 18, 14, 10,  8 ),
  UACR = c(25, 80, 200, 500, 800, 1200, 2500, 4500, 8000, 6000)
)
```

## Column map (required)

```{r}
col_map <- list(age = "age", sex = "sex", eGFR = "eGFR", UACR = "UACR")
```

## Core calculation
Default `na_action = "keep"`: missing inputs propagate to `NA` risks; no extreme checks.

```{r}
kfre_default <- kidney_failure_risk(
  data = df,
  col_map = col_map,
  na_action = "keep",
  verbose = FALSE
)

kfre_default
```

## Extreme values
Extreme values will produce extreme risk estimates. Pre-filter implausible inputs before calling.

```{r}
# Pre-filter example
df_filtered <- df
df_filtered$eGFR[df_filtered$eGFR > 200] <- NA
df_filtered$UACR[df_filtered$UACR > 10000] <- NA

kfre_filtered <- kidney_failure_risk(
  data = df_filtered,
  col_map = col_map,
  na_action = "warn",
  verbose = FALSE
)

kfre_filtered
```

## Missing data handling
`keep` propagates NA risks; `omit` drops rows with missing required inputs; `error` stops on missingness.

```{r}
df_na <- df
df_na$eGFR[2] <- NA

keep_out <- kidney_failure_risk(df_na, col_map, na_action = "keep")
omit_out <- kidney_failure_risk(df_na, col_map, na_action = "omit")

list(keep_rows = nrow(keep_out), omit_rows = nrow(omit_out))
```

## Expectations
- All four predictors must be mapped and present; `sex` must be 1/2.
- Inputs must be numeric; non-finite become `NA` and may propagate depending on `na_action`.
- Output: tibble with `KFRE_2yr`, `KFRE_5yr`; row count follows `na_action` and any row filtering you apply.

## Verbose diagnostics

```{r, message=TRUE}
old_opt <- options(healthmarkers.verbose = "inform")
kidney_failure_risk(
  data    = df,
  col_map = col_map,
  verbose = TRUE
)
options(old_opt)
```

## Using real cohort data

The bundled simulated dataset includes `eGFR` and `UACR` columns. Age and sex
are also present, so the KFRE can run directly on `sim_small`:

```{r}
library(dplyr)
sim_path <- system.file("extdata", "simulated_hm_data.rds", package = "HealthMarkers")
sim_small <- dplyr::slice_head(readRDS(sim_path), n = 50)

kfre_sim <- kidney_failure_risk(
  data    = sim_small,
  col_map = list(age = "age", sex = "sex", eGFR = "eGFR", UACR = "UACR"),
  na_action   = "keep"
)

dplyr::bind_cols(dplyr::select(sim_small, age, sex, eGFR, UACR), kfre_sim) |> head()
```

> **Column recognition tip:** For biobank extracts where eGFR or UACR are named
> differently (e.g., `eGFR_CKD_EPI`, `ACR_mg_g`), run `hm_col_report(your_data)`
> to check what is auto-detected before building your `col_map`.

## Tips
- Check units before running (e.g., UACR mg/g vs mg/mmol); convert upstream if needed.
- Use `na_action = "error"` for strict QA; `warn` to surface high missingness while retaining rows.
