---
title: "Metabolic Risk Features"
output: rmarkdown::html_document
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  message = FALSE,
  warning = FALSE
)
knitr::opts_knit$set(purl = FALSE)
```

## Scope

Create pediatric-friendly binary risk flags (dyslipidemia, insulin resistance, hyperglycemia, hypertension) from routine labs and BP z-scores with configurable NA and extreme handling.

## When to use

- You have lipid panel, glucose/HbA1c, BP z-scores, age, and z-scored HOMA-IR and want quick binary risk features.
- You need NA policies, high-missingness warnings, and optional extreme-value screening/capping built in.
- You are working in pediatrics/adolescents where triglyceride thresholds differ by age and BP is already standardized.

## Inputs

- `data`: data.frame/tibble with required numeric inputs.
- `col_map`: named list mapping required keys (defaults to same names): `chol_total`, `chol_ldl`, `chol_hdl`, `triglycerides`, `age_year`, `z_HOMA`, `glucose`, `HbA1c`, `bp_sys_z`, `bp_dia_z`.
- Units assumed: lipids mmol/L; glucose mmol/L; HbA1c mmol/mol (IFCC); BP inputs already z-scores; age in years; `z_HOMA` is a z-scored HOMA-IR.
- `na_action`: `keep` (default) propagates NA; `omit` drops rows missing required inputs; `error` aborts; `ignore` acts like keep; `warn` acts like keep plus high-missingness warnings via `na_warn_prop` (default 0.2).

## Preparing derived inputs

Two required inputs are **derived**, not raw lab values:

### z_HOMA (standardised HOMA-IR)

Compute HOMA-IR from fasting glucose (mmol/L) and fasting insulin (µIU/mL), then
standardise to a z-score relative to a healthy reference population:

```{r, eval=FALSE}
# Step 1: compute HOMA-IR
data$HOMA_IR <- (data$fasting_glucose_mmol * data$fasting_insulin_uIU) / 22.5

# Step 2: standardise within cohort (or use external reference parameters)
data$z_HOMA  <- normalize_vec(data$HOMA_IR, method = "z")
```

Published paediatric reference values: Keskin *et al.* (2005, *Pediatr Diabetes*);
for adults, a 90th-percentile cutpoint (z ≈ 1.28) is commonly applied.

### BP z-scores (bp_sys_z / bp_dia_z)

BP z-scores must be derived using sex- and age-specific normative tables (AAP 2017
Fourth Report or WHO paediatric reference). The `calc_sds()` function can apply
stratified reference parameters if you have them; alternatively the `childsds`
package (CRAN) provides ready-to-use paediatric normative tables.

```{r, eval=FALSE}
# Manual z-score: (observed − reference_mean) / reference_sd
data$bp_sys_z <- (data$sbp - ref_sbp_mean) / ref_sbp_sd
```

## Quick start

```{r}
library(HealthMarkers)
library(tibble)

# Six children/adolescents spanning multiple metabolic risk combinations
peds <- tibble::tibble(
  chol_total    = c(5.4, 4.8, 3.9, 6.2, 5.1, 4.5),
  chol_ldl      = c(3.6, 2.9, 2.4, 4.1, 3.2, 2.8),
  chol_hdl      = c(0.9, 1.2, 1.4, 0.8, 1.0, 1.3),
  triglycerides = c(1.6, 1.0, 0.8, 2.0, 1.3, 0.9),
  age_year      = c(15,  9,  12,  17,  11,   8 ),
  z_HOMA        = c(1.5, 0.8, 0.2, 2.1, 1.0, -0.3),
  glucose       = c(5.8, 5.1, 4.9, 6.5, 5.4, 4.8),
  HbA1c         = c(40,  35,  32,  46,  38,  30 ),
  bp_sys_z      = c(1.2, 1.8, 0.3, 2.1, 0.8, -0.2),
  bp_dia_z      = c(0.5, 1.7, 0.2, 1.9, 0.6,  0.1)
)

metabolic_risk_features(
  data = peds,
  col_map = list(
    chol_total = "chol_total", chol_ldl = "chol_ldl", chol_hdl = "chol_hdl", triglycerides = "triglycerides",
    age_year = "age_year", z_HOMA = "z_HOMA", glucose = "glucose", HbA1c = "HbA1c",
    bp_sys_z = "bp_sys_z", bp_dia_z = "bp_dia_z"
  ),
  na_action = "keep"
)
```

## Extreme values
Extreme inputs will produce extreme or incorrect flags. Pre-filter implausible values before calling.

```{r}
extreme <- peds
extreme$triglycerides[1] <- 40  # extreme; pre-filter
extreme$bp_sys_z[2] <- 9        # out of expected range; pre-filter
extreme$triglycerides[extreme$triglycerides > 30] <- NA
extreme$bp_sys_z[abs(extreme$bp_sys_z) > 5] <- NA

metabolic_risk_features(
  data = extreme,
  col_map = list(
    chol_total = "chol_total", chol_ldl = "chol_ldl", chol_hdl = "chol_hdl", triglycerides = "triglycerides",
    age_year = "age_year", z_HOMA = "z_HOMA", glucose = "glucose", HbA1c = "HbA1c",
    bp_sys_z = "bp_sys_z", bp_dia_z = "bp_dia_z"
  ),
  na_action = "warn",
  verbose = TRUE
)
```

## Missing data policy

```{r}
missing <- peds
missing$glucose[2] <- NA

metabolic_risk_features(
  data = missing,
  col_map = list(chol_total = "chol_total", chol_ldl = "chol_ldl", chol_hdl = "chol_hdl", triglycerides = "triglycerides",
                 age_year = "age_year", z_HOMA = "z_HOMA", glucose = "glucose", HbA1c = "HbA1c",
                 bp_sys_z = "bp_sys_z", bp_dia_z = "bp_dia_z"),
  na_action = "omit"
)

metabolic_risk_features(
  data = missing,
  col_map = list(chol_total = "chol_total", chol_ldl = "chol_ldl", chol_hdl = "chol_hdl", triglycerides = "triglycerides",
                 age_year = "age_year", z_HOMA = "z_HOMA", glucose = "glucose", HbA1c = "HbA1c",
                 bp_sys_z = "bp_sys_z", bp_dia_z = "bp_dia_z"),
  na_action = "warn"
)
```

## Expanded example: drop incomplete rows

```{r}
ext_cap <- peds
ext_cap$triglycerides[1] <- 40   # extreme high
ext_cap$bp_sys_z[2] <- 9         # extreme high
ext_cap$HbA1c[2] <- NA           # missing HbA1c to trigger drop

out_cap <- metabolic_risk_features(
  data = ext_cap,
  col_map = list(
    chol_total = "chol_total", chol_ldl = "chol_ldl", chol_hdl = "chol_hdl", triglycerides = "triglycerides",
    age_year = "age_year", z_HOMA = "z_HOMA", glucose = "glucose", HbA1c = "HbA1c",
    bp_sys_z = "bp_sys_z", bp_dia_z = "bp_dia_z"
  ),
  na_action = "omit",
  verbose = TRUE
)

list(rows_returned = nrow(out_cap), flags = out_cap)
```

## Threshold quick-reference

| Flag | Logic (values in mmol/L unless noted) |
| --- | --- |
| Dyslipidemia | CT > 5.2 OR LDL > 3.4 OR HDL < 1.0 OR TG > 1.1 (age 0–9) OR TG > 1.5 (age 10–19) |
| Insulin resistance | z_HOMA > 1.28 (≈90th percentile) |
| Hyperglycemia | Glucose 5.6–6.9 OR HbA1c 39–47 mmol/mol |
| Hypertension | BP z-score > 1.64 (systolic or diastolic) |

## Handling and outputs

- Missingness: `keep/ignore` propagate NA; `omit` drops rows with missing required inputs; `warn` logs high-missingness; `error` aborts on missing required inputs.
- Outputs: tibble with factor columns `dyslipidemia`, `insulin_resistance`, `hyperglycemia`, `hypertension` (levels `0`/`1`). Age-based TG cutoffs: >1.1 mmol/L (0–9 y) or >1.5 mmol/L (10–19 y) for dyslipidemia; BP z > 1.64 for hypertension; z_HOMA > 1.28 for insulin resistance; glucose 5.6–6.9 or HbA1c 39–47 for hyperglycemia.

## Verbose diagnostics

```{r, message=TRUE}
old_opt <- options(healthmarkers.verbose = "inform")
metabolic_risk_features(
  data = peds,
  col_map = list(
    chol_total = "chol_total", chol_ldl = "chol_ldl", chol_hdl = "chol_hdl",
    triglycerides = "triglycerides", age_year = "age_year", z_HOMA = "z_HOMA",
    glucose = "glucose", HbA1c = "HbA1c", bp_sys_z = "bp_sys_z", bp_dia_z = "bp_dia_z"
  ),
  verbose = TRUE
)
options(old_opt)
```

## Tips

- Keep lipids/glucose in mmol/L and HbA1c in mmol/mol; BP inputs must already be z-scores.
- Use `na_action = "omit"` for analysis-ready flags; keep/warn during QA to inspect missingness.
- Watch for HbA1c reported in % (values <=14 suggest percent) and lipids/glucose in mg/dL (would appear ~×38.7 or ×18 higher); standardize units before calling.
- BP inputs must already be z-scores; large magnitude (|z| > 5) usually indicates scaling issues.
