---
title: "Neurofilament light chain (NfL) marker"
output: rmarkdown::html_document
---

```{r, include=FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(purl = FALSE)
```

## Scope
Incorporate a neurofilament light chain (NfL) concentration measurement into the analysis pipeline. NfL is a biomarker of neuroaxonal injury; elevated plasma or CSF levels indicate neurodegeneration and typically increase with age and neurological disease. This function validates and passes through the input values (assumed plasma pg/mL) with optional NA handling, returning a tidy tibble ready for downstream analysis.

## Load packages and demo data

```{r}
library(HealthMarkers)
library(tibble)

df <- tibble::tibble(
  NfL = c(8.5, 14.2, 22.1, NA, 35.0)
)
```

## Column map
The only required key is `nfl`.

```{r}
cm <- list(nfl = "NfL")
```

## Core calculation

```{r}
out <- nfl_marker(
  data = df,
  col_map = cm,
  na_action = "keep",
  verbose = FALSE
)

out
```

## Missing data handling
`na_action = "keep"` preserves rows with `NA` outputs; `omit` drops rows with missing required inputs; `error` aborts if any required input is missing; `warn` keeps rows but emits a warning.

```{r}
df_na <- tibble::tibble(NfL = c(8.5, NA, 22.1))

keep_out <- nfl_marker(df_na, cm, na_action = "keep")
omit_out <- nfl_marker(df_na, cm, na_action = "omit")

list(keep_rows = nrow(keep_out), omit_rows = nrow(omit_out))
```

## Extreme values
Negative or implausibly high NfL values will be passed through as-is. Pre-filter before calling.

```{r, warning=TRUE}
df_ext <- tibble::tibble(NfL = c(8.5, -5, 22.1))  # negative value is implausible; pre-filter
# Pre-filter:
df_ext$NfL[df_ext$NfL < 0] <- NA
nfl_marker(df_ext, col_map = cm, na_action = "keep")
```

## Verbose diagnostics
Set `verbose = TRUE` (and `healthmarkers.verbose = "inform"`) to surface three structured messages on each call: preparing inputs, the column map, and a results summary.

```{r, message=TRUE}
old_opt <- options(healthmarkers.verbose = "inform")
nfl_marker(tibble::tibble(NfL = c(8.5, 14.2, 22.1)), cm, verbose = TRUE)
options(old_opt)
```

## Column recognition

Run `hm_col_report(your_data)` to check which analyte columns are auto-detected
before building your `col_map`. See the
[Multi-Biobank Compatibility](multi_biobank.html) article for recognised synonyms
across major biobanks.

```{r, eval=FALSE}
hm_col_report(your_data)
```

## Tips
- Units are preserved as-is; ensure all values are in the same unit (e.g., pg/mL) before calling.
- Age-adjusted reference ranges are not applied by this function; use downstream tools or stratified z-scores for age normalisation.
- For batch processing with imputed datasets, combine with `impute_missing()` upstream to handle systematic missingness before running `nfl_marker()`.
