---
title: "Normalising marker outputs"
output: rmarkdown::html_document
---

```{r, include=FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE, message = FALSE, warning = FALSE)
knitr::opts_knit$set(purl = FALSE)
```

## Scope

This article explains how to normalise biomarker outputs computed by HealthMarkers
functions using `hm_normalize()` and the underlying `normalize_vec()`.

Normalisation is useful before:

- Genome-wide association studies (GWAS) — inverse-normal transform is standard
- Principal component analysis (PCA) — z-score or min-max
- Polygenic score construction — z-score or robust scaling
- Visualisation — when variables are on very different scales

## When to use `hm_normalize()` vs the built-in `normalize` argument

The `normalize` argument is only implemented in the insulin sensitivity functions
(`fasting_is()`, `ogtt_is()`, `adipo_is()`, `all_insulin_indices()`).
For all other domain functions — `glycemic_markers()`, `lipid_markers()`,
`renal_markers()`, etc. — the `normalize` argument has no effect.

`hm_normalize()` works post-computation on **any** data frame and covers all cases.

## Setup

```{r}
if (requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all()
} else {
  library(HealthMarkers)
}

sim_path <- system.file("extdata", "simulated_hm_data.rds", package = "HealthMarkers")
sim      <- readRDS(sim_path)[1:50, ]
```

## Step 1 — compute markers

```{r}
out <- all_health_markers(
  data    = sim,
  which   = c("glycemic", "lipid", "renal", "inflammatory"),
  verbose = FALSE
)

# New marker columns added by all_health_markers()
new_cols <- setdiff(names(out), names(sim))
cat("New marker columns:", length(new_cols), "\n")
```

## Step 2 — normalise with `hm_normalize()`

### Z-score (mean 0, sd 1)

```{r}
out_z <- hm_normalize(out, cols = new_cols, method = "z")
round(head(out_z[new_cols[1:5]]), 3)
```

### Rank-based inverse-normal transform (Rankit)

The most common choice for GWAS. Normally distributed regardless of the original
distribution.

```{r}
out_int <- hm_normalize(out, cols = new_cols, method = "inverse")
round(head(out_int[new_cols[1:5]]), 3)
```

Use the Blom denominator instead of Rankit:

```{r}
out_blom <- hm_normalize(out, cols = new_cols, method = "inverse",
                          invnorm_denominator = "blom")
```

### Min-max scaling to [0, 1]

```{r}
out_range <- hm_normalize(out, cols = new_cols, method = "range")
round(head(out_range[new_cols[1:5]]), 3)
```

Scale to a custom interval (e.g. [-1, 1]):

```{r}
out_range2 <- hm_normalize(out, cols = new_cols, method = "range",
                            feature_range = c(-1, 1))
```

### Robust median/MAD scaling

Less sensitive to outliers than z-score.

```{r}
out_rob <- hm_normalize(out, cols = new_cols, method = "robust")
round(head(out_rob[new_cols[1:5]]), 3)
```

## Protecting covariate columns

Use `skip_cols` to exclude variables you want to keep on their original scale.
When `cols = NULL` (the default), `hm_normalize()` targets all numeric columns;
`skip_cols` removes covariates from that set.

```{r}
# Normalise all numeric columns except age, BMI, sex-encoded variables
out_protected <- hm_normalize(out, method = "inverse",
                               skip_cols = c("age", "BMI", "sex"))
```

## Using `normalize_vec()` on a single vector

`hm_normalize()` calls `normalize_vec()` internally. You can also use it directly:

```{r}
x <- out[[new_cols[1]]]

normalize_vec(x, method = "z")[1:10]
normalize_vec(x, method = "inverse")[1:10]
normalize_vec(x, method = "range")[1:10]
normalize_vec(x, method = "robust")[1:10]
```

## Comparison of methods

```{r}
compare <- data.frame(
  raw     = x,
  z       = normalize_vec(x, "z"),
  INT     = normalize_vec(x, "inverse"),
  range01 = normalize_vec(x, "range"),
  robust  = normalize_vec(x, "robust")
)
round(head(compare, 10), 3)
```

Summary statistics after normalisation:

```{r}
round(sapply(compare, function(v) c(mean = mean(v, na.rm = TRUE),
                                     sd   = sd(v, na.rm = TRUE),
                                     min  = min(v, na.rm = TRUE),
                                     max  = max(v, na.rm = TRUE))), 3)
```

## See also

- `?hm_normalize` — full argument reference
- `?normalize_vec` — single-vector normalisation details
- [Fasting insulin sensitivity](fasting_is.html) — `normalize` argument used at compute time
