---
title: "Nutrient Markers"
output: rmarkdown::html_document
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  message = FALSE,
  warning = FALSE
)
knitr::opts_knit$set(purl = FALSE)
```

## Scope

Compute nutrient-related ratios/products (FerritinTS, AGR, Omega3Index, Mg_Cr_Ratio, GlycatedAlbuminPct, UA_Cr_Ratio, BUN_Cr_Ratio, Ca_x_Phosphate, AnionGap, Tyr_Phe_Ratio) with NA/extreme handling and safe divisions; no unit conversion.

## When to use

- You have routine nutrient labs (iron, protein, omega-3, renal, electrolytes, amino acids) and want quick derived ratios.
- You need built-in NA policies, high-missingness diagnostics, and optional extreme-value scanning/capping.
- You can supply inputs in expected units and accept that missing inputs return NA for dependent markers.

## Inputs

- `data`: data frame/tibble containing any subset of recognized inputs.
- `col_map`: optional named list mapping keys (defaults to identity): `ferritin`, `transferrin_sat`, `albumin`, `total_protein`, `EPA`, `DHA`, `Mg`, `creatinine`, `glycated_albumin`, `uric_acid`, `BUN`, `phosphate`, `calcium`, `Na`, `K`, `Cl`, `HCO3`, `Tyr`, `Phe`. Unrecognized keys are ignored. Non-numeric values are coerced with warnings; non-finite set to `NA`.
- `na_action`: `keep` (default) propagates NA; `omit` drops rows with NA in used inputs; `error` aborts when used inputs contain NA.
- `na_warn_prop`: threshold for high-missingness diagnostics (default 0.2; shown in verbose/debug).
- `verbose`: optional progress and summary logging.

## Quick start

```{r}
library(HealthMarkers)
library(tibble)

df <- tibble::tibble(
  ferritin = c(50, 100), transferrin_sat = c(30, 50),
  albumin = c(45, 40), total_protein = c(70, 75),
  EPA = c(2.0, 2.5), DHA = c(4.0, 4.5),
  Mg = c(0.85, 0.90), creatinine = c(80, 90),
  glycated_albumin = c(12, 14), uric_acid = c(300, 400), BUN = c(14, 16),
  phosphate = c(1.0, 1.2), calcium = c(2.3, 2.4),
  Na = c(140, 138), K = c(4.2, 4.0), Cl = c(100, 102), HCO3 = c(24, 26),
  Tyr = c(60, 70), Phe = c(50, 55)
)

nutrient_markers(
  data = df,
  col_map = NULL,
  na_action = "keep",
  verbose = FALSE
)
```

## Extreme values
Extreme inputs will produce extreme derived markers. Pre-filter implausible values before calling.

```{r}
# Pre-filter extreme ferritin
df_clean <- df
df_clean$ferritin[df_clean$ferritin > 2000] <- NA

nutrient_markers(
  data = df_clean,
  col_map = NULL,
  na_action = "omit",
  verbose = TRUE
)
```

## Missing-data policy

```{r}
try(
  nutrient_markers(
    data = df,
    col_map = NULL,
    na_action = "error"
  )
)
```

## Outputs and expectations

- Returns a tibble with: FerritinTS, AGR, Omega3Index, Mg_Cr_Ratio, GlycatedAlbuminPct, UA_Cr_Ratio, BUN_Cr_Ratio, Ca_x_Phosphate, AnionGap, Tyr_Phe_Ratio.
- Each marker is computed only when its inputs are present; otherwise `NA` is returned for that marker.
- Zero/non-finite denominators are set to NA with consolidated warnings; no unit harmonization is applied.

## Verbose diagnostics

```{r, message=TRUE}
old_opt <- options(healthmarkers.verbose = "inform")
nutrient_markers(
  data = df,
  col_map = NULL,
  verbose = TRUE
)
options(old_opt)
```

## Column recognition

Run `hm_col_report(your_data)` to check which analyte columns are auto-detected
before building your `col_map`. See the
[Multi-Biobank Compatibility](multi_biobank.html) article for recognised synonyms
across major biobanks.

```{r, eval=FALSE}
hm_col_report(your_data)
```

## Tips

- Provide only the inputs you have; missing inputs yield NA for dependent markers.
- Omega3Index expects EPA/DHA as percentages; AGR uses globulin = total_protein - albumin.
- Use `na_action = "omit"` when you prefer row-complete outputs; use `keep` during QA to inspect missingness.
