---
title: "Sweat Markers"
output: rmarkdown::html_document
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  message = FALSE,
  warning = FALSE
)
knitr::opts_knit$set(purl = FALSE)
```

## Scope
Compute sweat chloride, sodium/potassium ratio, sweat lactate, and sweat rate (L/m²/h) from sweat assays and anthropometrics. Includes mapping validation, NA policies, optional extreme screening/capping on inputs, and safe handling of zero/invalid denominators.

## When to use this
- You have sweat chloride, sodium, potassium, lactate, and anthropometrics (weight before/after, duration, body surface area) and want to derive sweat rate and ratios.
- You want explicit NA handling, optional extreme-value capping, and consolidated zero-denominator warnings.

## What you need (inputs & options)

| Argument       | Purpose / Options                                               | Notes                                               |
|----------------|-----------------------------------------------------------------|-----------------------------------------------------|
| data           | Data frame/tibble with sweat and anthropometric measures        | Columns mapped via `col_map`                        |
| col_map        | Named list mapping required inputs                              | sweat_chloride, sweat_Na, sweat_K, sweat_lactate, weight_before, weight_after, duration, body_surface_area |
| na_action      | Missing-data policy for required inputs                         | "keep" (default), "omit", "error"                  |
| na_warn_prop   | Proportion threshold for high-missingness diagnostics           | Default 0.2 (shown in debug/verbose)                |
| verbose        | Emit progress and completion summaries                          | Default FALSE                                       |

**Required columns (col_map):** sweat_chloride, sweat_Na, sweat_K, sweat_lactate (mmol/L), weight_before, weight_after (kg), duration (h), body_surface_area (m²).

**Units:** No conversion is performed. Supply all values in the expected units above.

## Handling and expectations
- Validation & coercion: mapped columns must exist; non-numeric inputs are coerced with warnings if NAs are introduced; non-finite become NA.
- Missingness: `keep` propagates NA; `omit` drops rows with any required NA; `error` aborts if required inputs contain NA.
- High-missing diagnostics: controlled by `na_warn_prop`; shown when `verbose` (or debug) is enabled.
- Safe division: all ratios use a helper that sets NA for zero/invalid denominators and emits a consolidated warning.
- Sweat rate: $(\text{weight_before} - \text{weight_after}) / \text{duration} / \text{body_surface_area}$ (1 kg ≈ 1 L water).
- Empty result: if `na_action = "omit"` drops all rows, returns a zero-row tibble with expected columns.

## Outputs
- Returns a tibble (rows follow `na_action`) with: sweat_chloride, Na_K_ratio, sweat_lactate, sweat_rate.
- NA arises from missing/invalid inputs, zero/invalid denominators, or dropped rows under `omit`.

## Worked example 1: Standard usage, keep NAs
```{r}
library(HealthMarkers)
library(tibble)

df <- tibble::tibble(
  sweat_chloride = c(45, 60, NA),
  sweat_Na = c(55, 70, 60),
  sweat_K = c(5, 4, 0),
  sweat_lactate = c(4.8, 5.2, 6.0),
  weight_before = c(70.0, 80.0, 75.0),
  weight_after = c(69.5, 79.2, 74.8),
  duration = c(1.0, 1.5, 0.5),
  body_surface_area = c(1.9, 2.1, 2.0)
)

sweat_markers(
  data = df,
  col_map = list(
    sweat_chloride = "sweat_chloride",
    sweat_Na = "sweat_Na",
    sweat_K = "sweat_K",
    sweat_lactate = "sweat_lactate",
    weight_before = "weight_before",
    weight_after = "weight_after",
    duration = "duration",
    body_surface_area = "body_surface_area"
  ),
  na_action = "keep",
  verbose = TRUE
)
```
*Interpretation:* Returns all outputs; NA propagates where inputs are missing or denominators are zero (e.g., sweat_K = 0).

## Worked example 2: Pre-filter extremes, drop incomplete
```{r}
df2 <- tibble::tibble(
  cl = c(180, 250),       # extreme chloride; pre-filter
  na = c(60, 10),
  k = c(4, 0.5),
  lac = c(6, 60),         # extreme lactate; pre-filter
  w_before = c(75, 90),
  w_after = c(74.2, 89.0),
  dur = c(1, 0.2),
  bsa = c(2.0, 1.8)
)

# Pre-filter implausible values before calling
df2$cl[df2$cl > 200] <- NA
df2$lac[df2$lac > 50] <- NA

sweat_markers(
  data = df2,
  col_map = list(
    sweat_chloride = "cl",
    sweat_Na = "na",
    sweat_K = "k",
    sweat_lactate = "lac",
    weight_before = "w_before",
    weight_after = "w_after",
    duration = "dur",
    body_surface_area = "bsa"
  ),
  na_action = "omit",
  verbose = TRUE
)
```
*Interpretation:* Implausible values are set to NA before calling; incomplete rows are then dropped.

## Troubleshooting & common pitfalls
- Units: all values must be in the expected units; convert upstream if needed.
- Missing columns: ensure every required `col_map` key is mapped to an existing column.
- Zero/invalid denominators: produce NA outputs and a warning; check for sweat_K = 0, duration = 0, or BSA = 0.
- All NA outputs: usually due to missing/invalid inputs, zero denominators, or aggressive `na_action = "omit"`.

## Verbose diagnostics
```{r, message=TRUE}
old_opt <- options(healthmarkers.verbose = "inform")
sweat_markers(
  data = tibble::tibble(
    sweat_chloride    = 45,
    sweat_Na          = 55,
    sweat_K           = 5,
    sweat_lactate     = 4.8,
    weight_before     = 70,
    weight_after      = 69.5,
    duration          = 1.0,
    body_surface_area = 1.9
  ),
  col_map = list(
    sweat_chloride    = "sweat_chloride",
    sweat_Na          = "sweat_Na",
    sweat_K           = "sweat_K",
    sweat_lactate     = "sweat_lactate",
    weight_before     = "weight_before",
    weight_after      = "weight_after",
    duration          = "duration",
    body_surface_area = "body_surface_area"
  ),
  verbose = TRUE
)
options(old_opt)
```

## Column recognition

Run `hm_col_report(your_data)` to check which analyte columns are auto-detected
before building your `col_map`. See the
[Multi-Biobank Compatibility](multi_biobank.html) article for recognised synonyms
across major biobanks.

```{r, eval=FALSE}
hm_col_report(your_data)
```

## Tips for best results
- Choose `na_action = "omit"` for modeling datasets that require complete cases; use `keep` for exploratory review.
- Review warnings for zero denominators; these often indicate data entry or assay issues.

## Validation notes
- Na_K_ratio = sweat_Na / sweat_K; sweat_rate = (weight_before - weight_after) / duration / body_surface_area (1 kg ≈ 1 L).
- All divisions are safe: zero/invalid denominators yield NA and a warning.

## See also
- Function docs: `?sweat_markers`
- Related vignettes: health_markers, metabolic_markers, urine_markers
