---
title: "Urine Markers"
output: rmarkdown::html_document
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  message = FALSE,
  warning = FALSE
)
knitr::opts_knit$set(purl = FALSE)
```

## Scope
Compute urine-only renal and tubular markers: UACR, KDIGO albuminuria stage (A1–A3), microalbuminuria flag, UPCR (if protein provided), urine Na/K ratio (if electrolytes provided), and creatinine-normalized tubular markers (NGAL, KIM1, NAG, Beta2Micro, A1Micro, IL18, L_FABP). Includes mapping validation, NA policies, optional extreme screening/capping, and safe division with consolidated zero-denominator warnings. No unit conversion is applied.

## When to use this
- You have urine albumin and creatinine (mg/L, mg/dL) and want UACR and albuminuria staging.
- You also have urine protein, Na/K, or tubular markers and need creatinine-normalized ratios.
- You want robust NA/extreme handling with safe divisions to avoid Inf/NaN.

## What you need (inputs & options)

| Argument       | Purpose / Options                                               | Notes                                               |
|----------------|-----------------------------------------------------------------|-----------------------------------------------------|
| data           | Data frame/tibble with urine assays                             | Columns mapped by name (defaults assume same names) |
| na_action      | Missing-data policy for required inputs                         | "keep" (default), "omit", "error"                  |
| na_warn_prop   | Proportion threshold for high-missingness diagnostics           | Default 0.2 (shown in debug/verbose)                |
| verbose        | Emit progress and completion summaries                          | Default FALSE                                       |

**Required columns:** urine_albumin (mg/L), urine_creatinine (mg/dL).

**Optional columns:** urine_protein (mg/L), urine_Na (mmol/L), urine_K (mmol/L), NGAL, KIM1, NAG, beta2_micro (Beta2Micro output), a1_micro (A1Micro output), IL18, L_FABP.

**Units:** No conversion. Ensure albumin/protein mg/L, creatinine mg/dL, Na/K mmol/L; tubular marker units consistent across markers.

## Handling and expectations
- Validation & coercion: required columns must exist; numeric coercion with warnings if NAs introduced; non-finite set to NA.
- Missingness: `keep` propagates NA; `omit` drops rows with required NA; `error` aborts if required inputs contain NA.
- High-missing diagnostics: controlled by `na_warn_prop`; shown when `verbose` (or debug) is enabled.
- Safe division: zero/invalid denominators set outputs to NA with a consolidated warning.
- Albuminuria staging: A1 < 30 mg/g, A2 30–300 mg/g, A3 > 300 mg/g; microalbuminuria flag is `micro` for 30–300 mg/g, else `normal`.
- Empty result: if `na_action = "omit"` drops all rows, returns a zero-row tibble with expected columns.

## Outputs
- UACR, albuminuria_stage (A1/A2/A3), microalbuminuria (normal/micro), UPCR (if urine_protein present), U_Na_K_ratio (if Na & K present), NGAL_per_gCr, KIM1_per_gCr, NAG_per_gCr, Beta2Micro_per_gCr, A1Micro_per_gCr, IL18_per_gCr, L_FABP_per_gCr.
- NA arises from missing/invalid inputs, zero denominators, or dropped rows under `omit`.

## Worked example 1: Standard inputs, keep NAs
```{r}
library(HealthMarkers)
library(tibble)

df <- tibble::tibble(
  urine_albumin = c(25, 450, NA),      # mg/L
  urine_creatinine = c(1.0, 0.8, 0.5), # mg/dL
  urine_protein = c(120, NA, 90),      # mg/L
  urine_Na = c(100, 80, 90),
  urine_K = c(40, 30, 0),
  NGAL = c(15, 40, 20)
)

urine_markers(
  data = df,
  na_action = "keep",
  verbose = TRUE
)
```
*Interpretation:* Computes UACR, staging, micro flag, UPCR, U_Na_K_ratio, and NGAL_per_gCr where inputs exist; NA propagates for missing or zero denominators (e.g., urine_K = 0).

## Worked example 2: Drop incomplete rows
```{r}
df2 <- tibble::tibble(
  urine_albumin = c(20, 120000),    # extreme albumin; pre-filter
  urine_creatinine = c(1.2, 0.005), # very low creatinine; pre-filter
  urine_protein = c(80, 90000),
  urine_Na = c(90, 500),
  urine_K = c(35, 210),
  NGAL = c(10, 200000)
)

# Pre-filter implausible values before calling
df2$urine_albumin[df2$urine_albumin > 100000] <- NA
df2$NGAL[df2$NGAL > 100000] <- NA

urine_markers(
  data = df2,
  na_action = "omit",
  verbose = TRUE
)
```
*Interpretation:* Implausible values are set to NA before calling; incomplete rows are then dropped; zero-denominator warnings list counts.

## Troubleshooting & common pitfalls
- Units: ensure albumin/protein mg/L, creatinine mg/dL; Na/K mmol/L; tubular marker units consistent across markers.
- Missing columns: required columns must be present; optional markers only emit outputs when supplied.
- Zero denominators: check urine_creatinine, urine_K; warnings list how many cases; outputs set to NA.
- All NA outputs: often due to missing required inputs, zero denominators, or aggressive `na_action = "omit"`.

## Verbose diagnostics
```{r, message=TRUE}
old_opt <- options(healthmarkers.verbose = "inform")
urine_markers(
  data = tibble::tibble(urine_albumin = 30, urine_creatinine = 2),
  verbose = TRUE
)
options(old_opt)
```

## Column recognition

Run `hm_col_report(your_data)` to check which analyte columns are auto-detected
before building your `col_map`. See the
[Multi-Biobank Compatibility](multi_biobank.html) article for recognised synonyms
across major biobanks.

```{r, eval=FALSE}
hm_col_report(your_data)
```

## Tips for best results
- Use `na_action = "omit"` when downstream analyses need complete cases; `keep` for exploratory review.
- Monitor zero-denominator warnings—extremely low creatinine or K values will blank ratios and per-gCr markers.

## Validation notes
- UACR = (urine_albumin / urine_creatinine) * 1000 (mg/g); UPCR analogous when protein present.
- U_Na_K_ratio = urine_Na / urine_K.
- Tubular markers per g creatinine: marker / (urine_creatinine * 0.01) where denominator is g/L.
- Divisions are safe: non-finite/zero denominators yield NA with a consolidated warning.

## See also
- Function docs: `?urine_markers`
- Related vignettes: renal_markers, health_markers, corrected_calcium
